---
title: "Human Organoid Unsupervised Exploration"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE)
```

# Setup

```{r}
library(tidyverse)
library(here)
library(feather)
library(monocle3)
library(ggrastr)
library(cowplot)
library(princurve)

# install.packages('devtools')
# devtools::install_github('VPetukhov/ggrastr')
```

I load the dataset in question. 

```{r, eval = FALSE}
pca_raw <- feather::read_feather(path = here("vignettes/07_organoid_unsupervised_exploration/embeddings/hdf_pca_feather"))
#pca_raw <- pca_raw_store
```

I give a short overview of the dataset. 

```{r, eval = FALSE}
pca_raw %>% 
  sample_n(10000) %>%
  ggplot(aes(V1, V2, color = morphological_class)) + 
  geom_point() + 
  theme_classic() + 
  scale_color_viridis_d()
  
```


```{r, eval = FALSE}
# for demo: subsample the original dataset
set.seed(123)

pca_raw_store <- pca_raw
pca_raw <- pca_raw %>% 
  sample_n(100000)
```


# Import into Monocle 3

I continue importing the dataset into Monoocle3 

```{r, eval = FALSE}
# generating metadata objects
pca_anno_df <- pca_raw %>% dplyr::select(-(V1:V25)) %>% 
  mutate(uuid = paste(Plate, Well, Field, ObjectID, sep = "_")) %>% 
  mutate(uuid2 = uuid) %>% 
  mutate(Size = as.numeric(Size)) %>%
  mutate(Size_log = log(Size)) %>%
  as.data.frame() %>% 
  column_to_rownames("uuid2")

pca_matrix <- pca_raw %>% dplyr::select((V1:V25)) %>% as.data.frame() %>% magrittr::set_rownames(pca_anno_df$uuid) %>% magrittr::set_colnames(c(paste0("PC", c(1:25)))) %>% as.matrix()


ods <- new_cell_data_set(pca_matrix %>% t(),
                         cell_metadata = pca_anno_df)
```

I manually inject the PCA compression of the data into the object

```{r, eval = FALSE}
reducedDims(ods)$PCA <- pca_matrix
```

I run the dim reduction methods

```{r, eval = FALSE}
set.seed(123)
ods <- reduce_dimension(ods,
                        reduction_method = "UMAP",
                        umap.n_neighbors = 15L,
                        umap.fast_sgd=TRUE, 
                        cores = parallel::detectCores(),
                        verbose = TRUE)

saveRDS(ods, here("data/monocle/bulk_pca_cluster.Rds"))
```

I load the UMAP embedding of the organoid dataset. 

```{r}
ods <- readRDS(here("vignettes/07_organoid_unsupervised_exploration/monocle/bulk_pca_cluster.Rds"))
```

I plot a size-distribution. 

```{r, eval = TRUE}
gg_size_dist <- colData(ods) %>% as_tibble() %>% 
  ggplot(aes(Size)) + 
  geom_histogram() + 
  theme_classic() 

gg_size_dist + ggsave(here("results/figures/single_organoid/gg_size_dist.pdf"))
```


I plot the cells by size

```{r, eval = FALSE}
# gg_size <- plot_cells(ods, color_cells_by="Size", 
#            alpha = 0.1,
#            rasterize= TRUE) + 
#   scale_color_viridis_c() + 
#   theme(legend.position = "bottom")
```

I use an alterantive, manual, plotting method. In order to plot the object data, I export a subset of organoids. I plot the cells by size.

```{r}
set.seed(123)

df <- reducedDims(ods)$UMAP %>% cbind(colData(ods)) %>% as_tibble() %>%
  sample_frac(.01)

gg_size <- df %>%
  #filter(Size < 1000) %>%
  ggplot(aes(V1, V2, color = Size_log)) + 
  geom_point_rast(alpha = 0.5, size = 0.35) + 
  scale_color_viridis_c() +
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8))

```



```{r}
set.seed(123)

df <- reducedDims(ods)$UMAP %>% cbind(colData(ods)) %>% as_tibble() %>%
  sample_frac(0.1)

gg_size_supp <- df %>%
  mutate(drug = if_else(Drug == "DMSO", "DMSO", "other")) %>%
  ggplot(aes(V1, V2, color = Size_log)) + 
  geom_point_rast(alpha = 0.5, size = 0.35) + 
  scale_color_viridis_c() +
  theme_cowplot() +
  labs(x = "UMAP 1",
       y = "UMAP 2") + 
  theme(legend.position = "bottom") +
  facet_wrap(~ drug)

gg_size_supp + ggsave(here("figures/gg_size_all.pdf"))
```

I create a single plot showing the two extreme organoid lines and their distribution within the embedding. 

```{r}
set.seed(123)

df <- reducedDims(ods)$UMAP %>% cbind(colData(ods)) %>% as_tibble() %>%
  filter(Drug == "DMSO") 

gg_line <- df %>% dplyr::select(-Line) %>%
  ggplot(aes(V1, V2)) + 
  geom_point_rast(alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point_rast(data = df %>%
                   # filter(Line %in% c("D021T01")) %>%
    filter(Line %in% c("D055T01", "D007T01", "D021T01", "D019T01", "D027T01")) %>% 
    mutate(Line = factor(Line, levels = c("D055T01", "D007T01", "D021T01", "D019T01", "D027T01"))),
    #mutate(Line = factor(Line, levels = c("D021T01"))),
  aes(color = Line),alpha = .2, size = 0.35, shape=16) + 
  facet_wrap( ~ Line, ncol =1) + 
  scale_color_manual(values = c(c("#D80D12", "#461C01", "#9a4c91", "#70BE6F", "#24345E"))) + 	
  #geom_density2d(color = "black") + 
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) + 
  theme_cowplot(font_size = 8) + 
  theme(legend.position = "nothing")
```



```{r}
gg_line_overview <- plot_cells(ods, color_cells_by="morphological_class",
           alpha = 0.05, label_cell_groups = FALSE,
           rasterize = TRUE) + 
  facet_wrap(~ Line) + 
  theme(legend.position = "nothing") + 
  scale_color_manual(values = rev(c("#A2549B", "#D80D12", "#6E1614","#69B563", "#A9CEE1","#1E3B87"))) 
```

I directly write this supplementary plot to disc. 

```{r, eval = FALSE}
gg_line_overview + 
  ggsave(here("line.pdf"), height = 8, width = 8)
```


For the purpose of performance, I subsample my gene expression dataset and start clustering.

```{r, eval = FALSE}
set.seed(1234)
fraction = 0.01
ods_sub <- ods[, sample(seq_len(ncol(ods)), ncol(ods)*fraction, replace = FALSE)]

k_in = 100
ods_sub = cluster_cells(ods_sub, random_seed = 123, verbose = TRUE, 
                      reduction_method = "UMAP")
```

To get a general understanding of different phenotypes, I perform k-means clustering, which, I hope, is more scalable than NN-based graph clustering with a large number of neighbours. 

```{r}
kmeans_df <- readRDS(here("data/kmeans/pca_kmeans_df.Rds"))
```

I visualize the elbow plot. 

```{r}
gg_elbow <- kmeans_df %>% unnest(var_explained) %>% 
  ggplot(aes(k, var_explained)) +
  geom_point() + 
  theme_classic() + 
  geom_hline(yintercept = 0.9) + 
  geom_vline(xintercept = 8, linetype = "dotted")
```


```{r, eval = FALSE}
plot_clusgap = function(clusgap, title="Gap Statistic calculation results"){
    require("ggplot2")
    gstab = data.frame(clusgap$Tab, k=1:nrow(clusgap$Tab))
    p = ggplot(gstab, aes(k, gap)) + geom_line() + geom_point(size=4)
    p = p + geom_errorbar(aes(ymax=gap+SE.sim, ymin=gap-SE.sim), width = 0.1)
    p = p + theme_classic() + geom_vline(xintercept = 8, linetype = "dotted") + 
      scale_x_continuous(breaks = c(1:10))
    return(p)
}

gg_gap <- plot_clusgap(gap)
```

I visualize the clusters of the selected k=8. 

```{r, eval = FALSE}
kmeans_df %>% 
  filter(k == 8) %>% 
  .$kmeans %>% .[[1]] %>% .$cluster %>% 
  cbind(df) %>% janitor::clean_names() %>% 
  mutate(x = factor(x)) %>%
  ggplot(aes(v1, v2, color = x)) + 
  geom_point_rast(alpha = 0.7, size = 0.35) + 
  theme_classic() + 
  scale_color_brewer(type = "qual")
  
```

I am not absolutely hyped about the clustering. I think the clustering will be better if I run it on the first 25 principal components, rather than the 2 dimensional UMAP embedding. 

## k-means clustering of PCA embedding

```{r, eval = FALSE}
set.seed(123456)

df <- reducedDims(ods)$PCA %>% cbind(colData(ods)) %>% as_tibble()


pca_kmeans_df <- tibble(k = c(1:20)) %>% 
  mutate(kmeans = map(k, ~ kmeans(df %>% dplyr::select(PC1:PC25), centers = .x, iter.max = 10000, nstart=10, algorithm="MacQueen")),
         var_explained = map(kmeans, ~ .x$betweenss/.x$totss))
```

I visualize the elbow plot. 

```{r, eval = FALSE}
pca_kmeans_df <- readRDS(here("data/kmeans/pca_kmeans_df.Rds"))

gg_elbow_pca <- pca_kmeans_df %>% unnest(var_explained) %>% 
  ggplot(aes(k, var_explained)) +
  geom_point() + 
  theme_classic() + 
  geom_hline(yintercept = 0.9) + 
  geom_vline(xintercept = 8, linetype = "dotted")

gg_elbow_pca
```

Alterantively, I calculate the gap statistic. 

```{r, eval = TRUE}
gap_pca <- readRDS(here("data/kmeans/pca_gap.Rds"))
```

```{r, eval = FALSE}
plot_clusgap = function(clusgap, title="Gap Statistic calculation results"){
    require("ggplot2")
    gstab = data.frame(clusgap$Tab, k=1:nrow(clusgap$Tab))
    p = ggplot(gstab, aes(k, gap)) + geom_line() + geom_point(size=4)
    p = p + geom_errorbar(aes(ymax=gap+SE.sim, ymin=gap-SE.sim), width = 0.1)
    p = p + theme_classic() + geom_vline(xintercept = 9, linetype = "dotted") + 
      scale_x_continuous(breaks = c(1:10))
    return(p)
}

gg_gap <- plot_clusgap(gap_pca)

gg_gap + ggsave(here("figures/gap_umap.pdf"))
saveRDS(gg_gap, here("figures/gap_umap.Rds"))
```

Based on the two results, I decide to set k=8 and cluster the complete dataset. 

```{r, eval = FALSE}
set.seed(123456)
df_complete_pca <- reducedDims(ods)$PCA %>% cbind(colData(ods)) %>% as_tibble()

kmeans_complete <- kmeans(df_complete_pca %>% dplyr::select(P1:P25), centers = 8, iter.max = 10000, nstart=10, algorithm="MacQueen")
kmeans_complete$betweenss/kmeans_complete$totss
```

I visualize the clusters of the selected k=8. 

```{r, eval = FALSE}
df <- reducedDims(ods)$UMAP %>% cbind(colData(ods)) %>% as_tibble()

gg_kmeans <- kmeans_df %>% 
  filter(k == 9) %>% 
  .$kmeans %>% .[[1]] %>% .$cluster %>% 
  cbind(df) %>% janitor::clean_names() %>% 
  mutate(x = factor(x)) %>%
  sample_frac(0.01) %>%
  ggplot(aes(v1, v2, color = x)) + 
  geom_point_rast(alpha = 1, size = 0.35) + 
  theme_classic() 
  
gg_kmeans
```

kmeans clustering on the PCA does not recover complex relationships, such as the leftmost dead cluster, same as model-based clustering. Graph based clustering at the scale presented here did not recover an interpretable number of clusters with the available compute ressources. 
 
The final clustering method to move forward is the benchmark MST method proposed in "A comparison of single-cell trajectory inference methods" in which a model-based clustering was performed on the UMAP embedding of the dataset. For scalability reasons, the clustering was performend on a random sample of the complete dataset and labels were predicted.

Similar to the previously mentioned study, we next calculated a minimum spanning tree and created a graph of centroids. 

```{r}
clust <- readRDS(here("data/mclust/umap_mclust_k6.Rds"))
clust_predict  <- readRDS(here("data/mclust/umap_mclust_predict_k6.Rds")) # initially with 15 clusters

centers <- t(clust$parameters$mean)

milestone_ids <- paste0("M", seq_len(nrow(centers)))
rownames(centers) <- milestone_ids

# convert distance to similarity
dis <- as.matrix(dist(centers))
rownames(dis) <- colnames(dis) <- milestone_ids

disdf <- dis %>%
  reshape2::melt(varnames = c("from", "to"), value.name = "weight") %>%
  na.omit()

# calculate mst
gr <- igraph::graph_from_data_frame(disdf, directed = FALSE, vertices = milestone_ids)
mst <- igraph::minimum.spanning.tree(gr, weights = igraph::E(gr)$weight)

milestone_network <-
  igraph::as_data_frame(mst) %>%
  transmute(from, to, length = weight, directed = FALSE)

milestone_location <- centers %>% 
  as.data.frame() %>%
  rownames_to_column("milestone") %>% 
  janitor::clean_names()

milestone_network_ggplot <- milestone_network %>% 
  dplyr::select(milestone = from, -length, -directed) %>% 
  left_join(milestone_location, by = "milestone") %>% 
  dplyr::select(-milestone) %>% 
  cbind(milestone_network %>% 
  dplyr::select(milestone = to, -length, -directed) %>%
  left_join(milestone_location, by = "milestone") %>% 
    dplyr::select(v1_to = v1, v2_to = v2))
  

```

I plot the milestones on the previosly generated embedding. 

```{r}
df <- reducedDims(ods)$UMAP %>% cbind(colData(ods)) %>% cbind(clust_predict$classification) %>% as_tibble()

mclust_complete <- clust_predict

milestone_order <- c(1:6)
#milestone_order <- c(4,6,8,12,3,10,15,2,11,7,9,13,1,14,5)

gg_path <- clust$data %>%
  as.data.frame() %>%
  cbind(clust$classification) %>% 
  cbind(clust$uncertainty) %>% 
  janitor::clean_names() %>% 
  mutate(clust_uncertainty = (1-clust_uncertainty)*0.1,
         clust_classification = factor(clust_classification, levels = milestone_order)) %>%
  ggplot(aes(v1, v2, color = clust_classification)) + 
  geom_point_rast(size = 0.35, alpha = 0.1) + 
  # geom_segment(data = milestone_network_ggplot, aes(x = v1, y = v2, xend = v1_to, yend = v2_to), size = 0.7, color = "black") +
  # geom_point(data = milestone_location %>%
  #              mutate(x = substr(milestone, 2,2) %>% as.numeric()) %>%
  #              mutate(clust_classification = factor(x, levels = milestone_order)), alpha = 1,
  #            size = 1.3,
  #            color = "black") +
  
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  #theme(legend.position = "bottom") +
  theme(legend.position = "None") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) + 
  scale_color_manual(values = c("#6c80cc",
                                #"#6db13c", 
                                "#945fce",
                                #"#59c377", 
                                "#d34394",
                                #"#4a9e75", 
                                "#d0493a",
                                #"#45b0cf", 
                                "#db8f33",
                                #"#b76dac", 
                                "#b9af49"#,
                                #"#c05b6d", "#6f7a31",
                                #"#e0946e", "#9b5f2c"
                                )) 
  #scale_color_viridis_d() + 
  #scale_color_manual(values = scico::scico(15, palette = 'berlin')) + 
  #scale_color_manual(values = scico::scico(15, palette = 'roma')) 

#gg_path
```

I am interested in the names of the clusters 

```{r}
milestone_location %>%
               mutate(x = substr(milestone, 2,2) %>% as.numeric()) %>%
               mutate(clust_classification = factor(x, levels = milestone_order)) %>%
  ggplot(aes(v1,v2)) +
  geom_point(alpha = 1,
             size = 1.3,
             color = "black") + 
  
  
  geom_label(aes(label = milestone)) + 
  geom_segment(data = milestone_network_ggplot, aes(x = v1, y = v2, xend = v1_to, yend = v2_to), size = 0.7, color = "black") + 
  theme_classic()
```


It is hard to interpret the tree of clusters. For the following analysis I am going to use the clustering results and will create a barcode plot for each line and treatment. As the number of clusters is large (n > 15) I will introduce an order of clusters. 

```{r}
df_cluster <-reducedDims(ods)$UMAP %>% cbind(colData(ods)) %>% cbind(mclust_complete$classification) %>% as_tibble() %>% janitor::clean_names() %>% mutate(mclust_complete_classification = mclust_complete_classification %>% paste0("M", .) %>% factor(., levels = milestone_order %>% paste0("M", .)))
```

Now I have to validate how the predicted clusters look like for these maps. 

```{r}
gg_clust_predict <- df_cluster %>% 
  group_by(mclust_complete_classification) %>% 
  sample_n(100) %>% 
  ggplot(aes(v1, v2)) + 
  geom_point_rast(size = 1, alpha = 1) + 
  theme_cowplot() + 
  facet_wrap(~ mclust_complete_classification)
```


I visualize the proportion of clusters in a heatmap to identify similarities

```{r}
df_cluster %>% 
  group_by(line, mclust_complete_classification) %>% 
  summarize(n = n()) %>% 
  spread(mclust_complete_classification, n) %>% 
  as.data.frame() %>%
  column_to_rownames("line") %>% 
  pheatmap::pheatmap(scale = "row")
```


```{r}
clust_var <- df_cluster %>% 
  filter(drug == "DMSO") %>%
  group_by(line, mclust_complete_classification) %>% 
  summarize(n = n()) %>%
  group_by(line) %>% 
  mutate(prop = n/sum(n)) %>%
  dplyr::select(-n)  

clust_var_order <- clust_var %>% 
  group_by(mclust_complete_classification) %>% 
  summarise(var = var(prop)) %>% 
  arrange(desc(var)) %>%
  .$mclust_complete_classification %>% as.character()

clust_var %>% 
  group_by(mclust_complete_classification) %>% 
  summarise(var = var(prop)) %>% 
  arrange(desc(var)) %>% 
  mutate(mclust_complete_classification = factor(mclust_complete_classification, levels = clust_var_order)) %>%
  mutate(status = if_else(mclust_complete_classification %in% c("M14", "M13", "M1", "M5"), "dead", "alive")) %>%
  ggplot(aes(mclust_complete_classification, var, color = status)) + 
  geom_point() + 
  theme_cowplot()

# clust_var %>% 
#   mutate(mclust_complete_classification = factor(mclust_complete_classification, levels = clust_var_order)) %>%
#   
#   ggplot(aes(mclust_complete_classification, prop, color = status)) + 
#   geom_point()

# clust_mu_order <- clust_var %>% 
#   group_by(mclust_complete_classification) %>% 
#   summarise(mu = mean(prop)) %>% 
#   arrange(desc(mu)) %>%
#   .$mclust_complete_classification %>% as.character()
# 
# clust_var %>% 
#   mutate(mclust_complete_classification = factor(mclust_complete_classification, levels = clust_mu_order)) %>%
#   ggplot(aes(mclust_complete_classification, prop)) + 
#   geom_point()

line_order <- clust_var %>% 
  spread(mclust_complete_classification, prop) %>%
  janitor::clean_names() %>%
  arrange(desc(m5)) %>% 
  .$line
```

I write a csv with the proportions. 

```{r}
clust_var %>% 
  mutate(mclust_complete_classification = as.character(mclust_complete_classification)) %>%
  spread(mclust_complete_classification, prop) %>%
  janitor::clean_names() %>% 
  write_csv(here("data/cluster_proportions.csv"))
```


I associate proportions with a cluster of interest, M4

```{r}
clust_var %>% 
  filter(mclust_complete_classification == "M4") %>% 
  filter(line != "D021T01") %>% 
  mutate(pi3k_akt = if_else(line %in% c("D007T01", "D004T01",
                                        "D013T01", "D030T01",
                                        "D054T01"), TRUE, FALSE)) %>% 
  ggplot(aes(pi3k_akt, prop)) +
  geom_jitter(width = 0.2)+ 
  theme_cowplot() + 
  ggsignif::geom_signif(comparisons = list(c("TRUE", "FALSE")))
```

It is hard to define a mutation status as false given the limited coverage of our amplicon library.

```{r, eval = FALSE}
gg_proportion <- df_cluster %>% 
  mutate(line = factor(line, levels = line_order)) %>%
  #filter(!(line %in% c("D054T01", "D021T01", "D020T01", "D010T01"))) %>%
  filter(!(line %in% c("D020T02", "D010T01"))) %>%
  filter(drug == "DMSO") %>%
  ggplot(aes(line, fill = mclust_complete_classification)) + 
  geom_bar(position = "fill") + 
  theme_classic() + 
  coord_flip() + 
  #theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) + 
  scale_fill_manual(values = scico::scico(15, palette = 'roma')) +
  #facet_wrap(~ Drug)
  theme(legend.text = element_blank()) + 
  theme(legend.position = "None")

gg_proportion
```

Now I am interested how certain drugs and their concentrations shift the composition of morphological phenotypes

```{r, eval = FALSE}
drug_order <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Bortezomib")) %>%
    filter(line == "D046T01") %>% .$drug %>% unique()
drug_order <- drug_order[rev(c(3,2,5,4,6,1))]


gg_borte_prop <- df_cluster %>% 
  mutate(line = factor(line, levels = line_order)) %>%
  #filter(!(line %in% c("D054T01", "D021T01", "D020T01", "D010T01"))) %>%
  filter(!(line %in% c("D021T01", "D010T01"))) %>%
  filter(drug == "DMSO" | grepl(drug, pattern = "Bortezomib")) %>%
  filter(!grepl(drug, pattern = "Dabrafenib")) %>% 
  filter(line == "D046T01") %>%
  mutate(drug = factor(drug, levels =drug_order)) %>%
  ggplot(aes(drug, fill = mclust_complete_classification)) + 
  geom_bar(position = "fill") + 
  theme_classic() + 
  coord_flip() + 
  #theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) + 
  scale_fill_manual(values = scico::scico(15, palette = 'roma')) +
  #facet_wrap(~ concentration) +
  theme(legend.text = element_blank()) + 
  theme(legend.position = "None")
```

I visualize this transition in the UMAP.

```{r}
set.seed(123)

drug_order <- df_cluster %>%
    #filter(drug == "DMSO" | grepl(drug, pattern = "Bortezomib")) %>%
  filter(grepl(drug, pattern = "Bortezomib")) %>%
    filter(line == "D046T01") %>% .$drug %>% unique()
drug_order <- drug_order[rev(c(3,2,5,4,6,1))]

gg_bortezomib <- df_cluster %>% sample_frac(0.01) %>%
  dplyr::select(-line, -concentration) %>%
  ggplot(aes(v1, v2)) + 
  geom_point_rast(alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point_rast(data = df_cluster %>%
    #filter(drug == "DMSO" | grepl(drug, pattern = "Bortezomib")) %>%
      filter(grepl(drug, pattern = "Bortezomib")) %>%
  filter(line %in% c("D055T01", "D054T01", "D046T01")) %>%
    mutate(concentration = if_else(concentration == "nan", "0", concentration)) %>%
    mutate(drug = factor(drug, levels =drug_order)) %>%
    
    group_by(drug) %>% 
    sample_n(5000, replace = TRUE),
  aes(color = drug),alpha = 1, size = 1.5, shape=16, color = "black") + 
  facet_wrap( ~ concentration, ncol = 1) + 
  #scale_color_brewer(type = "seq", palette = "YlOrRd") + 
  #geom_density2d(color = "black") + 
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) + 
  theme_cowplot(font_size = 8) + 
  theme(legend.position = "bottom")
  


gg_bortezomib
```


I visualize the location of Bortezomib 

```{r}
drug_order <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Bortezomib")) %>%
    filter(line == "D046T01") %>% .$drug %>% unique()
drug_order <- drug_order[rev(c(3,2,5,4,6,1))]


center_df <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Bortezomib")) %>%
  # filter(line == "D046T01") %>% 
    mutate(drug = factor(drug, levels =drug_order)) %>%
    # group_by(drug) %>% 
    # sample_n(200, replace = FALSE) %>% 
  dplyr::select(v1, v2, drug, line) %>% 
  group_by(drug, line) %>% 
  summarize(v1 = mean(v1),
            v2 = mean(v2))

gg_bortezomib_path <- center_df %>%
  ggplot(aes(v1, v2, group = line)) + 
  
  scale_color_brewer(type = "seq", palette = "YlOrRd") + 
  geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point(aes(color = drug),alpha = 1, size = 3, shape=16) + 
  geom_path() +
  facet_wrap(~line ) +
  #geom_density2d(color = "black") + 
  theme_bw() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) 
```


I repeat the same figure with Trametinib

```{r}
df_cluster %>% 
  mutate(line = factor(line, levels = line_order)) %>%
  #filter(!(line %in% c("D054T01", "D021T01", "D020T01", "D010T01"))) %>%
  filter(!(line %in% c("D021T01", "D010T01"))) %>%
  filter(drug == "DMSO" | grepl(drug, pattern = "Trametinib")) %>%
  filter(!grepl(drug, pattern = "Dabrafenib")) %>% 
  filter(!grepl(drug, pattern = "GSK1120212")) %>%
  filter(line == "D046T01") %>%
  ggplot(aes(drug, fill = mclust_complete_classification)) + 
  geom_bar(position = "fill") + 
  theme_classic() + 
  coord_flip() + 
  #theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) + 
  scale_fill_manual(values = scico::scico(15, palette = 'roma')) +
  #facet_wrap(~ concentration) +
  theme(legend.text = element_blank()) + 
  theme(legend.position = "None")
```


```{r}
drug_order <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Trametinib")) %>%
  filter(!grepl(drug, pattern = "Dabrafenib")) %>% 
  filter(!grepl(drug, pattern = "GSK1120212")) %>% 
    filter(line == "D046T01") %>% .$drug %>% unique() %>% sort()


gg_trametinib <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Trametinib")) %>%
      filter(!grepl(drug, pattern = "Dabrafenib")) %>% 
       filter(!grepl(drug, pattern = "GSK1120212")) %>%
  filter(line == "D046T01") %>% 
    mutate(drug = factor(drug, levels =drug_order)) %>%
    group_by(drug) %>% 
    sample_n(200, replace = FALSE) %>%
  #sample_n(1000, replace = FALSE) %>%
  ggplot(aes(v1, v2)) + 
  
  geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_density2d(aes(color = drug)) +
  geom_point_rast(aes(color = drug),alpha = 1, size = 1, shape=16) + 
  facet_wrap(~ drug) + 
  scale_color_brewer(type = "seq", palette = "YlOrRd") + 
  #geom_density2d(color = "black") + 
  theme_bw() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) 

gg_trametinib
gg_trametinib + ggsave("trametinib.pdf", width = 15, height = 10)

```

```{r}
gg_trametinib_all <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Trametinib")) %>%
      filter(!grepl(drug, pattern = "Dabrafenib")) %>% 
      filter(grepl(drug, pattern = "GSK1120212")) %>%
  # #filter(line == "D046T01") %>% 
  #   mutate(drug = factor(drug, levels =drug_order)) %>%
  #   group_by(drug) %>% 
  #   sample_n(200, replace = FALSE) %>%
  #sample_n(1000, replace = FALSE) %>%
  ggplot(aes(v1, v2)) + 
  
  geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  #geom_density2d(aes(color = drug)) +
  geom_point_rast(alpha = 1, size = 0.35, shape=16, color = "black") + 
  #facet_wrap(~ drug) + 
  #scale_color_brewer(type = "seq", palette = "YlOrRd") + 
  #geom_density2d(color = "black") + 
  theme_bw() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) + 
  theme_cowplot() + theme(legend.position = "nothing")

gg_mtor_all <- df_cluster %>%
    #filter(drug == "DMSO" | grepl(drug, pattern = "WYE-132") | drug %in% c("PF-04691502", "PP121")) %>%
  filter(drug == "DMSO" | grepl(drug, pattern = "WYE-132")) %>%
  filter(line %in% c("D030T01", "D027T01", "D013T01")) %>%
  #filter(line %in% c("D055T01", "D054T01", "D046T01")) %>%
  group_by(drug) %>% 
  sample_n(700) %>%
  ggplot(aes(v1, v2)) + 
  
  geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  #geom_density2d(aes(color = drug)) +
  geom_point_rast(aes(color = drug),alpha = 1, size = 1, shape=16, color = "black") + 
  facet_wrap(~ drug) + 
  #scale_color_brewer(type = "seq", palette = "YlOrRd") + 
  #geom_density2d(color = "black") + 
  theme_bw() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) + 
  theme_cowplot(font_size = 8) + 
  theme(legend.position = "nothing")


gg_gski_all <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "CHIR-98014")) %>%
  filter(line %in% c("D055T01", "D054T01", "D046T01")) %>%
  group_by(drug) %>% 
  sample_n(1380) %>%
  ggplot(aes(v1, v2)) + 
  
  geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  #geom_density2d(aes(color = drug)) +
  geom_point_rast(aes(color = drug),alpha = 1, size = 1, shape=16, color = "black") + 
  facet_wrap(~ drug) + 
  #scale_color_brewer(type = "seq", palette = "YlOrRd") + 
  #geom_density2d(color = "black") + 
  theme_bw() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) + 
  theme_cowplot(font_size = 8) + 
  theme(legend.position = "nothing")

```


I need to visulaize these phenotypes in a better way. Thus I calculate the 2D median.

```{r}
drug_order <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Trametinib")) %>%
  filter(!grepl(drug, pattern = "Dabrafenib")) %>% 
  filter(!grepl(drug, pattern = "GSK1120212")) %>% 
    filter(line == "D046T01") %>% .$drug %>% unique() %>% sort()


center_df <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Trametinib")) %>%
      filter(!grepl(drug, pattern = "Dabrafenib")) %>% 
       filter(!grepl(drug, pattern = "GSK1120212")) %>%
  #filter(line == "D055T01") %>% 
    mutate(drug = factor(drug, levels =drug_order)) %>%
    # group_by(drug) %>% 
    # sample_n(200, replace = FALSE) %>% 
  dplyr::select(v1, v2, drug, line) %>% 
  group_by(drug, line) %>% 
  summarize(v1 = mean(v1),
            v2 = mean(v2))

gg_trametinib_path <- center_df %>%
  ggplot(aes(v1, v2, group = line)) + 
  
  scale_color_brewer(type = "seq", palette = "YlOrRd") + 
  geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point(aes(color = drug),alpha = 1, size = 3, shape=16) + 
  geom_path() +
  facet_wrap(~ line) +
  #geom_density2d(color = "black") + 
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) 
```

I want to create smoothed principal curves like in most trajectory inference methods, for example slingshot. 
This is the trajectory for Bortezomib:

```{r}
  


drug_order <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Bortezomib")) %>%
 .$drug %>% unique() %>% sort()

df_tmp <- df_cluster %>%
    filter(drug %in% drug_order) %>% 
  filter(line == "D055T01") %>%
  mutate(drug = factor(drug, levels =drug_order)) %>% 
  arrange((drug)) %>% 
  group_by(drug) %>%
  sample_n(1000, replace = TRUE) %>% 
  ungroup()

df_tmp_init <- df_tmp %>% 
  group_by(drug, line) %>% 
  summarize(v1 = mean(v1),
            v2 = mean(v2)) %>% 
  ungroup()

plot(df_tmp_init%>% dplyr::select(v1, v2) %>% as.matrix())
plot(df_tmp%>% dplyr::select(v1, v2) %>% as.matrix())

fit <- principal_curve(x = df_tmp %>% dplyr::select(v1, v2) %>% as.matrix(),
                       start = df_tmp_init %>% dplyr::select(v1, v2) %>% as.matrix(),
                       approx_points = FALSE, 
                       trace = TRUE, 
                       plot_iterations = TRUE, 
                       maxit = 30,
                       stretch =0)

plot(fit)
points(df_tmp_init%>% dplyr::select(v1, v2) %>% as.matrix())

fit$s[fit$ord,] %>% .[c(500:5000),] %>%
  as_tibble() %>%
  ggplot(aes(v1, v2)) + 
  geom_path() + 
  geom_point(data = df_tmp_init, size = 2)



```

Now I am building a function to create these tidy path objects.

```{r}
create_princurve_trace <- function(df){
  df_tmp_init <- df %>% 
  group_by(drug) %>% 
  summarize(v1 = mean(v1),
            v2 = mean(v2)) %>% 
  ungroup()
  
  fit <- principal_curve(x = df %>% dplyr::select(v1, v2) %>% as.matrix(),
                       start = df_tmp_init %>% dplyr::select(v1, v2) %>% as.matrix(),
                       approx_points = FALSE, 
                       trace = TRUE, 
                       plot_iterations = TRUE, 
                       maxit = 30,
                       stretch =2)
  
  plot(fit)
  points(df_tmp_init%>% dplyr::select(v1, v2) %>% as.matrix())

  result <- list()
  
  result$fit <- fit$s[fit$ord,] %>% as_tibble()
  result$center <- df_tmp_init
  result$input <- df
  
  return(result)
}
```


```{r}
gg_bortezomib_path + ggsave("bortezomib_path.pdf", width = 12, height =  7)
gg_trametinib_path + ggsave("trametininb_path.pdf", width = 12, height =  7)
```

It becomes clear that compounds have characteristic paths of drug response. 
To illustrate this, we focus on the organoids that start out with very disorganized archtecture and compare their response to bortezomib vs. trametinib. 


```{r}
drug_order <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Trametinib")| grepl(drug, pattern = "Bortezomib")) %>%
  filter(!grepl(drug, pattern = "Dabrafenib")) %>% 
  filter(!grepl(drug, pattern = "GSK1120212")) %>% .$drug %>% unique() %>% sort()

# center_df_mean <- df_cluster %>%
#     filter(drug %in% drug_order) %>%
#   filter(line %in% c("D055T01", "D054T01", "D046T01")) %>%
#   dplyr::select(v1, v2, drug, line) %>% 
#   group_by(drug) %>% 
#   summarize(v1 = mean(v1),
#             v2 = mean(v2)) %>% 
#   separate(drug, c("drug", "concentration"), sep = "__") %>% 
#   mutate(concentration = as.numeric(concentration))

center_df <- df_cluster %>%
    filter(drug %in% drug_order) %>%
  filter(line %in% c("D055T01", "D054T01", "D046T01")) %>%
  dplyr::select(v1, v2, drug, line) %>% 
  group_by(drug, line) %>% 
  summarize(v1 = mean(v1),
            v2 = mean(v2)) %>% 
  separate(drug, c("drug", "concentration"), sep = "__") %>% 
  mutate(concentration = as.numeric(concentration),
         drug_line = paste0(drug, "_", line))

gg_trambort <- center_df %>% 
  ggplot(aes(v1, v2)) + 
  geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point(aes(color = drug, group = drug_line),alpha = 1, size = 1.5, shape=16) + 
  #geom_smooth(aes(group = drug), se = FALSE) + 
  #geom_path(aes(group = drug, color = drug), arrow = arrow(angle =15, ends = "last", type = "closed", length = unit(0.15, "inches"))) + 
  geom_path(aes(group = drug_line, color = drug)) + 
  scale_color_manual(values = c("#707070", "#000000", "#4285F4")) +
  
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) 
```


I am creating a new version of the plot above

```{r}
drug_order <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Trametinib")| grepl(drug, pattern = "Bortezomib")) %>%
  filter(!grepl(drug, pattern = "Dabrafenib")) %>% 
  filter(!grepl(drug, pattern = "GSK1120212")) %>% .$drug %>% unique() %>% sort()

center_df <- df_cluster %>%
    filter(drug %in% drug_order) %>%
  filter(line %in% c("D055T01", "D054T01", "D046T01")) %>%
  mutate(drug = factor(drug, levels =drug_order)) %>% 
  arrange((drug)) %>% 
  group_by(drug, line) %>%
  sample_n(1000, replace = TRUE) %>% ungroup()

plot_df_bort <- center_df %>%
  filter(drug == "DMSO" | grepl(drug, pattern = "Bortezomib")) %>%
  filter(grepl(drug, pattern = "Bortezomib")) %>%
  nest(-line) %>%
  mutate(plot = purrr::map(data, ~ create_princurve_trace(.x)))

plot_df_tram <- center_df %>%
  filter(drug == "DMSO" | grepl(drug, pattern = "Trametinib")) %>% 
  filter(grepl(drug, pattern = "Trametinib")) %>%
  nest(-line) %>%
  mutate(plot = purrr::map(data, ~ create_princurve_trace(.x)))

gg_trambort_traj <- plot_df_tram %>% mutate(plot_trace = purrr::map(plot, ~ .x$fit %>% .[c(300:4500), ])) %>% unnest(plot_trace) %>%
  mutate(drug = "Trametinib") %>% 
  rbind(plot_df_bort %>% mutate(plot_trace = purrr::map(plot, ~ .x$fit %>% .[c(300:4000), ])) %>% unnest(plot_trace) %>%
  mutate(drug = "Bortezomib")) %>% 
  mutate(drug_line = paste0(drug, "_", line)) %>%
  filter(line == "D046T01") %>%
  ggplot(aes(v1, v2)) + 
  #geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  #geom_point(aes(color = drug, group = drug_line),alpha = 1, size = 1.5, shape=16) + 
  #geom_smooth(aes(group = drug), se = FALSE) + 
  geom_point(data = 
               plot_df_bort %>% unnest(data) %>% 
               rbind(plot_df_tram %>% unnest(data)) %>% 
               separate(drug, c("drug", "conc"), sep = "__"),
             aes(color = drug), alpha = 0.05, size = 0.35) +
  
  #geom_path(aes(group = drug, color = drug), arrow = arrow(angle =15, ends = "last", type = "closed", length = unit(0.15, "inches"))) + 
  geom_path(aes(group = drug_line, color = drug), size = 1.5, arrow = arrow(angle = 10, ends = "last", type = "closed", length = unit(0.15, "inches"))) + 
  scale_color_manual(values = c("#707070", "#4285F4")) +
  facet_grid(~ drug) +
  # geom_point(data = 
  #              plot_df_bort %>% mutate(center = map(plot, ~ .x$center)) %>% unnest(center) %>% 
  #              rbind(plot_df_tram %>% mutate(center = map(plot, ~ .x$center)) %>% unnest(center)) %>% 
  #              separate(drug, c("drug", "conc"), sep = "__"),
  #            aes(color = drug)) +
  
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) 
```

I create the same figure for Taxanes

```{r}
set.seed(12312)
drug_order <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Docetaxel")| grepl(drug, pattern = "Paclitaxel")) %>%
.$drug %>% unique() %>% sort()

center_df <- df_cluster %>%
    filter(drug %in% drug_order) %>%
  filter(line %in% c("D013T01", "D027T01")) %>%
  mutate(drug = factor(drug, levels =drug_order)) %>% 
  arrange((drug)) %>% 
  group_by(drug, line) %>%
  sample_n(1000, replace = TRUE) %>% ungroup()

plot_df_pacli <- center_df %>%
  filter(grepl(drug, pattern = "Paclitaxel")) %>%
  nest(-line) %>%
  mutate(plot = purrr::map(data, ~ create_princurve_trace(.x)))

plot_df_doce <- center_df %>%
  filter(grepl(drug, pattern = "Docetaxel")) %>%
  nest(-line) %>%
  mutate(plot = purrr::map(data, ~ create_princurve_trace(.x)))

gg_taxane_traj <- plot_df_pacli %>% mutate(plot_trace = purrr::map(plot, ~ .x$fit %>% .[c(300:4500), ])) %>% unnest(plot_trace) %>%
  mutate(drug = "Paclitaxel") %>% 
  rbind(plot_df_doce %>% mutate(plot_trace = purrr::map(plot, ~ .x$fit %>% .[c(300:4000), ])) %>% unnest(plot_trace) %>%
  mutate(drug = "Docetaxel")) %>% 
  mutate(drug_line = paste0(drug, "_", line)) %>%
  
  ggplot(aes(v1, v2)) + 
  geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  #geom_point(aes(color = drug, group = drug_line),alpha = 1, size = 1.5, shape=16) + 
  #geom_smooth(aes(group = drug), se = FALSE) + 
  geom_point(data = 
               plot_df_pacli %>% unnest(data) %>% 
               rbind(plot_df_doce %>% unnest(data)) %>% 
               separate(drug, c("drug", "conc"), sep = "__"),
             aes(color = drug), alpha = 0.05, size = 0.35) +
  
  #geom_path(aes(group = drug, color = drug), arrow = arrow(angle =15, ends = "last", type = "closed", length = unit(0.15, "inches"))) + 
  geom_path(aes(group = drug_line, color = drug), size = 1.5, arrow = arrow(angle = 10, ends = "last", type = "closed", length = unit(0.15, "inches"))) + 
  scale_color_manual(values = c("#F4B400", "#0F9D58")) +
  facet_grid(~ drug) +
  # geom_point(data = 
  #              plot_df_bort %>% mutate(center = map(plot, ~ .x$center)) %>% unnest(center) %>% 
  #              rbind(plot_df_tram %>% mutate(center = map(plot, ~ .x$center)) %>% unnest(center)) %>% 
  #              separate(drug, c("drug", "conc"), sep = "__"),
  #            aes(color = drug)) +
  
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) 
```



I am also interested in the OSI induced phenotype 

```{r}
drug_order <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Paclitaxel")) %>%
  #  filter(line == "D046T01") %>%
  .$drug %>% unique() %>% sort()

center_df <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Paclitaxel")) %>%
  #filter(line == "D046T01") %>% 
    mutate(drug = factor(drug, levels =drug_order)) %>%
    # group_by(drug) %>% 
    # sample_n(200, replace = FALSE) %>% 
  dplyr::select(v1, v2, drug, line) %>% 
  group_by(drug, line) %>% 
  summarize(v1 = median(v1),
            v2 = median(v2))

center_df %>%
  ggplot(aes(v1, v2, group = line)) + 
  
  scale_color_brewer(type = "seq", palette = "YlOrRd") + 
  geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point(aes(color = drug),alpha = 1, size = 3, shape=16) + 
  #facet_wrap(~ line) +
  #geom_path() +
  #geom_density2d(color = "black") + 
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) 
```

I also check out OSI-953

```{r}
drug_order <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "ONO-4059")) %>%
  #  filter(line == "D046T01") %>%
  .$drug %>% unique() %>% sort()

center_df <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "ONO-4059")) %>%
  #filter(line == "D046T01") %>% 
    mutate(drug = factor(drug, levels =drug_order)) %>%
    # group_by(drug) %>% 
    # sample_n(200, replace = FALSE) %>% 
  dplyr::select(v1, v2, drug, line) %>% 
  group_by(drug, line) %>% 
  summarize(v1 = median(v1),
            v2 = median(v2))

gg_ono <- center_df %>%
  filter(!(line %in% c("D021T01", "D010T01", "D018T01"))) %>%
  ggplot(aes(v1, v2, group = line)) + 
 
  scale_color_brewer(type = "seq", palette = "YlOrRd") + 
  geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  #geom_point(aes(color = drug),alpha = 1, size = 3, shape=16) + 
  # facet_wrap(~ line) +
  
  geom_line(aes(group = line), arrow = arrow(angle =15, ends = "first", type = "closed", length = unit(0.1, "inches")), color = "grey") +
  geom_line(data = center_df %>% filter(line == "D027T01"), aes(group = line), arrow = arrow(angle =15, ends = "first", type = "closed", length = unit(0.15, "inches")), color = "black") +
  #gghighlight::gghighlight(line == "D027T01") +
  #geom_density2d(color = "black") + 
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) 
```

I also add GSK inhibitors 

```{r}
drug_order <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "CHIR-99021")) %>%
  #  filter(line == "D046T01") %>%
  .$drug %>% unique() %>% sort()

center_df <- df_cluster %>%
    filter(drug %in% drug_order) %>%
  #filter(line == "D046T01") %>% 
  mutate(drug = if_else(drug == "DMSO", "DMSO", "Chir")) %>%
    mutate(drug = factor(drug)) %>%
    # group_by(drug) %>% 
    # sample_n(200, replace = FALSE) %>% 
  dplyr::select(v1, v2, drug, line) %>% 
  group_by(drug, line) %>% 
  summarize(v1 = median(v1),
            v2 = median(v2))

gg_chir <- center_df %>%
  filter(!(line %in% c("D021T01", "D010T01", "D018T01"))) %>%
  ggplot(aes(v1, v2, group = line)) + 
 
  scale_color_brewer(type = "seq", palette = "YlOrRd") + 
  geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  #geom_point(aes(color = drug),alpha = 1, size = 3, shape=16) + 
  # facet_wrap(~ line) +
  
  geom_line(aes(group = line), arrow = arrow(angle =15, ends = "first", type = "closed", length = unit(0.1, "inches")), color = "grey") +
  geom_line(data = center_df %>% filter(line == "D027T01"), aes(group = line), arrow = arrow(angle =15, ends = "first", type = "closed", length = unit(0.15, "inches")), color = "black") +
  #gghighlight::gghighlight(line == "D027T01") +
  #geom_density2d(color = "black") + 
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) 
```

# Example image generation. 

Based on the clustering I have produced, I want to visualize example objects belonging to each cluster. 

```{r}
source(here("R/display_organoid_complete.R"))
```

Now I create a list of organoids I am interested in. 

```{r}

```


# Cluster enrichment 

## Size per cluster 

```{r}
df_cluster %>% 
  dplyr::select(mclust_complete_classification, size) %>% 
  group_by(mclust_complete_classification) %>%
  summarise(size = mean(size)) %>% 
  arrange(size) %>% 
  mutate(mclust_complete_classification = factor(mclust_complete_classification) %>% fct_inorder()) %>% 
  ggplot(aes(mclust_complete_classification, size)) + 
  geom_point() + 
  theme_cowplot()
```

## Example organoids 

I tidy the membership porbability table. 

```{r, eval = FALSE}
score_df <- clust_predict$z %>% as.data.frame() %>% 
  rownames_to_column("id") %>% mutate(id = as.numeric(id)) %>% gather(m, prob, -id) %>% nest(-id) %>% mutate(score = furrr::future_map(data, ~ arrange(.x, desc(prob)) %>% head(1))) %>% 
  unnest(score) %>% dplyr::select(-data)
```


```{r}
library(imager)
html_dir = "/data/rauscher/promise/PROMISE/data-10x-4t-c-16z/htmldir/"

hdf5_all_uwot_stereotypes <- df_cluster%>%
  mutate(OriginalX = as.numeric(original_x),
         OriginalY = as.numeric(original_y),
         Field = field) %>%
  # I scale the information to the image's size and add 50% of the y-dim to target the first field
    mutate(ScaleX = round((OriginalX/2035.5)*(514/2)),
         ScaleY = round((OriginalY/2035.5)*(514/2))) %>%
  mutate(ScaleX = case_when(Field %in% c("1", "4") ~ ScaleX,
                            Field %in% c("2", "3") ~ ScaleX + (514/2)),
         ScaleY = case_when(Field %in% c("1", "2") ~ ScaleY,
                            Field %in% c("3", "4") ~ ScaleY + (514/2))) %>%
  mutate(lowerX = ScaleX - round(sqrt(as.numeric(size))/4),
         upperX = ScaleX + round(sqrt(as.numeric(size))/4), 
         lowerY = ScaleY - round(sqrt(as.numeric(size))/4), 
         upperY = ScaleY + round(sqrt(as.numeric(size))/4)) %>%
  filter(lowerX > 0,
         lowerY > 0,
         upperX < 514,
         upperY < 514) %>%
  group_by(mclust_complete_classification) %>% 
  #do(arrange(., desc(prob)) %>% head(10)) %>% 
    sample_n(10) %>%
  mutate(path_to_html = paste0(html_dir, bc, "/", bc, "_", substr(Well, 1,1), "_", substr(Well, 2,3), "_1.jpeg")) 

hdf5_all_uwot_stereotypes_img <- hdf5_all_uwot_stereotypes %>% ungroup() %>%
  #mutate(img_raw = purrr::map(path_to_html, ~ imager::load.image(.x))) %>%
  # mutate(img = purrr::map(img_raw, ~ .x %>% as.data.frame(wide = "c") %>%
  #          mutate(rgb.val = rgb(c.1, c.2, c.3))))
  mutate(img_raw = purrr::pmap(list(p = path_to_html, a =lowerX, b = upperX, c = lowerY, d= upperY, e = bc, f = cluster, g = index, h = score), function(p, a,b,c,d, e,f,g, h) EBImage::readImage(p) %>% 
                                 .[c(a:b), c(c:d),] %>%
                                 EBImage::writeImage(., files = here(paste0("data/cluster_tiles/", f, "_", e, "_", g, "_", round(h, 2), ".jpeg")))))
  

lapply(hdf5_all_uwot_stereotypes_img$img_raw, EBImage::display, method = "raster")

```



## Drug target

I wonder wether certain drugs are enriched in the 2D space I am observing

First I save my results.

```{r}
saveRDS(df_cluster, here("data/clustered_umap.Rds"))
```

Then I execute my contingency analysis

```{r, eval = FALSE}
chisq_drug <- df_cluster %>% 
  dplyr::select(mclust_complete_classification, drug) %>%
  filter(drug != "DMSO") %>%
  filter(drug != "Staurosporine_500nM") %>%
  #mutate(mclust_complete_classification = as.numeric(mclust_complete_classification)) %>%
  group_by(mclust_complete_classification, drug) %>% 
  summarize(n = n()) %>% 
  spread(drug, n, fill = 0) %>%
  as.data.frame() %>%
  column_to_rownames("mclust_complete_classification") %>% 
  as.matrix() %>%
  chisq.test()
  #stats::fisher.test(simulate.p.value = TRUE)
```

# Chi-Square 

```{r}
sd <- chisq_drug$residuals %>% as.matrix() %>% sd()

chi_tidy <- chisq_drug$residuals %>% as.data.frame() %>% 
  rownames_to_column("cluster") %>% 
  gather(drug, residual, -cluster)

chi_tidy %>% group_by(cluster) %>% 
  arrange(desc(residual)) %>% filter(cluster == "M1") #dead
```

```{r}
chi_tidy %>% group_by(cluster) %>% 
  arrange(desc(residual)) %>% filter(cluster == "M2")
```

```{r}
chi_tidy %>% group_by(cluster) %>% 
  arrange(desc(residual)) %>% filter(cluster == "M3") # dead
```

```{r}
chi_tidy %>% group_by(cluster) %>% 
  arrange(desc(residual)) %>% filter(cluster == "M4") # dense morphology #PF-04691502 #PP121 
```

```{r}
chi_tidy %>% group_by(cluster) %>% 
  arrange(desc(residual)) %>% filter(cluster == "M5") #dead
```

```{r}
chi_tidy %>% group_by(cluster) %>% 
  arrange(desc(residual)) %>% filter(cluster == "M6")
```


# Non- Chi-Sqaure

I explore the distribution of counts for each replicate

```{r}
df_drug_count <- df_cluster %>% 
  dplyr::select(mclust_complete_classification, drug, replicate) %>%
  filter(drug != "DMSO") %>%
  filter(drug != "Staurosporine_500nM") %>%
  filter(!grepl(drug, pattern = "__")) %>%
  group_by(mclust_complete_classification, drug, replicate) %>% 
  summarize(n = n()) 
```

I examine the dispersion of count distributions for each cluster across samples. Given the heterogeneity of dispersions, it might be useful to perform differential abundance analysis in a cluster-by-cluster fashion, as qCML assumes multiple small samples with similar dispesion.

```{r}
df_drug_count %>% 
  ggplot(aes(n, color = mclust_complete_classification)) + 
  geom_density() + 
  scale_x_log10() +
  theme_cowplot() + 
  scale_color_viridis_d()
```

I log transform the count data and plot histograms another time. As this is not sequencing data, there is no need to perform RNA-Seq specific preprocessing steps, such as depth and read-length adjustments. 

```{r}
df_drug_count %>%
  mutate(log_n = log(n)) %>% 
  ggplot(aes(mclust_complete_classification, log_n)) + 
  geom_boxplot()
```

Next, I remove drugs with a count below 50, as no statistical test would be useful in this case. It turns out that there are no drugs that do not meet this threshold.

```{r}
df_drug_count %>%
  ungroup() %>%
  group_by(drug) %>% 
  summarise(n = sum(n)) %>%
  arrange(n)
```

I continue to import the data into an biocodnuctor format to perform further processing. 

```{r}
library(edgeR)

df_drug_count_matrix <- df_drug_count %>%
  mutate(log_n = log(n)) %>% 
  ungroup() %>%
  mutate(cluster_replicate = paste0(as.character(mclust_complete_classification), "_", replicate)) %>%
  dplyr::select(-log_n, -mclust_complete_classification, -replicate) %>%
  spread(drug, n, fill = 0) %>% 
  as.data.frame() %>%
  column_to_rownames("cluster_replicate") %>% 
  t() %>% as.matrix()
  
dda <- DGEList(df_drug_count_matrix, 
        lib.size = NULL,
        group = rep(c(1,10,11,12,13,14,15, 2:9),each=2))
```

Now I follow thorugh a standard multi-group comparison workflow using edgeR.

```{r}
dda_norm <- calcNormFactors(dda)  
dda_norm$samples$norm.factors
lcpm_norm <- cpm(dda_norm, log=TRUE)
```

I compare the normalized data to unnormalized data. 

```{r}
lcpm <- cpm(dda, log=TRUE)

lcpm %>% as_tibble() %>% 
  gather(cluster, value) %>%
  mutate(normalization = FALSE) %>% 
  rbind(lcpm_norm %>% as_tibble() %>%
  gather(cluster, value) %>%
  mutate(normalization = TRUE)) %>% 
  ggplot(aes(value, color = cluster)) + 
  geom_density() + 
  facet_wrap(~ normalization) +
  scale_color_viridis_d() +
  theme_cowplot()
```

I create a small overview

```{r}
plotMDS(lcpm_norm)
```

Based on perturbations alone, we can identify clusters with similar patterns.

Now I run a test. 

```{r}
cluster <- factor(paste0("M", rep(c(1,10,11,12,13,14,15, 2:9),each=2)))
replicate <- factor(rep(c(1:2), times = 15))

design <- model.matrix(~ 0 + cluster + replicate)

colnames(design) <- c(paste0("M", rep(c(1,10,11,12,13,14,15, 2:9),each=1)), "replicate")

contrast_matrix <- makeContrasts(#"m1"=M1-(M2+M3+M4+M5+M6+M7+M8+M9+M10+M11+M12+M13+M14+M15)/14,
                            "m2"=M2-(M1+M3+M4+M5+M6+M7+M8+M9+M10+M11+M12+M13+M14+M15)/14,
                            "m3"=M3-(M1+M2+M4+M5+M6+M7+M8+M9+M10+M11+M12+M13+M14+M15)/14,
                           #"m4"=M4-(M1+M2+M3+M5+M6+M7+M8+M9+M10+M11+M12+M13+M14+M15)/14,
                           #"m5"=M5-(M1+M2+M3+M4+M6+M7+M8+M9+M10+M11+M12+M13+M14+M15)/14,
                           "m6"=M6-(M1+M2+M3+M4+M5+M7+M8+M9+M10+M11+M12+M13+M14+M15)/14,
                           #"m7"=M7-(M1+M2+M3+M4+M5+M6+M8+M9+M10+M11+M12+M13+M14+M15)/14,
                            "m8"=M8-(M1+M2+M3+M4+M5+M6+M7+M9+M10+M11+M12+M13+M14+M15)/14,
                           # "m9"=M9-(M1+M2+M3+M4+M5+M6+M7+M8+M10+M11+M12+M13+M14+M15)/14,
                           "m10"=M10-(M1+M2+M3+M4+M5+M6+M7+M8+M9+M11+M12+M13+M14+M15)/14,
                           #"m11"=M11-(M1+M2+M3+M4+M5+M6+M7+M8+M9+M10+M12+M13+M14+M15)/14,
                           "m12"=M12-(M1+M2+M3+M4+M5+M6+M7+M8+M9+M10+M11+M13+M14+M15)/14,
                           # "m13"=M13-(M1+M2+M3+M4+M5+M6+M7+M8+M9+M10+M11+M12+M14+M15)/14,
                           # "m14"=M12-(M1+M2+M3+M4+M5+M6+M7+M8+M9+M10+M11+M12+M13+M15)/14,
                            "m15"=M13-(M1+M2+M3+M4+M5+M6+M7+M8+M9+M10+M11+M12+M13+M14)/14,
                           levels=design)
```

I run a test

```{r}
par(mfrow=c(1,2))
v <- voom(dda, design, plot=TRUE)
vfit <- lmFit(v, design)
vfit <- contrasts.fit(vfit, contrasts=contrast_matrix)
efit <- eBayes(vfit)
plotSA(efit, main="Final model: Mean-variance trend")
```

Now I check the results. 

```{r}
df <- topTable(efit, number=Inf, coef= "m15") %>% as.data.frame() %>% rownames_to_column("drug")%>% as_tibble() %>% mutate(log_p = log(adj.P.Val))

df %>%
  ggplot(aes(t, log_p)) + 
  geom_point()
```

```{r}
df <- topTable(efit, number=Inf) %>% as.data.frame() %>% rownames_to_column("drug") %>% filter(adj.P.Val < 0.001) %>% dplyr::select(-(AveExpr:adj.P.Val)) %>% 
  column_to_rownames("drug") 

df %>%
  pheatmap::pheatmap(filename = "drug_heatmap.pdf", width = 20, height = 20, scale = "none")
```


```{r}
df_cluster_drugs <- df %>% rownames_to_column("product_name") %>% 
  gather(cluster, logFC, -product_name) %>%
  left_join(kistem %>% select(product_name, target,  pathway)) %>% 
  distinct()
  
```

```{r}
df_cluster_drugs %>% 
  arrange(logFC) %>%
  filter(abs(logFC) > 1) %>%
  filter(cluster == "m2")
```


I load the results

```{r}
fisher <- readRDS(here("data/drug_contig/fisher.Rds"))
```


```{r}



residual_drug_df <- chisq_drug$residuals %>% as.data.frame() %>%
  rownames_to_column("cluster") %>%
  as_tibble() %>% 
  gather(drug, residual, - cluster) %>% 
  arrange(desc(residual))

df <- residual_drug_df %>% 
  nest(-cluster) %>% 
  mutate(cluster = as.numeric(cluster)) %>%
  arrange(cluster) %>%
  filter(cluster == 2) %>% #13 Wnt, #11 dead, #12 (not cluster) mTOR
  .$data %>% 
  as.data.frame() %>%
  mutate(drug = factor(drug) %>% fct_inorder()) 

gg_chi <- df %>%
  ggplot(aes(drug, residual)) + 
  geom_point() + 
  theme_classic() +
  ggrepel::geom_text_repel(data = df %>% head(5) %>% rbind(df %>% tail(5)), aes(label = drug)) +
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_blank(),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) + 
  geom_hline(yintercept = 0)
  
  
```

```{r}
df_cluster$mclust_complete_classification %>% table()
```


```{r}
gg_clust2 <- df_cluster %>% 
  sample_frac(0.01) %>%
  filter(mclust_complete_classification == "2") %>%
  ggplot(aes(v1, v2, color = mclust_complete_classification)) + 
  geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") +
  geom_point_rast(alpha = 1, size = .35, shape=16) + 
  
  #facet_wrap(~ mclust_complete_classification) +
  #geom_path() +
  #geom_density2d(color = "black") + 
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) 
  
```


```{r}
clust$data %>%
  as.data.frame() %>%
  cbind(clust$classification) %>% 
  cbind(clust$uncertainty) %>% 
  janitor::clean_names() %>% 
  mutate(clust_uncertainty = (1-clust_uncertainty)*0.1,
         clust_classification = factor(clust_classification, levels = milestone_order)) %>%
  filter(clust_classification == 6) %>%
  ggplot(aes(v1, v2, color = clust_classification)) + 
  geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") +
  geom_point_rast(size = 0.35, alpha = 0.1) + 
  theme_classic()
```


```{r, eval = FALSE}
scheme <- rwantshue::iwanthue(seed = 42, force_init = TRUE) # recreate with a seed
# generate a new color palette (vector of hex values) with presets...


gg_cluster <- plot_cells(ods_sub, color_cells_by="cluster",
           alpha = 0.05) + 
  scale_color_manual(values = scheme$hex(23))
```

I load clustering results and plot them. 

```{r}
ods_sub_clust <- readRDS(here("R/ods_sub_clust_100"))
gg_cluster <- plot_cells(ods_sub_clust, color_cells_by="cluster", rasterize = TRUE,
           alpha = 0.5)
```

I create a graph following the following schema. I run this process in a separate script to speed up processing.

```{r, eval = FALSE}
ods_sub = learn_graph(ods_sub, verbose = TRUE, 
                        close_loop = TRUE, 
                        learn_graph_control = list(minimal_branch_len = 100))
```


```{r}
ods_sub_graph <- readRDS(here("R/ods_sub_graph_10"))
gg_path <- plot_cells(ods_sub_graph, rasterize = TRUE)
```


I save my results 

```{r}
saveRDS(ods_sub, here("data/monocle/bulk_pca_path_subsample.Rds"))
```

# Plot panel

I create a final graph

```{r}
# fig5 <- cowplot::plot_grid(gg_size, gg_line, 
#                    #gg_path, #gg_proportion,
#           gg_bortezomib,#gg_borte_prop,
#           cowplot::plot_grid(gg_trambort_traj , gg_taxane_traj,
#                              NULL, NULL,
#                            #  gg_ono, gg_chir, 
#                              align = "h", axis = "bt", labels = "AUTO", ncol = 2),
#           #gg_chi, gg_clust2,
#           #gg_path, #gg_proportion,
#           #gg_shift, gg_ridge, 
#           NULL, NULL,
#           align = "h", axis = "bt", labels = c("A", "B"), ncol = 2)
#   

# fig5 <- plot_grid(
#             plot_grid(gg_size + theme(legend.position = "nothing"), gg_line,
#                        ncol = 2,
#                        labels = c('A', 'B')),
#             gg_bortezomib + theme(legend.position = "nothing"),
#             # plot_grid(gg_trambort_traj,
#             #           gg_taxane_traj,
#             #           
#             #           NULL, NULL,
#             #           ncol = 2,
#             #           labels = c('D', "E", "F", "G",
#             #                      "H", "I", "J", "K")),
#             ncol = 1,
#             align = "h", axis = "bt", labels = c("", "C", ""))

fig5 <- plot_grid(NULL, 
                  NULL,
                  # plot_grid(gg_line, 
                  #           gg_bortezomib,
                  #           ncol = 2),
                  gg_mtor_all,
                  NULL, # NULL,
                  ncol = 2)

fig5 + 
  ggsave(here("panels/figure_5.pdf"), 
         width = 8.3, 
         height = 11.7)
```

```{r}
gg_trambort_traj + theme_cowplot() theme(legend.position = "none") + ggsave(here("panels/figure_5_aux.pdf"), width = 8, height = )
```


I create a supplemental figure

```{r, eval = TRUE}
cowplot::plot_grid(cowplot::plot_grid(gg_size_dist,
                                      NULL,
                                      ncol = 2),
                   gg_line_overview,
                   cowplot::plot_grid(gg_taxane_traj + theme_cowplot() + theme(legend.position = "none"),
                                      NULL,
                                      ncol = 2), 
                   align = "h", axis = "bt", labels = "AUTO", ncol = 1) + 
  ggsave("figure_S5.pdf", width = 6, height = 12)
```

