---
title: "Organoid Unuspervised Exploration"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE)
```

# Setup

```{r}
library(tidyverse)
library(here)
library(feather)
library(monocle3)
library(ggrastr)
library(cowplot)
library(platetools)
#library(princurve)

# local data access
library(rhdf5) #needed for local data access
library(furrr) #needed for local data access

# install.packages('devtools')
# devtools::install_github('VPetukhov/ggrastr')
mco_colors <- c("#52bde0", "#629d85", "#ffb54d" ,"#cccccc")
here()
```

I use this notebook to analyze phenotypic differences in mouse colon organoids.

```{r, eval = TRUE}
pca_filtered <- readRDS(here("vignettes/06_profiling_inspection/hdf5_pca_filtered.Rds"))
```


```{r, eval = FALSE}
# for demo: subsample the original dataset
set.seed(123)

pca_filtered_store <- pca_filtered
pca_filtered <- pca_filtered %>% 
  sample_n(100000)
```


# UMAP representation of single-organoids

I continue importing the dataset into Monoocle3 

```{r, eval = TRUE}
# generating metadata objects
pca_anno_df <- pca_filtered %>% dplyr::select(-(V1:V25)) %>% 
  mutate(uuid = paste(Plate, Well, Field, ObjectID, sep = "_")) %>% 
  mutate(uuid2 = uuid) %>% 
  mutate(Size = as.numeric(Size)) %>%
  mutate(Size_log = log(Size)) %>%
  as.data.frame() %>% 
  column_to_rownames("uuid2")

pca_matrix <- pca_filtered %>% dplyr::select((V1:V25)) %>% as.data.frame() %>% magrittr::set_rownames(pca_anno_df$uuid) %>% magrittr::set_colnames(c(paste0("PC", c(1:25)))) %>% as.matrix()


ods <- new_cell_data_set(pca_matrix %>% t(),
                         cell_metadata = pca_anno_df)
```

I manually inject the PCA compression of the data into the object

```{r, eval = TRUE}
reducedDims(ods)$PCA <- pca_matrix
```

I run the dim reduction methods

```{r, eval = FALSE}
set.seed(123)
ods <- reduce_dimension(ods,
                        reduction_method = "UMAP",
                        umap.n_neighbors = 15L,
                        umap.fast_sgd=TRUE, 
                        cores = parallel::detectCores(),
                        verbose = TRUE)

saveRDS(ods, here("vignettes/07_organoid_unsupervised_exploration/monocle/filtered_pca_cluster.Rds"))
```

I load the UMAP embedding of the organoid dataset. 

```{r}
ods <- readRDS(here("vignettes/07_organoid_unsupervised_exploration/monocle/filtered_pca_cluster.Rds"))
```

I plot a size-distribution. 

```{r, eval = TRUE}
gg_size_dist <- colData(ods) %>% as_tibble() %>% 
  ggplot(aes(Size)) + 
  geom_histogram() + 
  theme_classic()

gg_size_dist + ggsave(here("results/figures/gg_size_dist.pdf"))
```



```{r, eval = FALSE}
# gg_size <- plot_cells(ods, color_cells_by="Size", 
#            alpha = 0.1,
#            rasterize= TRUE) + 
#   scale_color_viridis_c() + 
#   theme(legend.position = "bottom")
```

I use an alterantive, manual, plotting method. In order to plot the object data, I export a subset of organoids. I plot the cells by size.

```{r}
set.seed(123)

df <- reducedDims(ods)$UMAP %>% cbind(colData(ods)) %>% as_tibble() %>%
  sample_frac(.1)

gg_size <- df %>%
  #filter(Size < 1000) %>%
  ggplot(aes(V1, V2, color = Size_log)) + 
  geom_point_rast(alpha = 0.5, size = 0.35) + 
  scale_color_viridis_c() +
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8))

```

Now I compare the distribution of DMSO treated organoids with drug-treated organoids.

```{r}
set.seed(123)

df <- reducedDims(ods)$UMAP %>% cbind(colData(ods)) %>% as_tibble() %>%
  sample_frac(0.1)

gg_size_supp <- df %>%
  mutate(drug = if_else(Drug == "DMSO", "DMSO", "other")) %>%
  ggplot(aes(V1, V2, color = Size_log)) + 
  geom_point_rast(alpha = 0.5, size = 0.35) + 
  scale_color_viridis_c() +
  theme_cowplot() +
  labs(x = "UMAP 1",
       y = "UMAP 2") + 
  theme(legend.position = "bottom") +
  facet_wrap(~ drug)

gg_size_supp + ggsave(here("results/figures/gg_size_all.pdf"))
```

I create a single plot showing organoid lines and their distribution within the embedding. 

```{r}
set.seed(123)

df <- reducedDims(ods)$UMAP %>% cbind(colData(ods)) %>% as_tibble() %>%
  filter(Drug == "DMSO") 

gg_line <- df %>% dplyr::select(-Line) %>%
  ggplot(aes(V1, V2)) + 
  geom_point_rast(alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point_rast(data = df,
    #mutate(Line = factor(Line, levels = c("D021T01"))),
  aes(color = Line),alpha = .7, size = 0.35, shape=16) + 
  facet_grid(Replicate ~ Line) + 
  scale_color_manual(values = c(c("#D80D12", "#461C01", "#9a4c91", "#70BE6F"))) + 	
  #geom_density2d(color = "black") + 
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) + 
  theme_cowplot(font_size = 8) + 
  theme(legend.position = "nothing") + 
  ggsave(here("results/figures/line.pdf"), height = 8, width = 8)

gg_line
```


Next to replicate-effects, I want to further understands possible effects due to plate number, drugs, position within the plate etc. 

```{r}
set.seed(123)

umap_well <- reducedDims(ods)$UMAP %>% cbind(colData(ods)) %>% as_tibble() %>% group_by(Plate, Well, Line, Replicate) %>% 
  summarise_at(vars(contains("V")), funs(mean))

df <- umap_well %>% 
  ungroup() %>%
  mutate(Plate = if_else(Plate == "M001W01P908L07", "M001W01P008L07", Plate))

platetools::raw_grid(data = df$V2, 
                     well = df$Well,
                     ncols = 12,
                     plate_id = df$Plate,
                     plate = 384) + 
  scale_fill_viridis_c()
```


```{r, eval = FALSE}
ods <- cluster_cells(ods, verbose = TRUE, random_seed = 123, k = 100)
# ods <- learn_graph(ods, 
#                    use_partition = FALSE,
#                    verbose = TRUE)
saveRDS(ods, here("vignettes/07_organoid_unsupervised_exploration/monocle/filtered_pca_cluster.Rds"))
```


```{r}
plot_cells(ods, show_trajectory_graph = FALSE, group_label_size = 3)
```

```{r}
set.seed(123)

df <- reducedDims(ods)$UMAP %>% cbind(colData(ods)) %>% as_tibble() %>%
  sample_frac(0.1)

df %>%
  ggplot(aes(V1, V2, color = Size_log)) + 
  geom_point_rast(alpha = 0.5, size = 0.35) + 
  scale_color_viridis_c() +
  theme_cowplot() +
  labs(x = "UMAP 1",
       y = "UMAP 2") + 
  theme(legend.position = "bottom") +
  facet_wrap(~ Replicate)
```

Three cluster regions appear to be heavily influenced by batch effects (3 per batch).

```{r}
cluster_df <- clusters(ods) %>% as.data.frame() %>% rownames_to_column("id") %>% as_tibble() %>%
  magrittr::set_colnames(c("id", "cluster")) %>%
  left_join(colData(ods) %>% as.data.frame() %>% 
              rownames_to_column("id") %>% as_tibble()) %>% 
  mutate(Drug_Line = paste0(Drug, "_", Line),
         Drug_Line = factor(Drug_Line),
         Drug = factor(Drug),
         Plate_Well_Drug = paste0(Plate, "_", Well, "_", Drug),
         Plate_Well_Drug = factor(Plate_Well_Drug)
         )
```

```{r}
chi_line <- chisq.test(cluster_df$cluster, cluster_df$Line)
chi_line$residuals %>% pheatmap::pheatmap()
```

```{r}
chi_rep <- chisq.test(cluster_df$cluster, cluster_df$Replicate)
chi_rep$residuals %>%
  as_tibble() %>%
  janitor::clean_names() %>%
  arrange(n)
```

```{r}
chi_drug <- chisq.test(cluster_df$cluster, cluster_df$Drug)
chi_drug$residuals %>% pheatmap::pheatmap(show_colnames = FALSE)

df <- chi_drug$residuals %>% as_tibble() %>% janitor::clean_names() %>%
  mutate(cluster = as.numeric(cluster_df_cluster),
         drug = cluster_df_drug) %>% 
  dplyr::select(-cluster_df_cluster, -cluster_df_drug) %>%
  filter(drug == "DMSO") %>%
  arrange((n))

df %>% 
  ggplot(aes(n)) + 
  geom_dotplot()
```

```{r}

```


```{r}
chi_drug_line <- chisq.test(cluster_df$cluster, cluster_df$Drug_Line)
chi_drug_line$residuals %>% pheatmap::pheatmap(show_colnames = FALSE)
```

Find clusters where DMSO is depleted. 

```{r}
drug_line_res <- chi_drug_line$residuals %>% as_tibble() %>% janitor::clean_names()

drug_line_res %>%
  separate(cluster_df_drug_line, c("drug", "line"), sep = "_M001") %>%
  mutate(cluster = as.numeric(cluster_df_cluster),
         line = paste0("M001", line)) %>% 
  dplyr::select(-cluster_df_cluster) %>%
  filter(drug == "DMSO") %>% 
  ggplot(aes(cluster, n, color = line)) + 
  geom_boxplot(aes(group = cluster))
```


```{r}
chi_well_drug <- chisq.test(cluster_df$cluster, cluster_df$Plate_Well_Drug)
well_drug_res <- chi_well_drug$residuals %>% as_tibble() %>% janitor::clean_names()

well_drug_res %>%
  group_by(cluster_df_cluster) %>%
  do(arrange(., desc(n)) %>% head(10)) %>%
  separate(cluster_df_plate_well_drug, c("plate", "well", "drug"), sep = "_", extra = "merge") %>%
  mutate(file_name = paste0(plate, "_", substr(well, 1, 1), "_", substr(well, 2, 3), "_1.jpeg")) %>%
  filter(cluster_df_cluster %in% c("1"))
```


# Removing not reproducible organoid phenotypes using an empirical Bayesian approach

I do not trust my embedding, because certain areas of the embedding are populated exclusively by one replicate of one sample. In order to estimate the reproducibility of certain organoid phenotypes, I estimate the proportion of replicates.

I start out setting an arbitrary number of clusters.

```{r}
kmeans <- df_cluster %>% 
  select(v1, v2) %>% 
  kmeans(centers = 100)

kmeans$size %>% hist(main = "Size of clusters")
```

Now I assess the composition of the kmeans clusters. 

```{r}
df_cluster_km <- df_cluster %>% cbind(kmeans = kmeans$cluster)

km_lr <- df_cluster_km %>% 
  group_by(line, replicate, kmeans) %>% 
  summarise(n = n()) %>% 
  ungroup() %>%
  complete(line, replicate, kmeans, fill = list(n  = 0))
```

I define a prior poportion of replicates. 

Next I use the tidy empirical bayes package by Dave robinson to define certaint estimates that a given proportion is a deviation of my null hypothesis, which states a equal distribution between replicates.

```{r}
km_lr_prop_tidy <- km_lr %>% 
  mutate(replicate = paste0("R", replicate)) %>% 
  spread(replicate, n) %>% 
  mutate(total = R1 + R2) %>% 
  ungroup() %>% 
  group_by(line) %>% 
  mutate(prop_r1 = sum(R1)/(sum(R1)+sum(R2)),
         prop_r2 = sum(R2)/(sum(R1)+sum(R2)))

km_lr_prop_tidy %>%
  ggplot(aes(total, R1/total)) + 
  geom_hline(aes(yintercept = (prop_r1), color = line), linetype = "dashed", alpha= 0.5) + 
  geom_hline(aes(yintercept = (prop_r2), color = line), linetype = "dashed", alpha= 0.5) +
  geom_point(aes(color = line)) + 
  scale_x_log10() + 
  theme_cowplot() 
  
```

I visualize the distrubtions and priors. 

```{r}
km_lr_prop_tidy %>% 
  ggplot(aes(R1/total)) + 
  geom_histogram() + 
  facet_grid(~ line) + 
  theme_cowplot() + 
  geom_vline(aes(xintercept = prop_r1))
```

Given the histograms, I proceed without a mixture model.

I start with a simple model. 

```{r}
km_shrunken <- km_lr_prop_tidy %>% add_ebb_estimate(R1, total+1)
```

I plot the shrinked values. 

```{r}
ggplot(km_shrunken, aes(.raw, .fitted, color = log10(total))) +
  
  geom_abline(color = "red") + 
  geom_point() +
  theme_cowplot() + 
  geom_hline(yintercept = 0.5)
```


And their consequence with 20% credible intervals.

```{r}
km_shrunken %>% 
ggplot(aes(total, .fitted)) + 
  geom_hline(aes(yintercept = (prop_r1), color = line), linetype = "dashed", alpha= 0.5) + 
  geom_hline(aes(yintercept = (prop_r2), color = line), linetype = "dashed", alpha= 0.5) +
  geom_point(aes(color = line)) + 
  scale_x_log10() +
  geom_hline(aes(yintercept = 0.8), linetype = "dashed", alpha= 1) +
  geom_hline(aes(yintercept = 0.2), linetype = "dashed", alpha= 1) +
  theme_cowplot() 
```


I define a hierarchical model which considers both the number of observations and the line of the organoids. 

```{r, eval = FALSE}
library(ebbr)
library(gamlss)
fit <- gamlss(cbind(R1, total - R1) ~ log(total + 1) + line,
               data = km_lr_prop_tidy,
               family = BB(mu.link = "identity"))

tidy(fit)
```

I identify the clusters in question and highlight them. 

```{r}
volatile_clusters <- km_shrunken %>% filter(.fitted < 0.1 | .fitted > 0.9) %>% group_by(kmeans) %>% summarize(n = n()) %>% .$kmeans
```

I remove these clusters from the dataset and plot the initial line plot again. 

```{r}
df_cluster_km %>% 
  filter(!kmeans %in% volatile_clusters) %>%
  group_by(line, replicate, mclust_complete_classification) %>% 
  summarize(n = n()) %>% 
  ungroup() %>%
  mutate(line = paste0(line, "_", replicate)) %>%
  dplyr::select(-replicate) %>%
  spread(mclust_complete_classification, n) %>% 
  as.data.frame() %>%
  column_to_rownames("line") %>% 
  pheatmap::pheatmap(scale = "row")
```

I also create the classic line plots. 

```{r}
df <- df_cluster_km %>% 
  filter(!kmeans %in% volatile_clusters) %>%
  filter(drug == "DMSO") 

gg_line_trim <- df %>% dplyr::select(-line) %>%
  ggplot(aes(v1, v2)) + 
  geom_point_rast(alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point_rast(data = df,
    #mutate(Line = factor(Line, levels = c("D021T01"))),
  aes(color = line),alpha = .2, size = 0.35, shape=16) + 
  facet_grid(replicate ~ line) + 
  scale_color_manual(values = c(c("#D80D12", "#461C01", "#9a4c91", "#70BE6F"))) + 	
  #geom_density2d(color = "black") + 
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) + 
  theme_cowplot(font_size = 8) + 
  theme(legend.position = "nothing")
```

I create a function to iterate over different numbers of k and return the ids of filtered organoids. 

```{r}
bayes_proportion_test <- function(df, k){
  kmeans <- df %>% 
  dplyr::select(v1, v2) %>% 
  kmeans(centers = k)
  
  df_cluster_km <- df %>% cbind(kmeans = kmeans$cluster)

  km_lr <- df_cluster_km %>% 
  group_by(line, replicate, kmeans) %>% 
  summarise(n = n()) %>% 
  ungroup() %>%
  complete(line, replicate, kmeans, fill = list(n  = 0))
  
  km_lr_prop_tidy <- km_lr %>% 
  mutate(replicate = paste0("R", replicate)) %>% 
  spread(replicate, n) %>% 
  mutate(total = R1 + R2) %>% 
  ungroup() %>% 
  group_by(line) %>% 
  mutate(prop_r1 = sum(R1)/(sum(R1)+sum(R2)),
         prop_r2 = sum(R2)/(sum(R1)+sum(R2)))
  
  km_shrunken <- km_lr_prop_tidy %>% add_ebb_estimate(R1, total+1)
  
  return(km_shrunken)
}
```

Let's run this for up to k=10000

```{r}
set.seed(48574)

bpk_df <- tibble(k = seq(10,500,10)) %>% 
  mutate(km = map(k, ~ bayes_proportion_test(df_cluster, .x)))

saveRDS(bpk_df, here("data/kmeans/bpk_1_200.Rds"))
```

I load the data

```{r}
df_cluster <-reducedDims(ods)$UMAP %>% cbind(colData(ods))

bpk_df <- readRDS(here("data/kmeans/bpk_1_200.Rds"))
```


I plot the number of organoids removed for each k and a threshold of 90% disbalance. 

```{r}
bpk_df %>% 
  unnest(km) %>% 
  filter(.fitted < 0.1 | .fitted > 0.9) %>% 
  group_by(line, k) %>% 
  summarize(total = sum(total)) %>%
  ggplot(aes(k, total, color = line)) + 
  geom_point()  + 
  geom_vline(xintercept = 200) +
#  scale_y_log10() +
  theme_cowplot() + 
  geom_smooth(se = FALSE, linetype = "dashed") 
```

I set k=200, repeat the procedure and remove datapoints that belong to irreproducible clusters.

```{r}
df <- df_cluster
k <- 200

set.seed(48574)

kmeans <- df %>% 
dplyr::select(v1, v2) %>% 
kmeans(centers = k)

df_cluster_km <- df %>% cbind(kmeans = kmeans$cluster)

km_lr <- df_cluster_km %>% 
group_by(line, replicate, kmeans) %>% 
summarise(n = n()) %>% 
ungroup() %>%
complete(line, replicate, kmeans, fill = list(n  = 0))

km_lr_prop_tidy <- km_lr %>% 
mutate(replicate = paste0("R", replicate)) %>% 
spread(replicate, n) %>% 
mutate(total = R1 + R2) %>% 
ungroup() %>% 
group_by(line) %>% 
mutate(prop_r1 = sum(R1)/(sum(R1)+sum(R2)),
       prop_r2 = sum(R2)/(sum(R1)+sum(R2)))

km_shrunken <- km_lr_prop_tidy %>% add_ebb_estimate(R1, total+1)

```

I identify clusters that are not reproducible and subsample overrepresented organoids. 

```{r}
set.seed(1231)

index <- km_shrunken %>% filter(.fitted < 0.1 | .fitted > 0.9) %>% 
  dplyr::select(line:R2) %>% 
  gather(replicate, n, -line, -kmeans) %>% 
  mutate(replicate = substr(replicate, 2,2)) %>% 
  group_by(line, kmeans) %>% 
  do(arrange(., n) %>% head(1)) %>% 
  # now I know the cluster that is underrepresented and their counts from the underrepresented replicate. I now turn to the corresponding replicate to downsample from.
  mutate(replicate = if_else(replicate == "1", "2", "1")) 

df_cluster_strat <- index %>% 
  right_join(df_cluster_km, .) %>% 
  group_by(line, kmeans) %>%
  sample_n(n) %>% 
  ungroup() %>%
  dplyr::select(-n) %>%
  rbind(df_cluster_km %>% anti_join(index %>% dplyr::select(-n)) %>% as_tibble()) %>% as_tibble()

```

```{r}
df <- df_cluster_strat %>% 
  filter(drug == "DMSO") 

gg_line_strat <- df %>% dplyr::select(-line) %>%
  ggplot(aes(v1, v2)) + 
  geom_point_rast(alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point_rast(data = df,
    #mutate(Line = factor(Line, levels = c("D021T01"))),
  aes(color = line),alpha = .2, size = 0.35, shape=16) + 
  facet_grid(replicate ~ line) + 
  scale_color_manual(values = c(c("#D80D12", "#461C01", "#9a4c91", "#70BE6F"))) + 	
  #geom_density2d(color = "black") + 
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) + 
  theme_cowplot(font_size = 8) + 
  theme(legend.position = "nothing")

gg_line_strat
```


Once I have subsampled labile clusters I continue by reembedding the whole dataset. I do so by creating a new object and re-running the whole pipeline. For subsampling, I set n to be number of organoids in the cluster from the underrepresented replicate. 

First I define a modified version of the pca raw object that I can use for object creation. 

```{r}
pca_filtered_strat <- df_cluster_strat %>% 
  dplyr::select(Plate = plate, 
                Replicate = replicate, 
                OriginalX = original_x, 
                OriginalY = original_y) %>% 
  semi_join(pca_filtered, .)
```


```{r, eval = TRUE}
# generating metadata objects
pca_anno_df <- pca_filtered_strat %>% dplyr::select(-(V1:V25)) %>% 
  mutate(uuid = paste(Plate, Well, Field, ObjectID, sep = "_")) %>% 
  mutate(uuid2 = uuid) %>% 
  mutate(Size = as.numeric(Size)) %>%
  mutate(Size_log = log(Size)) %>%
  as.data.frame() %>% 
  column_to_rownames("uuid2")

pca_matrix <- pca_filtered_strat %>% dplyr::select((V1:V25)) %>% as.data.frame() %>% magrittr::set_rownames(pca_anno_df$uuid) %>% magrittr::set_colnames(c(paste0("PC", c(1:25)))) %>% as.matrix()


ods <- new_cell_data_set(pca_matrix %>% t(),
                         cell_metadata = pca_anno_df)

# Again, I manually inject the PCA compression of the data into the object
reducedDims(ods)$PCA <- pca_matrix
```

I run the dim reduction methods

```{r, eval = FALSE}
set.seed(123)
ods <- reduce_dimension(ods,
                        reduction_method = "UMAP",
                        umap.n_neighbors = 15L,
                        umap.fast_sgd=TRUE, 
                        cores = parallel::detectCores(),
                        verbose = TRUE)

saveRDS(ods, here("data/monocle/bulk_pca_cluster_strat.Rds"))
```

I load the data after pre-processing 

```{r}
ods <- readRDS(here("phenotypespectrum/vignettes/organoid_pseudotime/monocle/bulk_pca_cluster_strat.Rds"))
```


I plot a size-distribution. 

```{r, eval = FALSE}
gg_size_dist <- colData(ods) %>% as_tibble() %>% 
  ggplot(aes(Size)) + 
  geom_histogram() + 
  theme_classic() + 
  scale_x_log10()

gg_size_dist + ggsave(here("figures/gg_size_dist.pdf"))
```


I plot the cells by size

```{r, eval = FALSE}
gg_size <- plot_cells(ods, color_cells_by="Size", 
           alpha = 0.1,
           rasterize= TRUE) + 
  scale_color_viridis_c() + 
  theme(legend.position = "bottom")
```

I use an alterantive, manual, plotting method. 

```{r}
set.seed(123)

df <- reducedDims(ods)$UMAP %>% cbind(colData(ods)) %>% as_tibble() %>%
  sample_frac(.1)

gg_size <- df %>%
  #filter(Size < 1000) %>%
  ggplot(aes(V1, V2, color = Size_log)) + 
  geom_point_rast(alpha = 0.5, size = 0.35) + 
  scale_color_viridis_c() +
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8))

gg_size

```



```{r}
set.seed(123)

df <- reducedDims(ods)$UMAP %>% cbind(colData(ods)) %>% as_tibble() %>%
  sample_frac(0.1)

gg_size_supp <- df %>%
  mutate(drug = if_else(Drug == "DMSO", "DMSO", "other")) %>%
  ggplot(aes(V1, V2, color = Size_log)) + 
  geom_point_rast(alpha = 0.5, size = 0.35) + 
  scale_color_viridis_c() +
  theme_cowplot() +
  labs(x = "UMAP 1",
       y = "UMAP 2") + 
  theme(legend.position = "bottom") +
  facet_wrap(~ drug)

gg_size_supp + ggsave(here("figures/gg_size_all.pdf"))
```

I create a single plot showing the two extreme organoid lines and their distribution within the embedding. 

```{r}
set.seed(123)

df <- reducedDims(ods)$UMAP %>% cbind(colData(ods)) %>% as_tibble() %>%
  filter(Drug == "DMSO") %>% 
  sample_frac(0.2)

gg_line <- df %>% dplyr::select(-Line) %>%
  ggplot(aes(V1, V2)) + 
  geom_point_rast(alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point_rast(data = df,
    #mutate(Line = factor(Line, levels = c("D021T01"))),
  aes(color = Line),alpha = .2, size = 0.35, shape=16) + 
  facet_wrap( ~ Line) + 
  scale_color_manual(values = c(c("#D80D12", "#461C01", "#9a4c91", "#70BE6F"))) + 	
  #geom_density2d(color = "black") + 
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) + 
  theme_cowplot(font_size = 8) + 
  theme(legend.position = "nothing")

gg_line
```



```{r}
gg_line_overview <- plot_cells(ods, color_cells_by="morphological_class",
           alpha = 0.05, label_cell_groups = FALSE,
           rasterize = TRUE) + 
  facet_wrap(~ Line) + 
  theme(legend.position = "nothing") + 
  scale_color_manual(values = rev(c("#A2549B", "#D80D12", "#6E1614","#69B563"))) 
```

I directly write this supplementary plot to disc. 

```{r, eval = FALSE}
gg_line_overview + 
  ggsave(here("line.pdf"), height = 8, width = 8)
```

## Identifying a generalizable model for organoid death. 

```{r}
mco_umap_tidy <- reducedDims(ods)$UMAP %>% cbind(colData(ods)) %>% as_tibble()
mco_pca_tidy <- counts(ods) %>% as.matrix() %>% t() %>% as.data.frame() %>% cbind(colData(ods)) %>% as.data.frame()%>% rownames_to_column("id") %>% as_tibble()
```

```{r}
cluster <- mco_pca_tidy %>%
 select(contains("PC")) %>% 
  kmeans(centers = 3)

mco_umap_tidy %>% 
  cbind(cluster = cluster$cluster %>% as.character()) %>%
  ggplot(aes(V1, V2, color = cluster)) + 
  geom_point() + 
  theme_cowplot()
```


## Learning labels from single organoid embeddings 



Now I go back to the PCA data and train a set of classifiers to distinguish between their genotype. 
First, I validate that the Bortezomib label contains mostly dead organoids.

```{r}
set.seed(123)

df <- reducedDims(ods)$UMAP %>% cbind(colData(ods)) %>% as_tibble() %>%
  filter(Drug == "Bortezomib (PS-341)")

gg_dead <- df %>% dplyr::select(-Line) %>%
  ggplot(aes(V1, V2)) + 
  geom_point_rast(alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point_rast(data = df,
    #mutate(Line = factor(Line, levels = c("D021T01"))),
  aes(color = Line),alpha = .8, size = 0.35, shape=16) + 
  facet_wrap( ~ Line) + 
  scale_color_manual(values = c(c("#D80D12", "#461C01", "#9a4c91", "#70BE6F"))) + 	
  #geom_density2d(color = "black") + 
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) + 
  theme_cowplot(font_size = 8) + 
  theme(legend.position = "nothing")

gg_dead
```

TODO: define better positive control
TODO: expand dataset with neigherst neighbour search

To validate I look at the treatments with the smallest average diameter per line. 

```{r}
drug_size <- mco_umap_tidy %>% 
  group_by(Drug) %>% 
  mutate(avg_size = mean(Size),
         n = n()) %>% 
  ungroup() %>%
  arrange(avg_size) %>% mutate(Drug = factor(Drug) %>% fct_inorder())

lethal <- drug_size %>% 
  dplyr::select(Drug, avg_size, n) %>% distinct() %>% head(10) %>% .$Drug
  # distinct() %>% 
  # ggplot(aes(Drug, avg_size)) + 
  # geom_point()
```


```{r}
drug_size %>% 
  group_by(Drug, Line) %>% 
  summarise(size = mean(Size_log),
            n = n()) %>%
  ggplot(aes(size, fill = Line)) + 
  geom_histogram() + 
  theme_cowplot()
```

I define a label that is associated with being dead or alive. I assume that Bortezomib is completely wiping out all organoids. I only have 1200 examples of Bortzomib treatment, so there is definetly room for improvement when it comes to the size of my dataset. 

I load the latest version of the model

```{r}
mco_scored_ldc <- read_csv(here("data/train/mco_pca_tidy_predcited_xgb_ldc.csv")) %>% 
  clean_names() %>%
  magrittr::set_colnames(c("no", 'positive_pred', 'negative_pred')) %>% cbind(., mco_pca_tidy %>%
  dplyr::select(-(PC1:PC25)))
```

```{r}
mco_scored_ldc %>% 
  ggplot(aes(positive_pred, fill = Line)) + 
  geom_histogram() +
  theme_cowplot() + 
  facet_wrap(~ Line) + 
  scale_y_log10()
```

```{r}
mco_scored_ldc %>% 
  ggplot(aes(Size, fill = Line)) + 
  geom_histogram() +
  theme_cowplot() + 
  facet_wrap(~ Line)
```

I wonder about the performance of the genotype classifier.

```{r}
mco_scored_gene <- read_csv(here("data/train/mco_pca_tidy_predcited_xgb_gene.csv")) %>% 
  magrittr::set_colnames(c("no", "0", "1", "2", "3")) %>% cbind(., mco_pca_tidy %>%
  dplyr::select(contains("Line"))) %>% 
  mutate(Line = case_when(Line == "M001W01" ~ 0,
                          Line == "M001K02" ~ 1,
                          Line == "M001A03" ~ 2,
                          Line == "M001B04" ~ 3)) %>% 
  dplyr::select(-no)

auroc <- mco_scored_gene %>% 
  mutate(Line = factor(Line)) %>%
  yardstick::roc_curve(Line, `0`:`3`)

auroc %>% autoplot()
```




Now I average the score for each treatment for each score.

```{r}
df <- mco_scored_gene %>%
  select(-Line) %>%
  cbind(mco_scored_ldc %>% select(positive_pred)) %>%
  cbind(mco_pca_tidy %>% select(-contains("PC"))) %>% 
  cbind(mco_umap_tidy %>% select(V1)) %>%
  group_by(Drug, Line) %>% 
  summarise(x0 = mean(`0`),
            x1 = mean(`1`),
            x2 = mean(`2`),
            x3 = mean(`3`),
            size = mean(Size),
            n = n(),
            positive = mean(positive_pred),
            umap1 = mean(V1))

loi = "M001W01"

df %>% 
  filter(Line == loi) %>%
  ggplot(aes(x0, umap1, color = positive, size = n)) + 
  geom_hline(yintercept = -2, linetype = "dashed") +
  geom_point() + 
  theme_cowplot() + 
  facet_wrap(~ Line) + 
  scale_color_viridis_c() + 
  labs(x = "Score",
       y = "UMAP 1") + 
  ggrepel::geom_text_repel(data = df %>% filter(Line == loi & umap1 > -2 & x0 < 0.41),
                           aes(label = Drug),
                           color = "black",
                           size = 4)
  

```

I plot the distribution of intesitites

```{r}
mean(df$umap1) - 2*sd(df$umap1)
```

```{r}
df %>% 
  filter(Line == "M001A03") %>%
  arrange(desc(x0))
```


### I run load the results of the APC classifier


```{r}
mco_scored_apc <- read_csv(here("data/train/mco_pca_tidy_predcited_xgb_apc.csv")) %>% .[,c(1:3)] %>%
  magrittr::set_colnames(c("no", "0", "1")) %>% cbind(., mco_pca_tidy %>%
  dplyr::select(contains("Line"))) %>% 
  mutate(apc = case_when(Line == "M001W01" ~ 0,
                          Line == "M001K02" ~ 0,
                          Line == "M001A03" ~ 1,
                          Line == "M001B04" ~ 1)) %>% 
  dplyr::select(-no)

auroc <- mco_scored_apc %>% dplyr::select(-Line) %>%
  mutate(apc = factor(apc)) %>%
  yardstick::roc_curve(apc, `0`,`1`)

auroc %>% autoplot()
```


Now I average the score for each treatment for each score.

```{r}
df <- mco_scored_apc %>%
  select(-Line) %>%
  cbind(mco_scored_ldc %>% select(positive_pred)) %>%
  cbind(mco_pca_tidy %>% select(-contains("PC"))) %>% 
  cbind(mco_umap_tidy %>% select(V1)) %>%
  group_by(Drug, Line) %>% 
  summarise(x0 = mean(`0`),
            x1 = mean(`1`),
            #x2 = mean(`2`),
            #x3 = mean(`3`),
            size = mean(Size),
            n = n(),
            positive = mean(positive_pred),
            umap1 = mean(V1))

loi = c("M001W01", "M001K02")

df %>% 
  filter(Line %in% loi) %>%
  ggplot(aes(x0, umap1, color = positive, size = n)) + 
  geom_hline(yintercept = -2, linetype = "dashed") +
  geom_point() + 
  theme_cowplot() + 
  facet_wrap(~ Line) + 
  scale_color_viridis_c() + 
  labs(x = "Score",
       y = "UMAP 1") + 
  ggrepel::geom_text_repel(data = df %>% filter(Line %in% loi & umap1 > -1 & x0 < 0.75),
                           aes(label = Drug),
                           color = "black",
                           size = 4)
  

```

```{r}
df <- mco_scored_apc %>%
  select(-Line) %>%
  cbind(mco_scored_ldc %>% select(positive_pred)) %>%
  cbind(mco_pca_tidy %>% select(-contains("PC"))) %>% 
  cbind(mco_umap_tidy %>% select(V1)) %>%
  mutate(Line = if_else(Line == "M001W01" | Line == "M001K02", "WT & KRAS", "APC & APC-KRAS")) %>%
  group_by(Drug, Line) %>% 
  summarise(x0 = mean(`0`),
            x1 = mean(`1`),
            #x2 = mean(`2`),
            #x3 = mean(`3`),
            size = mean(Size),
            n = n(),
            positive = mean(positive_pred),
            umap1 = mean(V1))

loi = c("WT & KRAS")

df %>% 
  filter(Line %in% loi) %>%
  ggplot(aes(x0, umap1, color = positive, size = n)) + 
  geom_hline(yintercept = -2, linetype = "dashed") +
  geom_point() + 
  theme_cowplot() + 
  facet_wrap(~ Line) + 
  scale_color_viridis_c() + 
  labs(x = "Score",
       y = "UMAP 1") + 
  ggrepel::geom_text_repel(data = df %>% filter(Line %in% loi & umap1 > -4 & x0 < 0.8),
                           aes(label = Drug),
                           color = "black",
                           size = 4)
```

CHIR-98014 - GSK3a/b inh.
CP-673451 - PDGFRa/b inh.
Crenolanib - PDGFRa/b inh.
AZD1080 - GSK3a/b inh.

```{r}
df <- mco_scored_apc %>%
  select(-Line) %>%
  cbind(mco_scored_ldc %>% select(positive_pred)) %>%
  cbind(mco_pca_tidy %>% select(-contains("PC"))) %>% 
  cbind(mco_umap_tidy %>% select(V1)) %>%
  group_by(Drug, Line) %>% 
  summarise(x0 = mean(`0`),
            x1 = mean(`1`),
            #x2 = mean(`2`),
            #x3 = mean(`3`),
            size = mean(Size),
            n = n(),
            positive = mean(positive_pred),
            umap1 = mean(V1))


loi = c("M001A03", "M001B04")

df %>% 
  filter(Line %in% loi) %>%
  ggplot(aes(x0, umap1, color = positive, size = n)) + 
  geom_hline(yintercept = -2, linetype = "dashed") +
  geom_point() + 
  theme_cowplot() + 
  facet_wrap(~ Line) + 
  scale_color_viridis_c() + 
  labs(x = "Score",
       y = "UMAP 1") + 
  ggrepel::geom_text_repel(data = df %>% filter(Line %in% loi & umap1 > -1.5 & x0 > 0.15),
                           aes(label = Drug),
                           color = "black",
                           size = 4)
```

```{r}
df <- mco_scored_apc %>%
  select(-Line) %>%
  cbind(mco_scored_ldc %>% select(positive_pred)) %>%
  cbind(mco_pca_tidy %>% select(-contains("PC"))) %>% 
  cbind(mco_umap_tidy %>% select(V1)) %>%
  mutate(Line = if_else(Line == "M001W01" | Line == "M001K02", "WT & KRAS", "APC & APC-KRAS")) %>%
  group_by(Drug, Line) %>% 
  summarise(x0 = mean(`0`),
            x1 = mean(`1`),
            #x2 = mean(`2`),
            #x3 = mean(`3`),
            size = mean(Size),
            n = n(),
            positive = mean(positive_pred),
            umap1 = mean(V1))

loi = c("APC & APC-KRAS")

cutoff <- 3*sd(df %>% filter(Line %in% loi) %>% .$x0) + mean(df %>% filter(Line %in% loi) %>% .$x0)

df %>% 
  filter(Line %in% loi) %>%
  ggplot(aes(x0, umap1, color = positive, size = n)) + 
  geom_hline(yintercept = -2, linetype = "dashed") +
  geom_point() + 
  theme_cowplot() + 
  facet_wrap(~ Line) + 
  scale_color_viridis_c() + 
  labs(x = "Score",
       y = "UMAP 1") + 
  ggrepel::geom_text_repel(data = df %>% filter(Line %in% loi & umap1 > -2 & x0 > cutoff),
                           aes(label = Drug),
                           color = "black",
                           size = 4)
```

Buparlisib - Pan-PI3K inhibitor
Hesparidin - Citrus Fruit extract 
KW-2449 - multi-target, FLT3 inhibtor
AZD7762 - CHK inhibitor
Ellagic acid - Natural product

```{r, eval = FALSE}
df %>% 
  ggplot(aes(x0, umap1, color = positive, size = n)) + 
  geom_hline(yintercept = -2, linetype = "dashed") +
  geom_point() + 
  theme_cowplot() + 
  #facet_wrap(~ Line) + 
  scale_color_viridis_c() + 
  labs(x = "Score",
       y = "UMAP 1") + 
  ggrepel::geom_text_repel(data = df %>% filter(umap1 > -2 & (x0 > 0.1 & x0 < 0.8)),
                           aes(label = Drug),
                           color = "black",
                           size = 4)
```

I test if logistic regression can also be used for this. 

```{r}
model = lm(apc ~ ., data = mco_pca_tidy_ml_split %>%
  training() %>% 
    dplyr::select(contains("PC"), apc))

prediction <- predict(model, newdata = mco_pca_tidy_ml)


df_logit <- mco_scored_apc %>%
  select(-Line, -`0`, -`1`) %>%
  cbind(x0 = prediction) %>%
  cbind(mco_scored_ldc %>% select(positive_pred)) %>%
  cbind(mco_pca_tidy %>% select(-contains("PC"))) %>% 
  cbind(mco_umap_tidy %>% select(V1)) %>%
  mutate(Line = if_else(Line == "M001W01" | Line == "M001K02", "WT & KRAS", "APC & APC-KRAS")) %>%
  group_by(Drug, Line) %>% 
  summarise(x0 = mean(x0),
            #x1 = mean(`1`),
            #x2 = mean(`2`),
            #x3 = mean(`3`),
            size = mean(Size),
            n = n(),
            positive = mean(positive_pred),
            umap1 = mean(V1))
```



```{r}
df <- reducedDims(ods)$UMAP %>% cbind(mco_scored_ldc) %>% as_tibble()

gg_ldc <- df %>% 
  janitor::clean_names() %>%
  ggplot(aes(x1, x2, color = positive_pred)) + 
  geom_point_rast(alpha = .2, size = 0.35, shape=16) + 
  facet_wrap( ~ line) + 
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme_cowplot(font_size = 8) + 
  scale_color_viridis_c()

gg_ldc
```




## Regions of the embedding



## Information gain for distributions

```{r}
set.seed(123)

df <- reducedDims(ods)$UMAP %>% cbind(colData(ods)) %>% as_tibble() %>%
  filter(Drug == "DMSO")

kl_frac <- function(frac, df){
  X <- df %>% filter(Line == "M001B04", Replicate == "1") %>% sample_frac(frac) %>% dplyr::select(V1, V2) %>% as.matrix()
  Y <- df %>% filter(Line == "M001B04", Replicate == "2") %>% sample_frac(frac) %>% dplyr::select(V1, V2)%>% as.matrix()
kl = mean(KL.divergence(X,Y, k = 10))
return(kl)
}

kl_frac_inv <- function(frac, df){
  Y <- df %>% filter(Line == "M001B04", Replicate == "1") %>% sample_frac(frac) %>% dplyr::select(V1, V2) %>% as.matrix()
  X <- df %>% filter(Line == "M001B04", Replicate == "2") %>% sample_frac(frac) %>% dplyr::select(V1, V2)%>% as.matrix()
kl = mean(KL.divergence(X,Y, k = 10))
return(kl)
}


kl_frac_benchmark <- tibble(frac = seq(0.05, 1, 0.01)) %>% 
  mutate(KL = map(frac, ~ kl_frac(.x, df))) %>%
  mutate(KL_inv = map(frac, ~ kl_frac_inv(.x, df)))

kl_frac_benchmark %>% unnest() %>%
  gather(div, value, -frac) %>%
  ggplot(aes(frac, value, color = div)) + 
  geom_point() + 
  theme_cowplot() + 
  scale_color_brewer(type = "qual")+ 
  ggtitle("Distributions of same size")
```

Comparing samples of different sizes

```{r}
set.seed(123)

df <- reducedDims(ods)$UMAP %>% cbind(colData(ods)) %>% as_tibble() %>%
  filter(Drug == "DMSO")

kl_frac <- function(frac, df){
  X <- df %>% filter(Line == "M001B04", Replicate == "1") %>% sample_frac(frac) %>% dplyr::select(V1, V2) %>% as.matrix()
  Y <- df %>% filter(Line == "M001B04", Replicate == "2") %>% sample_frac(frac) %>% sample_frac(0.1) %>% dplyr::select(V1, V2)%>% as.matrix()
kl = mean(KL.divergence(X,Y, k = 10))
return(kl)
}

kl_frac_inv <- function(frac, df){
  Y <- df %>% filter(Line == "M001B04", Replicate == "1") %>% sample_frac(frac) %>% dplyr::select(V1, V2) %>% as.matrix()
  X <- df %>% filter(Line == "M001B04", Replicate == "2") %>% sample_frac(frac) %>% sample_frac(0.1) %>% dplyr::select(V1, V2)%>% as.matrix()
kl = mean(KL.divergence(X,Y, k = 10))
return(kl)
}


kl_frac_benchmark <- tibble(frac = seq(0.05, 1, 0.01)) %>% 
  mutate(KL = map(frac, ~ kl_frac(.x, df))) %>%
  mutate(KL_inv = map(frac, ~ kl_frac_inv(.x, df)))

kl_frac_benchmark %>% unnest() %>%
  gather(div, value, -frac) %>%
  ggplot(aes(frac, value, color = div)) + 
  geom_point() + 
  theme_cowplot() + 
  scale_color_brewer(type = "qual") + 
  ggtitle("Minor Distribution of 10%")
```


What is the average number of observations per treatment in the dataset? 

First I simplify my dataset.

```{r}
ods_df <- reducedDims(ods)$UMAP %>% cbind(colData(ods)) %>% as_tibble()
```

Now I move on. 

```{r}
median <- ods_df %>% group_by(Line, Drug) %>% summarise(n = n()) %>% 
  filter(Drug != "Drug") %>% .$n %>% median()

ods_df %>% group_by(Line, Drug) %>% summarise(n = n()) %>% 
  filter(Drug != "Drug") %>%
  ggplot(aes(n)) + 
  geom_histogram() + 
  scale_x_log10() + 
  #scale_y_log10() + 
  theme_cowplot() + 
  geom_vline(xintercept = median)
```


Now I calculate the KLD for every APC treatment and the joined distribution of WT organoids. 

```{r}
KLD_revert_wt <- ods_df %>% 
  nest(-Drug, -Line) %>%
  #head(10) %>%
  mutate(kl = map(data, ~ mean(KL.divergence(.x %>% dplyr::select(V1, V2) %>% as.matrix(),
                                             ods_df %>% filter(Line == "M001W01" & Drug == "DMSO") %>% dplyr::select(V1, V2) %>% as.matrix(),
                                             k = 10)))) %>% 
  unnest(kl)

KLD_revert_apc <- ods_df %>% 
  nest(-Drug, -Line) %>%
  #head(10) %>%
  mutate(kl = map(data, ~ mean(KL.divergence(.x %>% dplyr::select(V1, V2) %>% as.matrix(),
                                             ods_df %>% filter(Line == "M001A03" & Drug == "DMSO") %>% dplyr::select(V1, V2) %>% as.matrix(),
                                             k = 10)))) %>% 
  unnest(kl)

KLD_revert_kras <- ods_df %>% 
  nest(-Drug, -Line) %>%
  #head(10) %>%
  mutate(kl = map(data, ~ mean(KL.divergence(.x %>% dplyr::select(V1, V2) %>% as.matrix(),
                                             ods_df %>% filter(Line == "M001K02" & Drug == "DMSO") %>% dplyr::select(V1, V2) %>% as.matrix(),
                                             k = 10)))) %>% 
  unnest(kl)

KLD_revert_ak <- ods_df %>% 
  nest(-Drug, -Line) %>%
  #head(10) %>%
  mutate(kl = map(data, ~ mean(KL.divergence(.x %>% dplyr::select(V1, V2) %>% as.matrix(),
                                             ods_df %>% filter(Line == "M001B04" & Drug == "DMSO") %>% dplyr::select(V1, V2) %>% as.matrix(),
                                             k = 10)))) %>% 
  unnest(kl)

```

Now I plot the distribution of KLDs. 


```{r}
KLD_revert_wt %>% 
  ggplot(aes(kl, fill = Line)) + 
  geom_histogram(alpha = 0.7, position = "identity") + 
  scale_x_log10() + 
  theme_cowplot() + 
  facet_grid(~ Line)
```

I load the drug anno table 

```{r}
ccp <- read_delim(here("data/layouts/Clin_Cancer_Panel_V170511.csv"), 
    ";", escape_double = FALSE, trim_ws = TRUE) %>% .[,-1] %>% 
  janitor::clean_names()

kistem <- read_delim(here("data/layouts/KiStem_edt_V170125.csv"), 
    ";", escape_double = FALSE, trim_ws = TRUE) %>% .[,-1] %>% 
  janitor::clean_names()
```



```{r}
KLD_revert_wt %>% 
  filter(Line == "M001B04") %>% arrange(kl) %>% 
  left_join(kistem %>% dplyr::select(Drug = product_name, target, pathway)) %>% distinct()
```

```{r}
KLD_revert_wt %>% 
  filter(Line == "M001A03") %>% arrange(kl)%>% 
  left_join(kistem %>% dplyr::select(Drug = product_name, target, pathway)) %>% distinct()
```



```{r}
df <- reducedDims(ods)$UMAP %>% cbind(colData(ods)) %>% as_tibble() %>%
  filter(Drug %in% c("DMSO", "CHIR-98014")) %>%
  group_by(Drug) %>%
  #sample_frac(0.1) %>% 
  ungroup()

gg_gsk <- df %>% dplyr::select(-Line, -Drug) %>%
  ggplot(aes(V1, V2)) + 
  geom_point_rast(alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point_rast(data = df,
  aes(color = Line),alpha = .2, size = 0.7, shape=16) + 
  facet_grid(Drug ~ Line) + 
  scale_color_manual(values = c(c("#D80D12", "#461C01", "#9a4c91", "#70BE6F"))) + 	
  labs(x = "UMAP 1",
       y = "UMAP 2")+
  theme_cowplot(font_size = 8) + 
  theme(legend.position = "nothing")
```


```{r}
KLD_revert_wt %>% 
  filter(Line != "M001W01") %>%
  filter(Line != "M001K02") %>%
  filter(Drug != "DMSO") %>%
  mutate(n = map(data, ~ nrow(.x))) %>% 
  unnest(n) %>% 
  ggplot(aes(n, kl, color = Line)) + 
  geom_point() + 
  theme_cowplot() + 
  #scale_x_log10() + 
  scale_y_log10() + 
  facet_wrap(~ Line, ncol =2)
```


```{r}
KLD_revert_wt <- ods_df %>% 
  nest(-Drug, -Line) %>%
  #head(10) %>%
  mutate(kl = map(data, ~ mean(KL.divergence(.x %>% dplyr::select(V1, V2) %>% as.matrix(),
                                             ods_df %>% filter(Line == "M001W01" & Drug == "DMSO") %>% dplyr::select(V1, V2) %>% as.matrix(),
                                             k = 10)))) %>% 
  unnest(kl)
```


```{r}
KLD_revert_wt %>% 
    #filter(Line == "M001B04") %>% arrange(kl) %>% 
    left_join(kistem %>% dplyr::select(Drug = product_name, target, pathway)) %>% distinct() %>% 
  mutate(target = if_else(target == "GSK-3", TRUE, FALSE)) %>%
  mutate(target = if_else(is.na(target), FALSE, target)) %>% 
  filter(Line != "M001W01") %>%
  filter(Line != "M001K02") %>%
  filter(Drug != "DMSO") %>%
  mutate(n = map(data, ~ nrow(.x))) %>% 
  unnest(n) %>% 
  ggplot(aes(n, kl, color = target)) + 
  geom_point() + 
  theme_cowplot() + 
  #scale_x_log10() + 
  scale_y_log10() + 
  facet_wrap(~ Line, ncol =2)
```

Permutation test of distribution

```{r}
KLD_revert_wt_null <- KLD_revert_wt %>% 
  mutate(n = map(data, ~ nrow(.x))) %>% 
  unnest(n) %>% 
  #head(1) %>% 
  #dplyr::slice(rep(1:n(), each = 2)) %>%
  dplyr::slice(rep(1:n(), each = 10000)) %>%
  mutate(kl_null = map2(Line, n, ~ mean(KL.divergence(ods_df %>% filter(Line == .x) %>% sample_n(.y) %>% dplyr::select(V1, V2) %>% as.matrix(),
                                                  ods_df %>% filter(Line == "M001K02" & Drug == "DMSO") %>% dplyr::select(V1, V2) %>% as.matrix(),
                                                  k = 10)))) %>% 
  unnest(kl_null)

saveRDS(KLD_revert_wt_null, "KLD_revert_wt_null.Rds")
```


















































```{r}
clust_var <- df_cluster %>% 
  filter(drug == "DMSO") %>%
  group_by(line, mclust_complete_classification) %>% 
  summarize(n = n()) %>%
  group_by(line) %>% 
  mutate(prop = n/sum(n)) %>%
  dplyr::select(-n)  

clust_var_order <- clust_var %>% 
  group_by(mclust_complete_classification) %>% 
  summarise(var = var(prop)) %>% 
  arrange(desc(var)) %>%
  .$mclust_complete_classification %>% as.character()

clust_var %>% 
  group_by(mclust_complete_classification) %>% 
  summarise(var = var(prop)) %>% 
  arrange(desc(var)) %>% 
  mutate(mclust_complete_classification = factor(mclust_complete_classification, levels = clust_var_order)) %>%
  mutate(status = if_else(mclust_complete_classification %in% c("M14", "M13", "M1", "M5"), "dead", "alive")) %>%
  ggplot(aes(mclust_complete_classification, var, color = status)) + 
  geom_point() + 
  theme_cowplot()

# clust_var %>% 
#   mutate(mclust_complete_classification = factor(mclust_complete_classification, levels = clust_var_order)) %>%
#   
#   ggplot(aes(mclust_complete_classification, prop, color = status)) + 
#   geom_point()

# clust_mu_order <- clust_var %>% 
#   group_by(mclust_complete_classification) %>% 
#   summarise(mu = mean(prop)) %>% 
#   arrange(desc(mu)) %>%
#   .$mclust_complete_classification %>% as.character()
# 
# clust_var %>% 
#   mutate(mclust_complete_classification = factor(mclust_complete_classification, levels = clust_mu_order)) %>%
#   ggplot(aes(mclust_complete_classification, prop)) + 
#   geom_point()

line_order <- clust_var %>% 
  spread(mclust_complete_classification, prop) %>%
  janitor::clean_names() %>%
  arrange(desc(m5)) %>% 
  .$line
```

I write a csv with the proportions. 

```{r}
clust_var %>% 
  mutate(mclust_complete_classification = as.character(mclust_complete_classification)) %>%
  spread(mclust_complete_classification, prop) %>%
  janitor::clean_names() %>% 
  write_csv(here("data/cluster_proportions.csv"))
```


I associate proportions with a cluster of interest, M4

```{r}
clust_var %>% 
  filter(mclust_complete_classification == "M4") %>% 
  filter(line != "D021T01") %>% 
  mutate(pi3k_akt = if_else(line %in% c("D007T01", "D004T01",
                                        "D013T01", "D030T01",
                                        "D054T01"), TRUE, FALSE)) %>% 
  ggplot(aes(pi3k_akt, prop)) +
  geom_jitter(width = 0.2)+ 
  theme_cowplot() + 
  ggsignif::geom_signif(comparisons = list(c("TRUE", "FALSE")))
```

It is hard to define a mutation status as false given the limited coverage of our amplicon library.

```{r, eval = FALSE}
gg_proportion <- df_cluster %>% 
  mutate(line = factor(line, levels = line_order)) %>%
  #filter(!(line %in% c("D054T01", "D021T01", "D020T01", "D010T01"))) %>%
  filter(!(line %in% c("D020T02", "D010T01"))) %>%
  filter(drug == "DMSO") %>%
  ggplot(aes(line, fill = mclust_complete_classification)) + 
  geom_bar(position = "fill") + 
  theme_classic() + 
  coord_flip() + 
  #theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) + 
  scale_fill_manual(values = scico::scico(15, palette = 'roma')) +
  #facet_wrap(~ Drug)
  theme(legend.text = element_blank()) + 
  theme(legend.position = "None")

gg_proportion
```

Now I am interested how certain drugs and their concentrations shift the composition of morphological phenotypes

```{r, eval = FALSE}
drug_order <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Bortezomib")) %>%
    filter(line == "D046T01") %>% .$drug %>% unique()
drug_order <- drug_order[rev(c(3,2,5,4,6,1))]


gg_borte_prop <- df_cluster %>% 
  mutate(line = factor(line, levels = line_order)) %>%
  #filter(!(line %in% c("D054T01", "D021T01", "D020T01", "D010T01"))) %>%
  filter(!(line %in% c("D021T01", "D010T01"))) %>%
  filter(drug == "DMSO" | grepl(drug, pattern = "Bortezomib")) %>%
  filter(!grepl(drug, pattern = "Dabrafenib")) %>% 
  filter(line == "D046T01") %>%
  mutate(drug = factor(drug, levels =drug_order)) %>%
  ggplot(aes(drug, fill = mclust_complete_classification)) + 
  geom_bar(position = "fill") + 
  theme_classic() + 
  coord_flip() + 
  #theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) + 
  scale_fill_manual(values = scico::scico(15, palette = 'roma')) +
  #facet_wrap(~ concentration) +
  theme(legend.text = element_blank()) + 
  theme(legend.position = "None")
```

I visualize this transition in the UMAP.

```{r}
set.seed(123)

drug_order <- df_cluster %>%
    #filter(drug == "DMSO" | grepl(drug, pattern = "Bortezomib")) %>%
  filter(grepl(drug, pattern = "Bortezomib")) %>%
    filter(line == "D046T01") %>% .$drug %>% unique()
drug_order <- drug_order[rev(c(3,2,5,4,6,1))]

gg_bortezomib <- df_cluster %>% sample_frac(0.01) %>%
  dplyr::select(-line, -concentration) %>%
  ggplot(aes(v1, v2)) + 
  geom_point_rast(alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point_rast(data = df_cluster %>%
    #filter(drug == "DMSO" | grepl(drug, pattern = "Bortezomib")) %>%
      filter(grepl(drug, pattern = "Bortezomib")) %>%
  filter(line %in% c("D055T01", "D054T01", "D046T01")) %>%
    mutate(concentration = if_else(concentration == "nan", "0", concentration)) %>%
    mutate(drug = factor(drug, levels =drug_order)) %>%
    
    group_by(drug) %>% 
    sample_n(5000, replace = TRUE),
  aes(color = drug),alpha = 1, size = 1.5, shape=16, color = "black") + 
  facet_wrap( ~ concentration, ncol = 1) + 
  #scale_color_brewer(type = "seq", palette = "YlOrRd") + 
  #geom_density2d(color = "black") + 
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) + 
  theme_cowplot(font_size = 8) + 
  theme(legend.position = "bottom")
  


gg_bortezomib
```


I visualize the location of Bortezomib 

```{r}
drug_order <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Bortezomib")) %>%
    filter(line == "D046T01") %>% .$drug %>% unique()
drug_order <- drug_order[rev(c(3,2,5,4,6,1))]


center_df <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Bortezomib")) %>%
  # filter(line == "D046T01") %>% 
    mutate(drug = factor(drug, levels =drug_order)) %>%
    # group_by(drug) %>% 
    # sample_n(200, replace = FALSE) %>% 
  dplyr::select(v1, v2, drug, line) %>% 
  group_by(drug, line) %>% 
  summarize(v1 = mean(v1),
            v2 = mean(v2))

gg_bortezomib_path <- center_df %>%
  ggplot(aes(v1, v2, group = line)) + 
  
  scale_color_brewer(type = "seq", palette = "YlOrRd") + 
  geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point(aes(color = drug),alpha = 1, size = 3, shape=16) + 
  geom_path() +
  facet_wrap(~line ) +
  #geom_density2d(color = "black") + 
  theme_bw() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) 
```


I repeat the same figure with Trametinib

```{r}
df_cluster %>% 
  mutate(line = factor(line, levels = line_order)) %>%
  #filter(!(line %in% c("D054T01", "D021T01", "D020T01", "D010T01"))) %>%
  filter(!(line %in% c("D021T01", "D010T01"))) %>%
  filter(drug == "DMSO" | grepl(drug, pattern = "Trametinib")) %>%
  filter(!grepl(drug, pattern = "Dabrafenib")) %>% 
  filter(!grepl(drug, pattern = "GSK1120212")) %>%
  filter(line == "D046T01") %>%
  ggplot(aes(drug, fill = mclust_complete_classification)) + 
  geom_bar(position = "fill") + 
  theme_classic() + 
  coord_flip() + 
  #theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) + 
  scale_fill_manual(values = scico::scico(15, palette = 'roma')) +
  #facet_wrap(~ concentration) +
  theme(legend.text = element_blank()) + 
  theme(legend.position = "None")
```


```{r}
drug_order <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Trametinib")) %>%
  filter(!grepl(drug, pattern = "Dabrafenib")) %>% 
  filter(!grepl(drug, pattern = "GSK1120212")) %>% 
    filter(line == "D046T01") %>% .$drug %>% unique() %>% sort()


gg_trametinib <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Trametinib")) %>%
      filter(!grepl(drug, pattern = "Dabrafenib")) %>% 
       filter(!grepl(drug, pattern = "GSK1120212")) %>%
  filter(line == "D046T01") %>% 
    mutate(drug = factor(drug, levels =drug_order)) %>%
    group_by(drug) %>% 
    sample_n(200, replace = FALSE) %>%
  #sample_n(1000, replace = FALSE) %>%
  ggplot(aes(v1, v2)) + 
  
  geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_density2d(aes(color = drug)) +
  geom_point_rast(aes(color = drug),alpha = 1, size = 1, shape=16) + 
  facet_wrap(~ drug) + 
  scale_color_brewer(type = "seq", palette = "YlOrRd") + 
  #geom_density2d(color = "black") + 
  theme_bw() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) 

gg_trametinib
gg_trametinib + ggsave("trametinib.pdf", width = 15, height = 10)

```

```{r}
gg_trametinib_all <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Trametinib")) %>%
      filter(!grepl(drug, pattern = "Dabrafenib")) %>% 
      filter(grepl(drug, pattern = "GSK1120212")) %>%
  # #filter(line == "D046T01") %>% 
  #   mutate(drug = factor(drug, levels =drug_order)) %>%
  #   group_by(drug) %>% 
  #   sample_n(200, replace = FALSE) %>%
  #sample_n(1000, replace = FALSE) %>%
  ggplot(aes(v1, v2)) + 
  
  geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  #geom_density2d(aes(color = drug)) +
  geom_point_rast(alpha = 1, size = 0.35, shape=16, color = "black") + 
  #facet_wrap(~ drug) + 
  #scale_color_brewer(type = "seq", palette = "YlOrRd") + 
  #geom_density2d(color = "black") + 
  theme_bw() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) + 
  theme_cowplot() + theme(legend.position = "nothing")

gg_mtor_all <- df_cluster %>%
    #filter(drug == "DMSO" | grepl(drug, pattern = "WYE-132") | drug %in% c("PF-04691502", "PP121")) %>%
  filter(drug == "DMSO" | grepl(drug, pattern = "WYE-132")) %>%
  filter(line %in% c("D030T01", "D027T01", "D013T01")) %>%
  #filter(line %in% c("D055T01", "D054T01", "D046T01")) %>%
  group_by(drug) %>% 
  sample_n(700) %>%
  ggplot(aes(v1, v2)) + 
  
  geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  #geom_density2d(aes(color = drug)) +
  geom_point_rast(aes(color = drug),alpha = 1, size = 1, shape=16, color = "black") + 
  facet_wrap(~ drug) + 
  #scale_color_brewer(type = "seq", palette = "YlOrRd") + 
  #geom_density2d(color = "black") + 
  theme_bw() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) + 
  theme_cowplot(font_size = 8) + 
  theme(legend.position = "nothing")


gg_gski_all <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "CHIR-98014")) %>%
  filter(line %in% c("D055T01", "D054T01", "D046T01")) %>%
  group_by(drug) %>% 
  sample_n(1380) %>%
  ggplot(aes(v1, v2)) + 
  
  geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  #geom_density2d(aes(color = drug)) +
  geom_point_rast(aes(color = drug),alpha = 1, size = 1, shape=16, color = "black") + 
  facet_wrap(~ drug) + 
  #scale_color_brewer(type = "seq", palette = "YlOrRd") + 
  #geom_density2d(color = "black") + 
  theme_bw() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) + 
  theme_cowplot(font_size = 8) + 
  theme(legend.position = "nothing")

```


I need to visulaize these phenotypes in a better way. Thus I calculate the 2D median.

```{r}
drug_order <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Trametinib")) %>%
  filter(!grepl(drug, pattern = "Dabrafenib")) %>% 
  filter(!grepl(drug, pattern = "GSK1120212")) %>% 
    filter(line == "D046T01") %>% .$drug %>% unique() %>% sort()


center_df <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Trametinib")) %>%
      filter(!grepl(drug, pattern = "Dabrafenib")) %>% 
       filter(!grepl(drug, pattern = "GSK1120212")) %>%
  #filter(line == "D055T01") %>% 
    mutate(drug = factor(drug, levels =drug_order)) %>%
    # group_by(drug) %>% 
    # sample_n(200, replace = FALSE) %>% 
  dplyr::select(v1, v2, drug, line) %>% 
  group_by(drug, line) %>% 
  summarize(v1 = mean(v1),
            v2 = mean(v2))

gg_trametinib_path <- center_df %>%
  ggplot(aes(v1, v2, group = line)) + 
  
  scale_color_brewer(type = "seq", palette = "YlOrRd") + 
  geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point(aes(color = drug),alpha = 1, size = 3, shape=16) + 
  geom_path() +
  facet_wrap(~ line) +
  #geom_density2d(color = "black") + 
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) 
```

I want to create smoothed principal curves like in most trajectory inference methods, for example slingshot. 
This is the trajectory for Bortezomib:

```{r}
  


drug_order <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Bortezomib")) %>%
 .$drug %>% unique() %>% sort()

df_tmp <- df_cluster %>%
    filter(drug %in% drug_order) %>% 
  filter(line == "D055T01") %>%
  mutate(drug = factor(drug, levels =drug_order)) %>% 
  arrange((drug)) %>% 
  group_by(drug) %>%
  sample_n(1000, replace = TRUE) %>% 
  ungroup()

df_tmp_init <- df_tmp %>% 
  group_by(drug, line) %>% 
  summarize(v1 = mean(v1),
            v2 = mean(v2)) %>% 
  ungroup()

plot(df_tmp_init%>% dplyr::select(v1, v2) %>% as.matrix())
plot(df_tmp%>% dplyr::select(v1, v2) %>% as.matrix())

fit <- principal_curve(x = df_tmp %>% dplyr::select(v1, v2) %>% as.matrix(),
                       start = df_tmp_init %>% dplyr::select(v1, v2) %>% as.matrix(),
                       approx_points = FALSE, 
                       trace = TRUE, 
                       plot_iterations = TRUE, 
                       maxit = 30,
                       stretch =0)

plot(fit)
points(df_tmp_init%>% dplyr::select(v1, v2) %>% as.matrix())

fit$s[fit$ord,] %>% .[c(500:5000),] %>%
  as_tibble() %>%
  ggplot(aes(v1, v2)) + 
  geom_path() + 
  geom_point(data = df_tmp_init, size = 2)



```

Now I am building a function to create these tidy path objects.

```{r}
create_princurve_trace <- function(df){
  df_tmp_init <- df %>% 
  group_by(drug) %>% 
  summarize(v1 = mean(v1),
            v2 = mean(v2)) %>% 
  ungroup()
  
  fit <- principal_curve(x = df %>% dplyr::select(v1, v2) %>% as.matrix(),
                       start = df_tmp_init %>% dplyr::select(v1, v2) %>% as.matrix(),
                       approx_points = FALSE, 
                       trace = TRUE, 
                       plot_iterations = TRUE, 
                       maxit = 30,
                       stretch =2)
  
  plot(fit)
  points(df_tmp_init%>% dplyr::select(v1, v2) %>% as.matrix())

  result <- list()
  
  result$fit <- fit$s[fit$ord,] %>% as_tibble()
  result$center <- df_tmp_init
  result$input <- df
  
  return(result)
}
```


```{r}
gg_bortezomib_path + ggsave("bortezomib_path.pdf", width = 12, height =  7)
gg_trametinib_path + ggsave("trametininb_path.pdf", width = 12, height =  7)
```

It becomes clear that compounds have characteristic paths of drug response. 
To illustrate this, we focus on the organoids that start out with very disorganized archtecture and compare their response to bortezomib vs. trametinib. 


```{r}
drug_order <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Trametinib")| grepl(drug, pattern = "Bortezomib")) %>%
  filter(!grepl(drug, pattern = "Dabrafenib")) %>% 
  filter(!grepl(drug, pattern = "GSK1120212")) %>% .$drug %>% unique() %>% sort()

# center_df_mean <- df_cluster %>%
#     filter(drug %in% drug_order) %>%
#   filter(line %in% c("D055T01", "D054T01", "D046T01")) %>%
#   dplyr::select(v1, v2, drug, line) %>% 
#   group_by(drug) %>% 
#   summarize(v1 = mean(v1),
#             v2 = mean(v2)) %>% 
#   separate(drug, c("drug", "concentration"), sep = "__") %>% 
#   mutate(concentration = as.numeric(concentration))

center_df <- df_cluster %>%
    filter(drug %in% drug_order) %>%
  filter(line %in% c("D055T01", "D054T01", "D046T01")) %>%
  dplyr::select(v1, v2, drug, line) %>% 
  group_by(drug, line) %>% 
  summarize(v1 = mean(v1),
            v2 = mean(v2)) %>% 
  separate(drug, c("drug", "concentration"), sep = "__") %>% 
  mutate(concentration = as.numeric(concentration),
         drug_line = paste0(drug, "_", line))

gg_trambort <- center_df %>% 
  ggplot(aes(v1, v2)) + 
  geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point(aes(color = drug, group = drug_line),alpha = 1, size = 1.5, shape=16) + 
  #geom_smooth(aes(group = drug), se = FALSE) + 
  #geom_path(aes(group = drug, color = drug), arrow = arrow(angle =15, ends = "last", type = "closed", length = unit(0.15, "inches"))) + 
  geom_path(aes(group = drug_line, color = drug)) + 
  scale_color_manual(values = c("#707070", "#000000", "#4285F4")) +
  
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) 
```


I am creating a new version of the plot above

```{r}
drug_order <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Trametinib")| grepl(drug, pattern = "Bortezomib")) %>%
  filter(!grepl(drug, pattern = "Dabrafenib")) %>% 
  filter(!grepl(drug, pattern = "GSK1120212")) %>% .$drug %>% unique() %>% sort()

center_df <- df_cluster %>%
    filter(drug %in% drug_order) %>%
  filter(line %in% c("D055T01", "D054T01", "D046T01")) %>%
  mutate(drug = factor(drug, levels =drug_order)) %>% 
  arrange((drug)) %>% 
  group_by(drug, line) %>%
  sample_n(1000, replace = TRUE) %>% ungroup()

plot_df_bort <- center_df %>%
  filter(drug == "DMSO" | grepl(drug, pattern = "Bortezomib")) %>%
  filter(grepl(drug, pattern = "Bortezomib")) %>%
  nest(-line) %>%
  mutate(plot = purrr::map(data, ~ create_princurve_trace(.x)))

plot_df_tram <- center_df %>%
  filter(drug == "DMSO" | grepl(drug, pattern = "Trametinib")) %>% 
  filter(grepl(drug, pattern = "Trametinib")) %>%
  nest(-line) %>%
  mutate(plot = purrr::map(data, ~ create_princurve_trace(.x)))

gg_trambort_traj <- plot_df_tram %>% mutate(plot_trace = purrr::map(plot, ~ .x$fit %>% .[c(300:4500), ])) %>% unnest(plot_trace) %>%
  mutate(drug = "Trametinib") %>% 
  rbind(plot_df_bort %>% mutate(plot_trace = purrr::map(plot, ~ .x$fit %>% .[c(300:4000), ])) %>% unnest(plot_trace) %>%
  mutate(drug = "Bortezomib")) %>% 
  mutate(drug_line = paste0(drug, "_", line)) %>%
  filter(line == "D046T01") %>%
  ggplot(aes(v1, v2)) + 
  #geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  #geom_point(aes(color = drug, group = drug_line),alpha = 1, size = 1.5, shape=16) + 
  #geom_smooth(aes(group = drug), se = FALSE) + 
  geom_point(data = 
               plot_df_bort %>% unnest(data) %>% 
               rbind(plot_df_tram %>% unnest(data)) %>% 
               separate(drug, c("drug", "conc"), sep = "__"),
             aes(color = drug), alpha = 0.05, size = 0.35) +
  
  #geom_path(aes(group = drug, color = drug), arrow = arrow(angle =15, ends = "last", type = "closed", length = unit(0.15, "inches"))) + 
  geom_path(aes(group = drug_line, color = drug), size = 1.5, arrow = arrow(angle = 10, ends = "last", type = "closed", length = unit(0.15, "inches"))) + 
  scale_color_manual(values = c("#707070", "#4285F4")) +
  facet_grid(~ drug) +
  # geom_point(data = 
  #              plot_df_bort %>% mutate(center = map(plot, ~ .x$center)) %>% unnest(center) %>% 
  #              rbind(plot_df_tram %>% mutate(center = map(plot, ~ .x$center)) %>% unnest(center)) %>% 
  #              separate(drug, c("drug", "conc"), sep = "__"),
  #            aes(color = drug)) +
  
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) 
```

I create the same figure for Taxanes

```{r}
set.seed(12312)
drug_order <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Docetaxel")| grepl(drug, pattern = "Paclitaxel")) %>%
.$drug %>% unique() %>% sort()

center_df <- df_cluster %>%
    filter(drug %in% drug_order) %>%
  filter(line %in% c("D013T01", "D027T01")) %>%
  mutate(drug = factor(drug, levels =drug_order)) %>% 
  arrange((drug)) %>% 
  group_by(drug, line) %>%
  sample_n(1000, replace = TRUE) %>% ungroup()

plot_df_pacli <- center_df %>%
  filter(grepl(drug, pattern = "Paclitaxel")) %>%
  nest(-line) %>%
  mutate(plot = purrr::map(data, ~ create_princurve_trace(.x)))

plot_df_doce <- center_df %>%
  filter(grepl(drug, pattern = "Docetaxel")) %>%
  nest(-line) %>%
  mutate(plot = purrr::map(data, ~ create_princurve_trace(.x)))

gg_taxane_traj <- plot_df_pacli %>% mutate(plot_trace = purrr::map(plot, ~ .x$fit %>% .[c(300:4500), ])) %>% unnest(plot_trace) %>%
  mutate(drug = "Paclitaxel") %>% 
  rbind(plot_df_doce %>% mutate(plot_trace = purrr::map(plot, ~ .x$fit %>% .[c(300:4000), ])) %>% unnest(plot_trace) %>%
  mutate(drug = "Docetaxel")) %>% 
  mutate(drug_line = paste0(drug, "_", line)) %>%
  
  ggplot(aes(v1, v2)) + 
  geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  #geom_point(aes(color = drug, group = drug_line),alpha = 1, size = 1.5, shape=16) + 
  #geom_smooth(aes(group = drug), se = FALSE) + 
  geom_point(data = 
               plot_df_pacli %>% unnest(data) %>% 
               rbind(plot_df_doce %>% unnest(data)) %>% 
               separate(drug, c("drug", "conc"), sep = "__"),
             aes(color = drug), alpha = 0.05, size = 0.35) +
  
  #geom_path(aes(group = drug, color = drug), arrow = arrow(angle =15, ends = "last", type = "closed", length = unit(0.15, "inches"))) + 
  geom_path(aes(group = drug_line, color = drug), size = 1.5, arrow = arrow(angle = 10, ends = "last", type = "closed", length = unit(0.15, "inches"))) + 
  scale_color_manual(values = c("#F4B400", "#0F9D58")) +
  facet_grid(~ drug) +
  # geom_point(data = 
  #              plot_df_bort %>% mutate(center = map(plot, ~ .x$center)) %>% unnest(center) %>% 
  #              rbind(plot_df_tram %>% mutate(center = map(plot, ~ .x$center)) %>% unnest(center)) %>% 
  #              separate(drug, c("drug", "conc"), sep = "__"),
  #            aes(color = drug)) +
  
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) 
```



I am also interested in the OSI induced phenotype 

```{r}
drug_order <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Paclitaxel")) %>%
  #  filter(line == "D046T01") %>%
  .$drug %>% unique() %>% sort()

center_df <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Paclitaxel")) %>%
  #filter(line == "D046T01") %>% 
    mutate(drug = factor(drug, levels =drug_order)) %>%
    # group_by(drug) %>% 
    # sample_n(200, replace = FALSE) %>% 
  dplyr::select(v1, v2, drug, line) %>% 
  group_by(drug, line) %>% 
  summarize(v1 = median(v1),
            v2 = median(v2))

center_df %>%
  ggplot(aes(v1, v2, group = line)) + 
  
  scale_color_brewer(type = "seq", palette = "YlOrRd") + 
  geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point(aes(color = drug),alpha = 1, size = 3, shape=16) + 
  #facet_wrap(~ line) +
  #geom_path() +
  #geom_density2d(color = "black") + 
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) 
```

I also check out OSI-953

```{r}
drug_order <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "ONO-4059")) %>%
  #  filter(line == "D046T01") %>%
  .$drug %>% unique() %>% sort()

center_df <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "ONO-4059")) %>%
  #filter(line == "D046T01") %>% 
    mutate(drug = factor(drug, levels =drug_order)) %>%
    # group_by(drug) %>% 
    # sample_n(200, replace = FALSE) %>% 
  dplyr::select(v1, v2, drug, line) %>% 
  group_by(drug, line) %>% 
  summarize(v1 = median(v1),
            v2 = median(v2))

gg_ono <- center_df %>%
  filter(!(line %in% c("D021T01", "D010T01", "D018T01"))) %>%
  ggplot(aes(v1, v2, group = line)) + 
 
  scale_color_brewer(type = "seq", palette = "YlOrRd") + 
  geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  #geom_point(aes(color = drug),alpha = 1, size = 3, shape=16) + 
  # facet_wrap(~ line) +
  
  geom_line(aes(group = line), arrow = arrow(angle =15, ends = "first", type = "closed", length = unit(0.1, "inches")), color = "grey") +
  geom_line(data = center_df %>% filter(line == "D027T01"), aes(group = line), arrow = arrow(angle =15, ends = "first", type = "closed", length = unit(0.15, "inches")), color = "black") +
  #gghighlight::gghighlight(line == "D027T01") +
  #geom_density2d(color = "black") + 
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) 
```

I also add GSK inhibitors 

```{r}
drug_order <- df_cluster %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "CHIR-99021")) %>%
  #  filter(line == "D046T01") %>%
  .$drug %>% unique() %>% sort()

center_df <- df_cluster %>%
    filter(drug %in% drug_order) %>%
  #filter(line == "D046T01") %>% 
  mutate(drug = if_else(drug == "DMSO", "DMSO", "Chir")) %>%
    mutate(drug = factor(drug)) %>%
    # group_by(drug) %>% 
    # sample_n(200, replace = FALSE) %>% 
  dplyr::select(v1, v2, drug, line) %>% 
  group_by(drug, line) %>% 
  summarize(v1 = median(v1),
            v2 = median(v2))

gg_chir <- center_df %>%
  filter(!(line %in% c("D021T01", "D010T01", "D018T01"))) %>%
  ggplot(aes(v1, v2, group = line)) + 
 
  scale_color_brewer(type = "seq", palette = "YlOrRd") + 
  geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  #geom_point(aes(color = drug),alpha = 1, size = 3, shape=16) + 
  # facet_wrap(~ line) +
  
  geom_line(aes(group = line), arrow = arrow(angle =15, ends = "first", type = "closed", length = unit(0.1, "inches")), color = "grey") +
  geom_line(data = center_df %>% filter(line == "D027T01"), aes(group = line), arrow = arrow(angle =15, ends = "first", type = "closed", length = unit(0.15, "inches")), color = "black") +
  #gghighlight::gghighlight(line == "D027T01") +
  #geom_density2d(color = "black") + 
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) 
```

# Example image generation. 

Based on the clustering I have produced, I want to visualize example objects belonging to each cluster. 

```{r}
source(here("R/display_organoid_complete.R"))
```

Now I create a list of organoids I am interested in. 

```{r}

```


# Cluster enrichment 

## Size per cluster 

```{r}
df_cluster %>% 
  dplyr::select(mclust_complete_classification, size) %>% 
  group_by(mclust_complete_classification) %>%
  summarise(size = mean(size)) %>% 
  arrange(size) %>% 
  mutate(mclust_complete_classification = factor(mclust_complete_classification) %>% fct_inorder()) %>% 
  ggplot(aes(mclust_complete_classification, size)) + 
  geom_point() + 
  theme_cowplot()
```

## Example organoids 

I tidy the membership porbability table. 

```{r, eval = FALSE}
score_df <- clust_predict$z %>% as.data.frame() %>% 
  rownames_to_column("id") %>% mutate(id = as.numeric(id)) %>% gather(m, prob, -id) %>% nest(-id) %>% mutate(score = furrr::future_map(data, ~ arrange(.x, desc(prob)) %>% head(1))) %>% 
  unnest(score) %>% dplyr::select(-data)
```


```{r}
library(imager)
html_dir = "/data/rauscher/promise/PROMISE/data-10x-4t-c-16z/htmldir/"

hdf5_all_uwot_stereotypes <- df_cluster%>%
  mutate(OriginalX = as.numeric(original_x),
         OriginalY = as.numeric(original_y),
         Field = field) %>%
  # I scale the information to the image's size and add 50% of the y-dim to target the first field
    mutate(ScaleX = round((OriginalX/2035.5)*(514/2)),
         ScaleY = round((OriginalY/2035.5)*(514/2))) %>%
  mutate(ScaleX = case_when(Field %in% c("1", "4") ~ ScaleX,
                            Field %in% c("2", "3") ~ ScaleX + (514/2)),
         ScaleY = case_when(Field %in% c("1", "2") ~ ScaleY,
                            Field %in% c("3", "4") ~ ScaleY + (514/2))) %>%
  mutate(lowerX = ScaleX - round(sqrt(as.numeric(size))/4),
         upperX = ScaleX + round(sqrt(as.numeric(size))/4), 
         lowerY = ScaleY - round(sqrt(as.numeric(size))/4), 
         upperY = ScaleY + round(sqrt(as.numeric(size))/4)) %>%
  filter(lowerX > 0,
         lowerY > 0,
         upperX < 514,
         upperY < 514) %>%
  group_by(mclust_complete_classification) %>% 
  #do(arrange(., desc(prob)) %>% head(10)) %>% 
    sample_n(10) %>%
  mutate(path_to_html = paste0(html_dir, bc, "/", bc, "_", substr(Well, 1,1), "_", substr(Well, 2,3), "_1.jpeg")) 

hdf5_all_uwot_stereotypes_img <- hdf5_all_uwot_stereotypes %>% ungroup() %>%
  #mutate(img_raw = purrr::map(path_to_html, ~ imager::load.image(.x))) %>%
  # mutate(img = purrr::map(img_raw, ~ .x %>% as.data.frame(wide = "c") %>%
  #          mutate(rgb.val = rgb(c.1, c.2, c.3))))
  mutate(img_raw = purrr::pmap(list(p = path_to_html, a =lowerX, b = upperX, c = lowerY, d= upperY, e = bc, f = cluster, g = index, h = score), function(p, a,b,c,d, e,f,g, h) EBImage::readImage(p) %>% 
                                 .[c(a:b), c(c:d),] %>%
                                 EBImage::writeImage(., files = here(paste0("data/cluster_tiles/", f, "_", e, "_", g, "_", round(h, 2), ".jpeg")))))
  

lapply(hdf5_all_uwot_stereotypes_img$img_raw, EBImage::display, method = "raster")

```



## Drug target

I wonder wether certain drugs are enriched in the 2D space I am observing

First I save my results.

```{r}
saveRDS(df_cluster, here("data/clustered_umap.Rds"))
```

Then I execute my contingency analysis

```{r, eval = FALSE}
chisq_drug <- df_cluster %>% 
  dplyr::select(mclust_complete_classification, drug) %>%
  filter(drug != "DMSO") %>%
  filter(drug != "Staurosporine_500nM") %>%
  #mutate(mclust_complete_classification = as.numeric(mclust_complete_classification)) %>%
  group_by(mclust_complete_classification, drug) %>% 
  summarize(n = n()) %>% 
  spread(drug, n, fill = 0) %>%
  as.data.frame() %>%
  column_to_rownames("mclust_complete_classification") %>% 
  as.matrix() %>%
  chisq.test()
  #stats::fisher.test(simulate.p.value = TRUE)
```

# Chi-Square 

```{r}
sd <- chisq_drug$residuals %>% as.matrix() %>% sd()

chi_tidy <- chisq_drug$residuals %>% as.data.frame() %>% 
  rownames_to_column("cluster") %>% 
  gather(drug, residual, -cluster)

chi_tidy %>% group_by(cluster) %>% 
  arrange(desc(residual)) %>% filter(cluster == "M1") #dead
```

```{r}
chi_tidy %>% group_by(cluster) %>% 
  arrange(desc(residual)) %>% filter(cluster == "M2")
```

```{r}
chi_tidy %>% group_by(cluster) %>% 
  arrange(desc(residual)) %>% filter(cluster == "M3") # dead
```

```{r}
chi_tidy %>% group_by(cluster) %>% 
  arrange(desc(residual)) %>% filter(cluster == "M4") # dense morphology #PF-04691502 #PP121 
```

```{r}
chi_tidy %>% group_by(cluster) %>% 
  arrange(desc(residual)) %>% filter(cluster == "M5") #dead
```

```{r}
chi_tidy %>% group_by(cluster) %>% 
  arrange(desc(residual)) %>% filter(cluster == "M6")
```


# Non- Chi-Sqaure

I explore the distribution of counts for each replicate

```{r}
df_drug_count <- df_cluster %>% 
  dplyr::select(mclust_complete_classification, drug, replicate) %>%
  filter(drug != "DMSO") %>%
  filter(drug != "Staurosporine_500nM") %>%
  filter(!grepl(drug, pattern = "__")) %>%
  group_by(mclust_complete_classification, drug, replicate) %>% 
  summarize(n = n()) 
```

I examine the dispersion of count distributions for each cluster across samples. Given the heterogeneity of dispersions, it might be useful to perform differential abundance analysis in a cluster-by-cluster fashion, as qCML assumes multiple small samples with similar dispesion.

```{r}
df_drug_count %>% 
  ggplot(aes(n, color = mclust_complete_classification)) + 
  geom_density() + 
  scale_x_log10() +
  theme_cowplot() + 
  scale_color_viridis_d()
```

I log transform the count data and plot histograms another time. As this is not sequencing data, there is no need to perform RNA-Seq specific preprocessing steps, such as depth and read-length adjustments. 

```{r}
df_drug_count %>%
  mutate(log_n = log(n)) %>% 
  ggplot(aes(mclust_complete_classification, log_n)) + 
  geom_boxplot()
```

Next, I remove drugs with a count below 50, as no statistical test would be useful in this case. It turns out that there are no drugs that do not meet this threshold.

```{r}
df_drug_count %>%
  ungroup() %>%
  group_by(drug) %>% 
  summarise(n = sum(n)) %>%
  arrange(n)
```

I continue to import the data into an biocodnuctor format to perform further processing. 

```{r}
library(edgeR)

df_drug_count_matrix <- df_drug_count %>%
  mutate(log_n = log(n)) %>% 
  ungroup() %>%
  mutate(cluster_replicate = paste0(as.character(mclust_complete_classification), "_", replicate)) %>%
  dplyr::select(-log_n, -mclust_complete_classification, -replicate) %>%
  spread(drug, n, fill = 0) %>% 
  as.data.frame() %>%
  column_to_rownames("cluster_replicate") %>% 
  t() %>% as.matrix()
  
dda <- DGEList(df_drug_count_matrix, 
        lib.size = NULL,
        group = rep(c(1,10,11,12,13,14,15, 2:9),each=2))
```

Now I follow thorugh a standard multi-group comparison workflow using edgeR.

```{r}
dda_norm <- calcNormFactors(dda)  
dda_norm$samples$norm.factors
lcpm_norm <- cpm(dda_norm, log=TRUE)
```

I compare the normalized data to unnormalized data. 

```{r}
lcpm <- cpm(dda, log=TRUE)

lcpm %>% as_tibble() %>% 
  gather(cluster, value) %>%
  mutate(normalization = FALSE) %>% 
  rbind(lcpm_norm %>% as_tibble() %>%
  gather(cluster, value) %>%
  mutate(normalization = TRUE)) %>% 
  ggplot(aes(value, color = cluster)) + 
  geom_density() + 
  facet_wrap(~ normalization) +
  scale_color_viridis_d() +
  theme_cowplot()
```

I create a small overview

```{r}
plotMDS(lcpm_norm)
```

Based on perturbations alone, we can identify clusters with similar patterns.

Now I run a test. 

```{r}
cluster <- factor(paste0("M", rep(c(1,10,11,12,13,14,15, 2:9),each=2)))
replicate <- factor(rep(c(1:2), times = 15))

design <- model.matrix(~ 0 + cluster + replicate)

colnames(design) <- c(paste0("M", rep(c(1,10,11,12,13,14,15, 2:9),each=1)), "replicate")

contrast_matrix <- makeContrasts(#"m1"=M1-(M2+M3+M4+M5+M6+M7+M8+M9+M10+M11+M12+M13+M14+M15)/14,
                            "m2"=M2-(M1+M3+M4+M5+M6+M7+M8+M9+M10+M11+M12+M13+M14+M15)/14,
                            "m3"=M3-(M1+M2+M4+M5+M6+M7+M8+M9+M10+M11+M12+M13+M14+M15)/14,
                           #"m4"=M4-(M1+M2+M3+M5+M6+M7+M8+M9+M10+M11+M12+M13+M14+M15)/14,
                           #"m5"=M5-(M1+M2+M3+M4+M6+M7+M8+M9+M10+M11+M12+M13+M14+M15)/14,
                           "m6"=M6-(M1+M2+M3+M4+M5+M7+M8+M9+M10+M11+M12+M13+M14+M15)/14,
                           #"m7"=M7-(M1+M2+M3+M4+M5+M6+M8+M9+M10+M11+M12+M13+M14+M15)/14,
                            "m8"=M8-(M1+M2+M3+M4+M5+M6+M7+M9+M10+M11+M12+M13+M14+M15)/14,
                           # "m9"=M9-(M1+M2+M3+M4+M5+M6+M7+M8+M10+M11+M12+M13+M14+M15)/14,
                           "m10"=M10-(M1+M2+M3+M4+M5+M6+M7+M8+M9+M11+M12+M13+M14+M15)/14,
                           #"m11"=M11-(M1+M2+M3+M4+M5+M6+M7+M8+M9+M10+M12+M13+M14+M15)/14,
                           "m12"=M12-(M1+M2+M3+M4+M5+M6+M7+M8+M9+M10+M11+M13+M14+M15)/14,
                           # "m13"=M13-(M1+M2+M3+M4+M5+M6+M7+M8+M9+M10+M11+M12+M14+M15)/14,
                           # "m14"=M12-(M1+M2+M3+M4+M5+M6+M7+M8+M9+M10+M11+M12+M13+M15)/14,
                            "m15"=M13-(M1+M2+M3+M4+M5+M6+M7+M8+M9+M10+M11+M12+M13+M14)/14,
                           levels=design)
```

I run a test

```{r}
par(mfrow=c(1,2))
v <- voom(dda, design, plot=TRUE)
vfit <- lmFit(v, design)
vfit <- contrasts.fit(vfit, contrasts=contrast_matrix)
efit <- eBayes(vfit)
plotSA(efit, main="Final model: Mean-variance trend")
```

Now I check the results. 

```{r}
df <- topTable(efit, number=Inf, coef= "m15") %>% as.data.frame() %>% rownames_to_column("drug")%>% as_tibble() %>% mutate(log_p = log(adj.P.Val))

df %>%
  ggplot(aes(t, log_p)) + 
  geom_point()
```

```{r}
df <- topTable(efit, number=Inf) %>% as.data.frame() %>% rownames_to_column("drug") %>% filter(adj.P.Val < 0.001) %>% dplyr::select(-(AveExpr:adj.P.Val)) %>% 
  column_to_rownames("drug") 

df %>%
  pheatmap::pheatmap(filename = "drug_heatmap.pdf", width = 20, height = 20, scale = "none")
```


```{r}
df_cluster_drugs <- df %>% rownames_to_column("product_name") %>% 
  gather(cluster, logFC, -product_name) %>%
  left_join(kistem %>% select(product_name, target,  pathway)) %>% 
  distinct()
  
```

```{r}
df_cluster_drugs %>% 
  arrange(logFC) %>%
  filter(abs(logFC) > 1) %>%
  filter(cluster == "m2")
```


I load the results

```{r}
fisher <- readRDS(here("data/drug_contig/fisher.Rds"))
```


```{r}



residual_drug_df <- chisq_drug$residuals %>% as.data.frame() %>%
  rownames_to_column("cluster") %>%
  as_tibble() %>% 
  gather(drug, residual, - cluster) %>% 
  arrange(desc(residual))

df <- residual_drug_df %>% 
  nest(-cluster) %>% 
  mutate(cluster = as.numeric(cluster)) %>%
  arrange(cluster) %>%
  filter(cluster == 2) %>% #13 Wnt, #11 dead, #12 (not cluster) mTOR
  .$data %>% 
  as.data.frame() %>%
  mutate(drug = factor(drug) %>% fct_inorder()) 

gg_chi <- df %>%
  ggplot(aes(drug, residual)) + 
  geom_point() + 
  theme_classic() +
  ggrepel::geom_text_repel(data = df %>% head(5) %>% rbind(df %>% tail(5)), aes(label = drug)) +
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_blank(),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) + 
  geom_hline(yintercept = 0)
  
  
```

```{r}
df_cluster$mclust_complete_classification %>% table()
```


```{r}
gg_clust2 <- df_cluster %>% 
  sample_frac(0.01) %>%
  filter(mclust_complete_classification == "2") %>%
  ggplot(aes(v1, v2, color = mclust_complete_classification)) + 
  geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") +
  geom_point_rast(alpha = 1, size = .35, shape=16) + 
  
  #facet_wrap(~ mclust_complete_classification) +
  #geom_path() +
  #geom_density2d(color = "black") + 
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme(axis.title.x = element_text(color = "black", size = 8),
        axis.title.y = element_text(color = "black", size = 8),
        axis.text.y = element_text(color = "black", size = 8),
        axis.text.x = element_text(color = "black", size = 8),
        legend.text = element_text(color = "black", size = 8),
        legend.title = element_text(color = "black", size = 8)) 
  
```


```{r}
clust$data %>%
  as.data.frame() %>%
  cbind(clust$classification) %>% 
  cbind(clust$uncertainty) %>% 
  janitor::clean_names() %>% 
  mutate(clust_uncertainty = (1-clust_uncertainty)*0.1,
         clust_classification = factor(clust_classification, levels = milestone_order)) %>%
  filter(clust_classification == 6) %>%
  ggplot(aes(v1, v2, color = clust_classification)) + 
  geom_point_rast(data = df_cluster %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") +
  geom_point_rast(size = 0.35, alpha = 0.1) + 
  theme_classic()
```


```{r, eval = FALSE}
scheme <- rwantshue::iwanthue(seed = 42, force_init = TRUE) # recreate with a seed
# generate a new color palette (vector of hex values) with presets...


gg_cluster <- plot_cells(ods_sub, color_cells_by="cluster",
           alpha = 0.05) + 
  scale_color_manual(values = scheme$hex(23))
```

I load clustering results and plot them. 

```{r}
ods_sub_clust <- readRDS(here("R/ods_sub_clust_100"))
gg_cluster <- plot_cells(ods_sub_clust, color_cells_by="cluster", rasterize = TRUE,
           alpha = 0.5)
```

I create a graph following the following schema. I run this process in a separate script to speed up processing.

```{r, eval = FALSE}
ods_sub = learn_graph(ods_sub, verbose = TRUE, 
                        close_loop = TRUE, 
                        learn_graph_control = list(minimal_branch_len = 100))
```


```{r}
ods_sub_graph <- readRDS(here("R/ods_sub_graph_10"))
gg_path <- plot_cells(ods_sub_graph, rasterize = TRUE)
```


I save my results 

```{r}
saveRDS(ods_sub, here("data/monocle/bulk_pca_path_subsample.Rds"))
```

# Plot panel

I create a final graph

```{r}
# fig5 <- cowplot::plot_grid(gg_size, gg_line, 
#                    #gg_path, #gg_proportion,
#           gg_bortezomib,#gg_borte_prop,
#           cowplot::plot_grid(gg_trambort_traj , gg_taxane_traj,
#                              NULL, NULL,
#                            #  gg_ono, gg_chir, 
#                              align = "h", axis = "bt", labels = "AUTO", ncol = 2),
#           #gg_chi, gg_clust2,
#           #gg_path, #gg_proportion,
#           #gg_shift, gg_ridge, 
#           NULL, NULL,
#           align = "h", axis = "bt", labels = c("A", "B"), ncol = 2)
#   

# fig5 <- plot_grid(
#             plot_grid(gg_size + theme(legend.position = "nothing"), gg_line,
#                        ncol = 2,
#                        labels = c('A', 'B')),
#             gg_bortezomib + theme(legend.position = "nothing"),
#             # plot_grid(gg_trambort_traj,
#             #           gg_taxane_traj,
#             #           
#             #           NULL, NULL,
#             #           ncol = 2,
#             #           labels = c('D', "E", "F", "G",
#             #                      "H", "I", "J", "K")),
#             ncol = 1,
#             align = "h", axis = "bt", labels = c("", "C", ""))

fig5 <- plot_grid(NULL, 
                  NULL,
                  # plot_grid(gg_line, 
                  #           gg_bortezomib,
                  #           ncol = 2),
                  gg_mtor_all,
                  NULL, # NULL,
                  ncol = 2)

fig5 + 
  ggsave(here("panels/figure_5.pdf"), 
         width = 8.3, 
         height = 11.7)
```

```{r}
gg_trambort_traj + theme_cowplot() theme(legend.position = "none") + ggsave(here("panels/figure_5_aux.pdf"), width = 8, height = )
```


I create a supplemental figure

```{r, eval = TRUE}
cowplot::plot_grid(cowplot::plot_grid(gg_size_dist,
                                      NULL,
                                      ncol = 2),
                   gg_line_overview,
                   cowplot::plot_grid(gg_taxane_traj + theme_cowplot() + theme(legend.position = "none"),
                                      NULL,
                                      ncol = 2), 
                   align = "h", axis = "bt", labels = "AUTO", ncol = 1) + 
  ggsave("figure_S5.pdf", width = 6, height = 12)
```

