---
title: "MOFA drug effect"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(MOFA2)
library(here)
#library(MOFAdata)
library(ReactomePA)
library(biomaRt)
library(gridExtra)
```


```{r}
# input
model <- load_model(here("models/mofa/model.hdf5"))

umap_df <- readRDS(here::here("data/processed/PhenotypeSpectrum/umap_absolute_all_drugs_sampled.Rds"))
promise_long_filtered_top <- readRDS(here::here('data/processed/expression/promise_expr_filtered_tidy_top.rds'))

weights <- model@expectations$Z$single_group %>% as.data.frame() %>% rownames_to_column("id") %>% as_tibble() %>% janitor::clean_names()
loadings_size <- model@expectations$W$size_view %>% as.data.frame() %>% rownames_to_column("id") %>% as_tibble() %>% janitor::clean_names()
loadings_expression <- model@expectations$W$expression %>% as.data.frame() %>% rownames_to_column("id") %>% as_tibble() %>% janitor::clean_names()
loadings_morphology <- model@expectations$W$morphology %>% as.data.frame() %>% rownames_to_column("id") %>% as_tibble() %>% janitor::clean_names()
loadings_drug_activity <- model@expectations$W$drug_activity %>% as.data.frame() %>% rownames_to_column("id") %>% as_tibble() %>% janitor::clean_names()
loadings_mutation <- model@expectations$W$mutation %>% as.data.frame() %>% rownames_to_column("id") %>% as_tibble() %>% janitor::clean_names()


```

# Inference on organoid level

```{r, eval = FALSE}
new_morphology <- readRDS(here::here("data/processed/PhenotypeSpectrum/pca_absolute_all_drugs_sampled.Rds")) %>% 
  dplyr::select(contains("pc"), uuid) %>% 
  column_to_rownames("uuid") %>%
  as.matrix() %>% scale(center = TRUE, scale = FALSE)

z_projected_organoid <-  new_morphology %*% t(Winv) %>% as.data.frame()%>% rownames_to_column("uuid")

umap_factor <- umap_df %>% 
  inner_join(z_projected_organoid) %>%
  filter(line != "D020T02") 

gg_factor_umap <- umap_factor %>% 
  dplyr::select(-size_factor) %>%
  dplyr::rename(factor1 = V1, factor2 = V2, factor3 = V3) %>%
  pivot_longer(cols = contains("factor"), names_to = "number", values_to = "value") %>%
  ggplot(aes(v1, v2, color = log(value))) + 
  ggrastr::geom_point_rast(alpha = 0.1, size = 0.35) + 
  scale_color_viridis_c() +
  cowplot::theme_cowplot() +
  labs(x = "UMAP 1",
       y = "UMAP 2",
       color = "factor value") + 
  facet_wrap(~ number) +
  theme(legend.position = "bottom") + 
    coord_fixed()

gg_factor_umap + ggsave(here("factor_overview_organoid_level.pdf"), height = 6 , width = 6)
```


# Inference on treatment-level

```{r}
W <- get_weights(model, "morphology_view")[[1]]
Winv <- pracma::pinv(W)

# Damit kannst du dann erstmal schauen, wo die Organoide im latenten Raum liegen (z_projected).
# Falls das Sinn macht, kannst du dann mit den gelernten Gewichten vom RNA view via Z_projected  * W_RNA  die RNA Werte vorhersagen und mit den Messungen, die du fuer einzelne Drugs hast, vergleichen, i.e.



new_morphology <- readRDS(here::here("data/processed/PhenotypeSpectrum/pca_absolute_all_drugs_sampled.Rds")) %>% 
  # dplyr::filter(drug %in% c("DMSO") | (drug %in% c("WYE-125132 (WYE-132)",
  #                                                  "Trametinib (GSK1120212)") & 
  #                                       line %in% c("D046T01", "D018T01"))) %>% 
  # dplyr::filter(drug %in% c("DMSO") | drug %in% c("WYE-125132 (WYE-132)",
  #                                                  "Trametinib (GSK1120212)",
  #                                                 "AZD3759")) %>%
  #dplyr::filter(drug %in% c("DMSO")) %>%
  filter(concentration == "nan") %>%
  mutate(rep = paste0("r", replicate)) %>% 
  mutate(line = substr(line, 1, 4)) %>% 
  mutate(line = paste0(line, "_", rep)) %>%
  # I treat every drug induced phenotype as a dedicated line
  mutate(line = if_else(drug == "DMSO", line, paste0(line, "_", drug))) %>%
  #filter(size_log > 7.5) %>%
  group_by(line) %>% 
  summarise_at(vars(contains("pc")), funs(mean)) %>% 
  ungroup() %>% column_to_rownames("line") %>% as.matrix() %>% 
  #scale(center = TRUE, scale = FALSE) %>% 
  sweep(2, model@intercepts$morphology_view %>% unlist() %>% unname())

z_projected <-  new_morphology %*% t(Winv) %>% as.data.frame()%>% rownames_to_column("id")
```

# Inference of gene expression

```{r}
z_projected_matrix <- z_projected %>% column_to_rownames("id") %>% as.matrix()

expression_inference <- z_projected_matrix %*% t(get_weights(model, "expression")[[1]]) %>% as.data.frame() %>% rownames_to_column("id") %>% as_tibble() %>% 
  separate(id, c("line", "drug"), sep = 7) %>% 
  mutate(drug = substr(drug , 2, nchar(drug))) %>% 
  mutate(drug = ifelse(drug == "", "DMSO", drug))
```

```{r}

```

```{r}
expression_inference_tidy <- expression_inference %>%  gather(symbol, expr, -drug, -line) %>% mutate(symbol = substr(symbol, 1, nchar(symbol)-nchar("_expression"))) 

expression_dmso_inference <- promise_long_filtered_top %>% 
  dplyr::select(-line) %>% 
  dplyr::rename(line = id) %>% 
  dplyr::select(line, expr, symbol) %>% 
  left_join(expression_inference_tidy %>% filter(drug == "DMSO") %>% dplyr::rename(inference = expr))
```

```{r}
expression_dmso_inference %>% 
  drop_na() %>%
  ggplot(aes(expr, inference)) + 
  geom_point(alpha = 0.2) + 
  geom_smooth(aes(group = line))
```



```{r}
z_projected_anno <- z_projected %>% 
  mutate(target = substr(id, 9, nchar(id))) %>% 
  mutate(line = substr(id, 1, 7)) %>%
  mutate(target = ifelse(target == "", "DMSO", target))

z_projected_anno %>% 
  filter(target == "DMSO") %>%
  left_join(weights) %>% 
  dplyr::select(-target, -line) %>% 
  gather(variable, value, -id) %>% 
  mutate(factor = if_else(variable %in% c("V1", "factor1"), "factor1", "none")) %>%
  mutate(factor = if_else(variable %in% c("V2", "factor2"), "factor2", factor)) %>%
  mutate(factor = if_else(variable %in% c("V3", "factor3"), "factor3", factor)) %>% 
  mutate(source = if_else(grepl(variable, pattern = "factor"), "MOFA", "pseudoinverse")) %>% 
  dplyr::select(-variable) %>%
  spread(source, value) %>% 
  ggplot(aes(MOFA, pseudoinverse)) + 
  geom_point() + 
  facet_wrap(~ factor) + 
  theme_cowplot() + 
  geom_smooth(method = "lm")

```


```{r}
## load drug annotation
drug_anno <- readxl::read_excel(here::here('references/layouts/Compound_Annotation_Libraries_New.xlsx')) %>% distinct(drug = `Compound name`, target = `Primary Target`)

z_projected_anno_tidy <- z_projected_anno %>% 
  rename(drug = target) %>%
  left_join(drug_anno)

z_projected_anno_test <- z_projected_anno_tidy %>% semi_join(z_projected_anno_tidy %>% 
                                                               dplyr::select(target, drug) %>%
                                                               distinct() %>%
                                            dplyr::count(target) %>% 
                                            filter(n >=5)) %>% 
  filter(drug != "DMSO")
```

```{r}
drug_projection_test_result <- rbind(
  lm(V1 ~ target, data = z_projected_anno_test) %>% summary() %>% broom::tidy() %>% mutate(factor = "factor1"),
  lm(V2 ~ target, data = z_projected_anno_test) %>% summary() %>% broom::tidy() %>% mutate(factor = "factor2"),
  lm(V3 ~ target, data = z_projected_anno_test) %>% summary() %>% broom::tidy() %>% mutate(factor = "factor3")) 

drug_projection_test_result <- drug_activity_test_result %>% 
  filter(term != "(Intercept)") %>% 
  dplyr::select(term, statistic, factor, p.value)
```

```{r}
hist(drug_projection_test_result$p.value)
```

```{r}
moa_factor_shift <- drug_projection_test_result %>% arrange(p.value) %>% 
  mutate(fdr = p.adjust(p.value, method = "BH")) %>% 
  filter(fdr <= 0.2) %>% 
  head(15) %>%
  tableGrob(., theme = ttheme_default(), rows = NULL)

grid.arrange(moa_factor_shift)
```

aurora kinase, PLK, CDK, leads to reduced factor 1 - smaller organoids
mTORi leads to increased factor 1 - larger organoids?
MEK, FAK leads to increased factor 2 - more cystic organoids


Wnt inhibitors lead to reduced factor 3 - 

```{r}
doi <- c("MEK", "ctrl")

z_projected_anno_tidy %>% 
  mutate(target = if_else(drug == "DMSO", "ctrl", target)) %>%
  left_join(drug_projection_test_result %>% rename(target = term)) %>%
  #filter(line %in% c("D004_r1", "D004_r2", "D007_r1", "D007_r2", "D019_r1", "D019_r2")) %>%
  filter(target %in% doi) %>%
  ggplot(aes(target, V2, label = id)) + 
  geom_jitter(width = 0.1) + 
  ggsignif::geom_signif(comparisons = list(doi)) + 
  theme_cowplot() + 
  labs(y = "factor 2")


```

```{r}
new_morphology <- readRDS(here::here("data/processed/PhenotypeSpectrum/pca_absolute_all_drugs_sampled.Rds")) %>% 
  # dplyr::filter(drug %in% c("DMSO") | (drug %in% c("WYE-125132 (WYE-132)",
  #                                                  "Trametinib (GSK1120212)") & 
  #                                       line %in% c("D046T01", "D018T01"))) %>% 
  # dplyr::filter(drug %in% c("DMSO") | drug %in% c("WYE-125132 (WYE-132)",
  #                                                  "Trametinib (GSK1120212)",
  #                                                 "AZD3759")) %>%
  #dplyr::filter(drug %in% c("DMSO")) %>%
  filter(concentration != "nan") %>%
  mutate(rep = paste0("r", replicate)) %>% 
  mutate(line = substr(line, 1, 4)) %>% 
  mutate(line = paste0(line, "_", rep)) %>%
  # I treat every drug induced phenotype as a dedicated line
  mutate(line = if_else(drug == "DMSO", line, paste0(line, "_", drug, "_", concentration))) %>%
  #filter(size_log > 7.5) %>%
  group_by(line) %>% 
  summarise_at(vars(contains("pc")), funs(mean)) %>% 
  ungroup() %>% column_to_rownames("line") %>% as.matrix()

z_projected_concentration <-  new_morphology %*% t(Winv) %>% as.data.frame()%>% rownames_to_column("id")

z_projected_conc_anno <- z_projected_concentration %>% 
  mutate(target = substr(id, 9, nchar(id))) %>% 
  separate(target, c("target", "concentration"), sep = "_") %>% 
  mutate(concentration = as.numeric(concentration)) %>%
  mutate(line = substr(id, 1, 7)) %>%
  mutate(target = ifelse(target == "", "DMSO", target))

z_projected_conc_anno_tidy <- z_projected_conc_anno %>% 
  rename(drug = target) %>%
  left_join(drug_anno)

```

```{r}
z_projected_conc_anno_tidy %>% 
  filter(drug %in% c("DMSO", "Binimetinib")) %>%
  mutate(concentration = if_else(drug == "DMSO", 0.00032, concentration)) %>% #adding a pseudo conc for plotting
  ggplot(aes(concentration, V2)) + 
  geom_jitter(width = 0.1, aes(color = drug)) + 
  scale_x_log10() + 
  geom_smooth(color = "black") + 
  geom_hline(yintercept = 0) +
  theme_cowplot() + 
  labs(y = "factor 2") + 
  scale_color_manual(values = c("black", "grey"))
```



```{r}
doi <- c("DMSO", "Binimetinib")

z_projected_conc_anno_tidy %>% 
  filter(drug %in% doi) %>%
  filter(concentration == 0.2 | drug == "DMSO") %>%
  ggplot(aes(drug, V2, label = id)) + 
  geom_jitter() + 
  ggsignif::geom_signif(comparisons = list(doi))
```

```{r}
umap_df %>%
  filter(drug == "Binimetinib") %>%
  mutate(concentration = as.numeric(concentration)) %>%
  ggplot(aes(concentration, size_log)) + 
  geom_jitter(width = 0.1) + 
  stat_summary(fun.y=mean, colour="blue", geom="line", size = 1) + # draw a mean line in the data
  scale_x_log10() + 
  labs(title = "Binimetinib") + 
  theme_cowplot()
```
```{r}
umap_df %>%
  filter(drug == "Binimetinib") %>%
  filter(prob_live == 1) %>%
  mutate(concentration = as.numeric(concentration)) %>%
  ggplot(aes(concentration, size_log)) + 
  geom_jitter(width = 0.1) + 
  stat_summary(fun.y=mean, colour="blue", geom="line", size = 1) + # draw a mean line in the data
  scale_x_log10() + 
  labs(title = "Binimetinib, filtering for alive organoids") +
  #facet_grid(prediction ~ line) + 
  theme_cowplot()
```



```{r}
doi <- c("mTOR", "ctrl")

z_projected_anno_tidy %>% 
  mutate(target = if_else(drug == "DMSO", "ctrl", target)) %>%
  left_join(drug_projection_test_result %>% rename(target = term)) %>%
  #filter(line %in% c("D004_r1", "D004_r2", "D007_r1", "D007_r2", "D019_r1", "D019_r2")) %>%
  filter(target %in% doi) %>%
  ggplot(aes(target, V1)) + 
  geom_jitter(width = 0.2) + 
  ggsignif::geom_signif(comparisons = list(doi)) + 
  labs(y = "factor 1") + 
  theme_cowplot()
```

```{r}
doi <- c("mTOR", "ctrl")

z_projected_anno_tidy %>% 
  mutate(target = if_else(drug == "DMSO", "ctrl", target)) %>%
  left_join(drug_projection_test_result %>% rename(target = term)) %>%
  #filter(line %in% c("D004_r1", "D004_r2", "D007_r1", "D007_r2", "D019_r1", "D019_r2")) %>%
  filter(target %in% doi) %>% filter(drug %in% c("DMSO", "WYE-125132 (WYE-132)")) %>% 
  
  ggplot(aes(target, V1, label = id, color = target)) + 
  geom_jitter() + 
  ggsignif::geom_signif(comparisons = list(doi))
```



```{r}
z_projected_anno %>%
  filter(target %in% c("CTRL", "Trametinib")) %>%
  ggplot(aes(V1, V2, color = target)) + 
  geom_point() + 
  geom_path(aes(group = line), color = "black")
```



```{r}
z_projected %>% ggplot(aes(V1, V3, label = id)) + 
  geom_point(size = 4) + 
  ggrepel::geom_text_repel(color = "black") + 
  #scale_color_viridis_c() + 
  theme_cowplot()
```


# QC

```{r}
plot_data_overview(model)
```

```{r}
head(model@cache$variance_explained$r2_total[[1]])
```


Breakdown by factors 

```{r}
model@cache$variance_explained$r2_per_factor$single_group
```

```{r}
model@cache$variance_explained$r2_per_factor$single_group %>% as.data.frame() %>% 
  rownames_to_column("factor") %>% 
  mutate(overall_variance = expression + morphology_view + size_view + drug_activity) %>%
  mutate(overall_variance = overall_variance/4) %>%
  ggplot(aes(factor, overall_variance)) + 
  geom_point(size = 2) + 
  theme_cowplot()
```


```{r}
gg_ve <- plot_variance_explained(model, x="view", y="factor")
gg_ve
```

```{r}
ph_cor <- weights %>% 
  as.data.frame() %>% 
  column_to_rownames("id") %>% 
  cor() %>% 
  pheatmap::pheatmap()
```

```{r}
weights %>% 
  as.data.frame() %>% 
  column_to_rownames("id") %>% 
  cor() %>% as.vector() %>% 
  .[. != 1] %>% 
  summary()
```


```{r}
plot_factor(model, 
  factor = 1:3
)
```

# factor overview

```{r}
umap_factor <- weights %>% 
  separate(id, c("line", "replicate"), sep = "_") %>% 
  mutate(line = paste0(line, "T01"),
         replicate = substr(replicate, 2,2)) %>% 
  left_join(umap_df %>% 
  filter(drug == "DMSO"), .) %>% 
  filter(line != "D020T02") 

gg_factor_umap <- umap_factor %>% 
  dplyr::select(-size_factor) %>%
  pivot_longer(cols = contains("factor"), names_to = "number", values_to = "value") %>%
  ggplot(aes(v1, v2, color = value)) + 
  ggrastr::geom_point_rast(alpha = 0.1, size = 0.35) + 
  scale_color_viridis_c() +
  cowplot::theme_cowplot() +
  labs(x = "UMAP 1",
       y = "UMAP 2",
       color = "factor value") + 
  facet_wrap(~ number) +
  theme(legend.position = "bottom") + 
    coord_fixed()

gg_factor_umap
```



```{r}
gg_f12 <- weights %>% 
  rename(line = id) %>%
  left_join(organoid_size_fit %>% mutate(line = paste0(line, "_", rep))) %>%
  left_join(organoid_morphology %>% 
            mutate(line = substr(line, 1, 4)) %>% 
  expand_grid(., rep = c("r1", "r2")) %>% 
  mutate(sample = paste0(line, "_", rep)) %>% dplyr::select(-line, line = sample)) %>%
  distinct() %>%
  mutate(drug = if_else(is.na(morphology), "treatment", "ctrl")) %>%
  ggplot(aes(factor1, factor2, label = line, color = drug)) + 
  geom_point(size = 4) + 
  ggrepel::geom_text_repel(color = "black") + 
  #scale_color_viridis_c() + 
  theme_cowplot()

gg_f12
```

```{r}
weights %>% 
  rename(line = id) %>%
  left_join(organoid_size_fit %>% mutate(line = paste0(line, "_", rep))) %>%
  left_join(organoid_morphology %>% 
            mutate(line = substr(line, 1, 4)) %>% 
  expand_grid(., rep = c("r1", "r2")) %>% 
  mutate(sample = paste0(line, "_", rep)) %>% dplyr::select(-line, line = sample)) %>%
  distinct() %>%
  mutate(drug = if_else(is.na(morphology), "treatment", "ctrl")) %>% 
  dplyr::select(contains("factor"), line) %>%
  column_to_rownames("line") %>% pheatmap::pheatmap()
```



```{r}
weights %>% 
  rename(line = id) %>%
  left_join(organoid_size_fit %>% mutate(line = paste0(line, "_", rep))) %>%
  ggplot(aes(factor1, factor3, label = line)) + 
  geom_point(size = 2) + 
  ggrepel::geom_text_repel(color = "black") + 
  scale_color_viridis_c() + 
  theme_cowplot()
```

## mutations

```{r}
loadings_mutation %>%
  column_to_rownames("id") %>% 
  pheatmap::pheatmap()
```

```{r}
df <- loadings_mutation %>% 
  gather(key = "factor", value = "weight", -id) 

df %>%
  ggplot(aes(factor, weight, label = id)) +
  geom_point() + 
  theme_cowplot() + 
  ggrepel::geom_text_repel(data = df %>% filter(id %in% c("NRAS", "ERBB2", "PTEN", "KRAS", "PIK3CA")))
```



## drug response


```{r}
## load drug annotation
drug_anno <- readxl::read_excel(here::here('references/layouts/Compound_Annotation_Libraries_New.xlsx')) %>% distinct(drug = `Compound name`, target = `Primary Target`)

loadings_drug_activity_tidy <- loadings_drug_activity %>% 
  mutate(drug = substr(id, 7, nchar(.))) %>% 
  dplyr::select(-id) %>% 
  arrange(desc(factor1)) %>% 
  left_join(drug_anno)


# 
# top_n <- 20
# 
# top_drugs <- rbind(
#   loadings_drug_activity_tidy %>% arrange(factor1) %>% head(top_n),
#   loadings_drug_activity_tidy %>% arrange(factor1) %>% tail(top_n),
#   loadings_drug_activity_tidy %>% arrange(factor2) %>% head(top_n),
#   loadings_drug_activity_tidy %>% arrange(factor2) %>% tail(top_n),
#   loadings_drug_activity_tidy %>% arrange(factor3) %>% head(top_n),
#   loadings_drug_activity_tidy %>% arrange(factor3) %>% tail(top_n)
# ) %>% distinct()
# 
# top_drugs %>% 
#   column_to_rownames("drug") %>% 
#   pheatmap::pheatmap()
```

```{r}
df <- loadings_drug_activity_tidy %>% dplyr::select(-target) %>% 
  gather(factor_name, value, -drug) %>% 
  distinct() 

row_anno <-  df %>% left_join(loadings_drug_activity_tidy) %>% dplyr::select(drug, target) %>%
  mutate(id = paste0(drug, "_", target)) %>% distinct() %>% as.data.frame() %>% column_to_rownames("id") 
           

df %>%
  spread(drug, value) %>% 
  dplyr::select(-factor_name) %>%
  cor() %>% 
  as.matrix() %>% 
  pheatmap::pheatmap(show_rownames = FALSE, 
                     show_colnames = FALSE)
```


