---
title: "Gene Expression Morphology"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE)

```


```{r}
library(SummarizedExperiment)
library(limma)
library(glmnet)
library(pheatmap)
library(reshape2)
library(tidyverse)
library(cowplot)
library(here)
library(DESeq2)


library(clusterProfiler)
library(ReactomePA)
library(enrichplot)
library(org.Hs.eg.db)
```

Define function namespace.

```{r}
filter <- dplyr::filter
select <- dplyr::select
rename <- dplyr::rename
slice <- dplyr::slice
count <- dplyr::count
```

Set parameters for plotting.

```{r}
theme_set(theme_cowplot())
```

# Data loading

We load the gene expression data from an Rdata object from file.

```{r}
load(here('data/processed/expression/promise_expr.rda'))

organoid_morphology <- read_delim(here::here("references/imaging/visual_classification_organoids.csv"), ";", escape_double = FALSE, trim_ws = TRUE) %>% 
  dplyr::select(line = organoid, morphology = visual_inspection_v2) %>%
  mutate(line = substr(line, 1, 4)) %>% 
  mutate(morphology = if_else(is.na(morphology), "other", morphology))

## annotate phenotype group
solid <- organoid_morphology  %>% 
  filter(morphology == "solid") %>%
  .$line
cystic <- organoid_morphology %>% 
  filter(morphology == "cystic") %>%
  .$line

## organoid size
organoid_size_fit <- readRDS(here::here("data/processed/morphology/organoid_size.Rds")) %>% 
  filter(!line %in% c('D055T01', 'D020T02', 'D021T01')) %>% 
  #filter(!line %in% c('D055T01','D020T02')) %>% 
  mutate(line = as.character(line)) %>% 
  dplyr::select(line, size = x, rep = replicate) %>% 
  distinct() %>% arrange(line) %>%
  mutate(line = substr(line, 1, 4)) %>% 
  mutate(rep = paste0("r", rep))

# adding metadata and filtering the promise_expr object
new_col <- colData(promise_expr) %>% as.data.frame() %>%
  left_join(., left_join(
             organoid_size_fit, 
             organoid_morphology)) %>% 
  dplyr::select(-(line:chip_name))
colData(promise_expr) <- cbind(colData(promise_expr), new_col)

promise_expr_filtered <- promise_expr[,promise_expr$line %in% (organoid_size_fit$line %>% unique())]

## long data frame
promise_long <- assays(promise_expr)$expr %>% 
  as_tibble(rownames = 'probe') %>% 
  pivot_longer(values_to = 'expr', names_to = 'id', -probe) %>%
  left_join(as_tibble(rowData(promise_expr), rownames = 'probe')) %>%
  inner_join(as_tibble(colData(promise_expr), rownames = 'id')) %>%
  select(-chip_name)

## exclude outlier
promise_long <- promise_long %>% filter(!line %in% c('D054', 'D055', 'D021'))
#promise_long <- promise_long %>% filter(!line %in% c('D054', 'D055'))

## adding phenotype information
promise_long <- promise_long %>% 
  mutate(phenotype = ifelse(line %in% solid, 'solid', 
                     ifelse(line %in% cystic, 'cystic', 'other'))) %>% 
  filter(phenotype != 'other')
```

# Quality control

We perform a clustering to check if the samples are grouped by line of origin.

```{r}
## make expr matrix
expr_mat <- assays(promise_expr_filtered)$expr

## generate heatmap annotation
anno <- promise_long %>% 
  distinct(line, rep, phenotype) %>% 
  unite(id, line, rep) %>% 
  mutate_all(as.factor) %>%
  as.data.frame() %>% column_to_rownames('id')

## visualize distance matrix
pheatmap(as.matrix(dist(t(expr_mat))),
         annotation_row = anno)

## visualize most variable probes
pheatmap(expr_mat[order(apply(expr_mat, 1, sd), decreasing = T)[1:100],], 
         annotation_col = anno)
```

```{r, eval = FALSE}
boxplot(log(expr)[sample(1:nrow(expr), 2000),])
```


## Feature Selection

```{r}
means <- rowMeans(assays(promise_expr_filtered)$expr)
sd <- apply(assays(promise_expr_filtered)$expr,1,sd)
cv <- sd/means
#par(mar=c(3.5,3.5,1,1),mgp=c(2,0.65,0),cex=0.9)
smoothScatter(log(means),log(cv))
```

```{r}
hist_res <- hist(cv, 100, col="#e7efd8", 
                 freq = FALSE, 
                 main = "Histogram of coefficients of variance", 
                 xlab = "CV")

top_n = cv %>% length()
top_n = top_n * 0.10

keep_probes = cv %>% sort(decreasing = TRUE) %>% .[1:top_n] 


thresh_n = min(keep_probes)

abline(v = thresh_n)
```


Keeping the top 5% probes based on the coefficient of variance. This corresponds to this number of genes:

```{r}
top_n
```


```{r}
promise_expr_filtered = promise_expr_filtered[rowData(promise_expr_filtered) %>% rownames() %in% names(keep_probes),]
```



```{r}
pca_hco <- assays(promise_expr_filtered)$expr %>% t() %>% prcomp(., scale = TRUE)
summary(pca_hco)

#Build colnames 
pca_percent <- paste(colnames(pca_hco$x), " ", "(", 
                     as.character(round(pca_hco$sdev^2 / sum(pca_hco$sdev^2) * 100, 1)),
                     "%", ")", sep="")

#Format PCA
pca_hco$x_anno <- pca_hco$x %>%
  as.data.frame() %>%
  rownames_to_column("id") %>% 
  separate(id, c("organoid", "line", "date", "array1", "array2"), sep="_") %>%
  unite(type, c("array1", "array2")) %>% mutate(type = substr(type, 1, nchar(type)-4))
```

```{r}
summary(pca_hco)$importance %>% as.data.frame() %>% 
  rownames_to_column("metric") %>% gather(PC, value, -metric) %>% 
  filter(metric == "Proportion of Variance") %>% 
  mutate(PC = factor(PC, levels = .$PC)) %>% 
  head(-1) %>% 
  ggplot(aes(PC, value)) + 
  geom_point() + 
  labs(y = "Variance explained")
```

```{r, eval = FALSE}
df = promise_long %>% mutate(symbol = ifelse(probe == '202410_x_at', 'IGF2', symbol)) %>% 
  group_by(probe, line, symbol, phenotype) %>% summarise(expr = mean(expr))

plot_df = pca_hco$rotation %>% as.data.frame() %>% rownames_to_column("id") %>% 
  gather(PC, value, -id) %>% 
  nest(-PC) %>%
  mutate(head = purrr::map(data, ~ head(.x, 10)),
         tail = purrr::map(data, ~ tail(.x, 10)),
         headtail = purrr::map2(head, tail, ~ rbind(.x, .y))) %>% 
  dplyr::select(PC, headtail) %>% 
  unnest() %>% 
  arrange(PC, value) %>% 
  filter(PC %in% c("PC1", "PC2", "PC3")) %>% 
  dplyr::select(-value, probe = id) %>%
  left_join(df)

plot_df %>% 
  ggplot(aes(line, probe, fill = expr)) + 
  geom_tile() + 
  facet_wrap(~ PC)
```


```{r}
pca_morph <- pca_hco$x_anno %>%
  left_join(organoid_morphology %>% rename(organoid = line)) %>%
  ggplot(aes(PC1, PC2, color = morphology, label = organoid)) + 
  geom_point(size = 3) + 
  theme_classic() + 
  xlab(pca_percent[1]) + 
  ylab(pca_percent[2]) +
  scale_color_manual(values = c("grey", "black")) +
  theme(legend.position = "bottom") + 
  ggrepel::geom_text_repel() +
  ggtitle("PCA on probe level - organoid morphology")

pca_morph
```

```{r}


pca_size <- pca_hco$x_anno %>%
  left_join(organoid_size_fit %>% rename(organoid = line)) %>%
  ggplot(aes(PC1, PC2, color = size)) + 
  geom_point(size = 3) + 
  theme_classic() + 
  xlab(pca_percent[1]) + 
  ylab(pca_percent[2]) +
  theme(legend.position = "bottom") + 
  #scale_color_viridis_c() +
  scale_color_gradient(low="blue", high="red") +
  #scale_colour_gradient2() +
  ggtitle("PCA on probe level - organoid size")

pca_size
```

```{r}
df = promise_long %>% 
  mutate(symbol = ifelse(probe == '202410_x_at', 'IGF2', symbol)) %>%
  filter(symbol %in% c("IGF2")) %>% 
  dplyr::select(organoid =line, expr, line = rep)

pca_igf <- pca_hco$x_anno %>%
  left_join(df) %>%
  ggplot(aes(PC1, PC2, color = expr)) + 
  geom_point(size = 3) + 
  theme_classic() + 
  xlab(pca_percent[1]) + 
  ylab(pca_percent[2]) +
  theme(legend.position = "bottom") + 
  scale_color_viridis_c() +
  #scale_color_gradient(low="blue", high="red") +
  #scale_colour_gradient2() +
  ggtitle("PCA on probe level - IGF2 expression")

pca_igf
```



```{r}
promise_expr_filtered

size_mat <- model.matrix(~size, colData(promise_expr_filtered))

size_fit <- limma::lmFit(assays(promise_expr_filtered)$expr, design = size_mat)
size_fit_bayes <- limma::eBayes(size_fit)

toptable_size = topTable(size_fit_bayes, number = Inf, coef = "size") %>% rownames_to_column("id") %>% 
  left_join(rowData(promise_expr_filtered) %>% as.data.frame() %>% rownames_to_column("id"))

toptable_size %>% ggplot(aes(P.Value)) + 
  geom_histogram(bins = 100) + 
  theme_classic() + 
  geom_vline(xintercept = 0.05) + 
  xlab("p-value")  +
  ggtitle("p-value distriutions") 
  #scale_y_log10()
```

```{r}
toptable_size %>% 
  mutate(symbol = ifelse(id == '202410_x_at', 'IGF2', symbol)) %>%
  filter(adj.P.Val <= 0.05) %>%
  arrange(desc(t))
```


```{r}
promise_long %>% 
  mutate(symbol = ifelse(probe == '202410_x_at', 'IGF2', symbol)) %>%
  filter(symbol %in% c("H19", "MIR483", "IRS1", "IGF2")) %>% arrange(expr) %>% arrange(line) %>% 
  group_by(line, rep, symbol) %>% 
  summarise(expr = mean(expr)) %>% 
  left_join(organoid_size_fit) %>% 
  arrange(symbol) %>%
  ggplot(aes(expr, size)) + 
  geom_point() + 
  geom_smooth(method = "lm") + 
  facet_wrap(~ symbol, ncol =2)
```


```{r, eval = FALSE}
## gene ranks for enrichment, best T-value for each gene
ranks <- top3_res %>%
  filter(!tp53, !sex) %>%
  arrange(desc(abs(t)))
ranks <- setNames(ranks$t, ranks$entrez)

## get pathway annotation
pathways <- fgsea::reactomePathways(names(ranks))

## perform GSEA
enr <- fgsea::fgsea(pathways, ranks, nperm = 1e5, maxSize=500)
enr <- as_tibble(enr) %>% arrange(pval)

## genes for interesting terms
eg <- enr %>% filter(pathway == 'IGF1R signaling cascade') %>% pull(leadingEdge) %>% unlist()


## a function to draw barcode plots
custom_barcode_plot <- function(stat_vector, sig_genes, term){
  ## generate barcode plot
  bc_plot <- plotEnrichment(sig_genes, stat_vector)
  
  ## remove unwanted layers
  bc_plot$layers <- list()
  
  ## add barcode at the bottom
  lowest_pos <- min(bc_plot$data[,2])
  dash_length <- abs(purrr::reduce(range(bc_plot$data[,2]), `-`)*0.1)
  middle <- which.min(abs(sort(stat_vector, decreasing=T)))
  
  bc_plot_custom <- bc_plot + geom_segment(aes(x=x, xend=x), y=lowest_pos,
                           yend=lowest_pos-dash_length) + 
    geom_line(colour='#4daf4a') + 
    geom_hline(yintercept=lowest_pos, colour='#cccccc') + 
    geom_hline(yintercept=0, colour='#cccccc') + xlab('') +
    theme_classic() +
    geom_tile(data=tibble(rank=1:length(stat_vector), 
                          y=lowest_pos-(1.25*dash_length)), 
              aes(x=rank, y=y, fill=rank),
                  width=1,
                  height=0.5*dash_length) +
    scale_fill_gradient2(low ='#b2182b', high='#2166ac', 
                         mid='#f7f7f7', midpoint = middle) + 
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    theme(panel.grid=element_blank(), 
          axis.text.x=element_blank(),
          axis.ticks.x = element_blank(),
          legend.position = 'none') + 
    ggtitle(term) +
    ylab('Enrichment score')
  
  return(bc_plot_custom)
}

signature_genes <- pathways['IGF1R signaling cascade'] %>% unlist()

custom_barcode_plot(ranks, signature_genes,
                    'IGF1R signaling')
```

# Organoid morphology

```{r}
promise_expr_filtered

morph_mat <- model.matrix(~morphology, colData(promise_expr_filtered))

morph_fit <- limma::lmFit(assays(promise_expr_filtered)$expr, design = morph_mat)
morph_fit_bayes <- limma::eBayes(morph_fit)

toptable_morph <- topTable(morph_fit_bayes, number = Inf, coef = "morphologysolid") %>% rownames_to_column("id") %>% 
  left_join(rowData(promise_expr_filtered) %>% as.data.frame() %>% rownames_to_column("id"))

toptable_morph %>% ggplot(aes(P.Value)) + 
  geom_histogram(bins = 100) + 
  theme_classic() + 
  geom_vline(xintercept = 0.05) + 
  xlab("p-value")  +
  ggtitle("p-value distriutions") 
  #scale_y_log10()
```


```{r}
toptable_morph %>% 
  filter(adj.P.Val <= 0.05) %>%
  arrange(desc(t))
```

```{r}
promise_long %>% 
  filter(symbol %in% c("DACH1", "SERPINA1", "DKK1", "IFITM1")) %>% arrange(expr) %>% arrange(line) %>% 
  group_by(line, rep, symbol, phenotype) %>% 
  summarise(expr = mean(expr)) %>% 
  left_join(organoid_size_fit) %>% 
  ggplot(aes(phenotype, expr)) + 
  ggrastr::geom_boxplot_jitter(width = .2) + 
  geom_jitter(width = .1) +
  #geom_boxplot() +
  #geom_smooth(method = "lm") + 
  facet_wrap(~ symbol, ncol =2)
```

```{r}
# I only feed the top 5% most variable genes into the GO enrichment
df = toptable_morph %>% drop_na() %>% 
  group_by(entrez) %>% 
  summarise(t = mean(t)) %>%
  arrange(desc(t))
ranks = setNames(df$t, as.character(df$entrez))
#ranks = setNames(as.character(df$t), df$entrez)


## GO term enrichment (biological process)
gse_go <- gseGO(
  geneList = ranks,
  OrgDb = org.Hs.eg.db,
  ont = 'BP',
  nPerm = 1e5,
  minGSSize = 10,
  maxGSSize = 500,
  pvalueCutoff = 0.2
)

go <- pairwise_termsim(gse_go) 
emapplot(go)
```

```{r, eval = FALSE}
## Reactome enrichment analysis
gse_reactome <- gsePathway(
  geneList = ranks,
  organism = 'human',
  nPerm = 1e5,
  minGSSize = 10,
  maxGSSize = 500,
  pvalueCutoff = 0.2
)

reactome <- pairwise_termsim(gse_reactome) 
emapplot(reactome)
```


```{r}
## parse signatures and make list
intestinal_sig <- readxl::read_excel(here('data/external/expression/merloz-suarez_sigantures.xls'), 
                                     sheet = 1, skip = 4) %>% .[,1:4] %>%
  gather(signature, symbol) %>% drop_na() %>% 
  mutate(symbol = gsub('\\*', '', symbol))

## run gsea with clusterprofiler
df = toptable_morph %>% drop_na() %>% 
  group_by(symbol) %>% 
  summarise(t = mean(t)) %>%
  arrange(desc(t))
ranks_symbol = setNames(df$t, as.character(df$symbol))

gse_sig <- GSEA(
  geneList = ranks_symbol,
  TERM2GENE = intestinal_sig,
  nPerm = 1e5,
  minGSSize = 1,
  maxGSSize = 1000,
  pvalueCutoff = 1
)

## output as tibble
gse_sig_tbl <- as_tibble(gse_sig)
```


```{r}
## lgr5 signature
gseaplot2(
  gse_sig, geneSetID = gse_sig$ID[2], 
  title = paste0(gse_sig$ID[2], 
                ' (p = ', round(gse_sig_tbl$pvalue[2], 3), 
                '; NES = ', round(gse_sig_tbl$NES[2], 2), ')')
) 
```


```{r}
## proliferation
gseaplot2(
  gse_sig, geneSetID = gse_sig$ID[1],
  title = paste0(gse_sig$ID[1], 
                ' (p = ', round(gse_sig_tbl$pvalue[1], 3), 
                '; NES = ', round(gse_sig_tbl$NES[1], 2), ')')
)
```

# MOFA

```{r}
library(MOFA2)

input_df = promise_long %>% 
  # renaming columns
  dplyr::select(sample = line,
                feature = probe,
                value = expr) %>% 
  mutate(view = "expression") %>% 
  # averaging feature value
  group_by(feature, sample, view) %>%
  summarise(value = mean(value)) %>%
  # renaming feature jic
  mutate(feature = paste0(feature, "_", view)) %>%
  dplyr::select(sample, feature, view, value) %>% 
  drop_na() %>%
  data.table::as.data.table()
```

```{r}
file <- system.file("extdata", "test_data.RData", package = "MOFA2")
load(file) 
dt[,group:=NULL]

dt$value <- input_df$value[1:nrow(dt)]
dt$feature <- input_df$feature[1:nrow(dt)]

MOFAmodel <- create_mofa_from_df(dt)
plot_data_overview(MOFAmodel)
```


```{r}
MOFAobject <- create_mofa(input_df)
print("created MOFA object")
plot_data_overview(MOFAobject)
```

```{r}
run_mofa(MOFAmodel, outfile, use_basilisk = TRUE)
```


```{r}
# MOFAobject <- prepare_mofa(
#   object = MOFAobject,
#   data_options = data_opts,
#   model_options = model_opts,
#   training_options = train_opts
# )

outfile = file.path(getwd(),"model.hdf5")
MOFAobject.trained <- run_mofa(MOFAobject, outfile, use_basilisk = TRUE)
```


# Organoid size and morphology


```{r, eval = FALSE}
## model matrix
mm <- model.matrix(~phenotype, data = sample_anno)

## set technical replicate plots
blocks <- as.factor(sample_anno$line)
## compute within-block correlations
corfit <- duplicateCorrelation(expr_mat, mm, block = blocks)

## fit model and compute model coefficients
fit_phenotype <- lmFit(expr_mat, mm, 
                       block = blocks, 
                       cor= corfit$consensus.correlation)
fit_phenotype <- eBayes(fit_phenotype)

## extract results and annotate gene symbols
probe_to_symbol <- promise_expr %>% distinct(probe, symbol) %>% 
  group_by(probe) %>% slice(1) %>% ungroup()
phenotype_res <- topTable(fit_phenotype, coef = 'phenotypesolid', n = Inf) %>% 
  as_tibble(rownames = 'probe') %>%
  left_join(probe_to_symbol)
```

## other

```{r, eval=FALSE}
volc_data <- top3_res %>% 
  mutate(group = ifelse(symbol %in% c('IGF2', 'MIR483', 'H19', 'INS-IGF2'), 'igf2', 'none')) %>%
  mutate(label = ifelse((group != 'none') & (adj.P.Val < 0.01), symbol, ''))

co05 <- volc_data %>% filter(adj.P.Val > 0.05) %>% pull(P.Value) %>%
  head(1) %>% log10() %>% `*`(-1)

ggplot() +
  ggrastr::geom_point_rast(data=subset(volc_data, group == 'none'),
             aes(logFC, -log10(P.Value)), colour='#dddddd') +
  geom_point(data=subset(volc_data, group == 'igf2'),
             aes(logFC, -log10(P.Value)), colour='#4285f4') +
  geom_hline(yintercept = co05, linetype='dashed') +
  ggrepel::geom_text_repel(data=subset(volc_data, group == 'igf2'),
                  aes(logFC, -log10(P.Value), label=label)) +
  ylab('P-value [-log10]') + xlab('Fold change [log2]') + 
  scale_y_continuous(expand=c(0,0))
```



# Organoid size and gene expression

```{r}
sessionInfo()
```

