---
title: "Create overview plots for compounds of interest"
output: html_notebook
---

```{r, message=FALSE}
library(tidyverse)
library(here)
library(EBImage)
library(janitor)
# library(imager)

knitr::opts_chunk$set(cache = TRUE,
                      echo = FALSE)


```

I start loading the annotation file for the library I am interested in. We are interested in both the CCP and the KiStem library information. 

```{r}
ccp <- read_delim(here::here("references/Clin_Cancer_Panel_V170511.csv"), 
    ";", escape_double = FALSE, trim_ws = TRUE) %>% .[,-1] %>% 
  janitor::clean_names()

kistem <- read_delim(here("references/KiStem_edt_V170125.csv"), 
    ";", escape_double = FALSE, trim_ws = TRUE) %>% .[,-1] %>% 
  janitor::clean_names()

fda <- read_csv(here("references/FDA_1443_edt_V170125.csv")) %>% .[,-1] %>% 
  janitor::clean_names()
```

I define the source image path. 

```{r}
#path_to_images_dir <- here("html") #only use for local analysis
path_to_images_dir <- here("data/profiling/htmldir")
list.files(path_to_images_dir)
```

I define three function that take an annotation file, a file path and a compound of interest. They produce figures of every available image and their respective activity.

```{r}
get_img_path_treatment <- function(path_to_images, treatment_of_interest, anno){
list.files(path_to_images) %>% 
  tibble(file = ., path = paste0(path_to_images, "/", file)) %>% 
    # removing files that I do not want to work with
  filter(! grepl(pattern = ".md5s", x = file)) %>% 
  filter(! grepl(pattern = ".html", x = file)) %>% 
  filter(! grepl(pattern = "copylog", x = file)) %>% 
  filter(! grepl(pattern = "D0", x = file)) %>% 
  mutate(library_id = substr(file, 13, 14) %>% as.numeric()) %>% 
  left_join(anno, by = "library_id") %>%
  drop_na() %>% 
  filter(product_name %in% treatment_of_interest) %>% 
  mutate(path_img = paste0(path, "/", file, "_", row_id_384, "_", 
                           stringr::str_pad(col_id_384, width = 2, side = "left", pad = "0"),
                           "_1.jpeg")) %>% 
  dplyr::select(file, path, path_img, product_name, concentration) %>% 
  mutate(line = substr(file, 1, 7)) %>%
  return()
}


get_img_path_well <- function(path_to_images, library, col_id, row_id, anno){
list.files(path_to_images) %>% 
  tibble(file = ., path = paste0(path_to_images, "/", file)) %>% 
    # removing files that I do not want to work with
  filter(! grepl(pattern = ".md5s", x = file)) %>% 
  filter(! grepl(pattern = ".html", x = file)) %>% 
  filter(! grepl(pattern = "copylog", x = file)) %>% 
  filter(! grepl(pattern = "D0", x = file)) %>% 
  mutate(library_id = substr(file, 13, 14) %>% as.numeric()) %>% 
  left_join(anno, by = "library_id") %>%
  drop_na() %>% 
  filter(col_id_384 == col_id,
         row_id_384 == row_id,
         library_id == library) %>% 
  mutate(path_img = paste0(path, "/", file, "_", row_id_384, "_", 
                           stringr::str_pad(col_id_384, width = 2, side = "left", pad = "0"),
                           "_1.jpeg")) %>% 
  dplyr::select(file, path, path_img, product_name, concentration) %>% 
  mutate(line = substr(file, 1, 7)) %>%
  return()
}

get_img_data <- function(path){
set.seed(312)

img_df <- path %>%
  group_by(concentration, product_name, line) %>% 
  sample_n(1) %>%
  filter(!(product_name == "DMSO" & concentration != 1)) %>%
  mutate(img_raw = purrr::map(path_img, ~ imager::load.image(.x))) %>%
  mutate(img = purrr::map(img_raw, ~ .x %>% as.data.frame(wide = "c") %>% 
           mutate(rgb.val = rgb(c.1, c.2, c.3))))

return(img_df)
}

plot_img <- function(data, treatment_of_interest){
img_df_plot <- data %>% 
  unnest(img) 

img_df_plot %>%  
  ggplot(aes(x,y)) + 
  geom_raster(aes(fill = rgb.val)) + 
  scale_fill_identity() + 
  scale_y_reverse() + 
  facet_grid(line ~ product_name + concentration) + 
  theme_bw() + 
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) + 
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  theme(panel.border = element_blank(), panel.grid.major = element_blank(),
panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) +
  ggsave(file = paste0(treatment_of_interest[1], "_mco.pdf"), 
         height = 3*(img_df_plot$line %>% unique() %>% length()),
         width = 3*(img_df_plot %>% ungroup %>% dplyr::select(product_name, concentration) %>% 
           distinct() %>% nrow()))
}
```

I define my annotation data. 

```{r}
anno_df <- rbind(ccp %>% dplyr::select(library_id, product_name, col_id_384, 
                                    row_id_384, concentration), 
              kistem %>% dplyr::select(library_id, product_name, 
                                       col_id_384, row_id_384) %>% 
                mutate(concentration = 1),
              fda %>% dplyr::select(product_name, 
                                       col_id_384, row_id_384,
                                    robot_destination) %>% 
                mutate(concentration = 1,
                       library_id = as.numeric(robot_destination)+3) %>% 
                dplyr::select(-robot_destination))

write_rds(anno_df, here::here("data/compound_libraries/annotation.Rds"))
```

Now I execute and create tables for drugs I am interested in. 
To improve speed, I execute my code in parallel. 

```{r, eval = FALSE}
tibble(drug = c()) %>% 
  
get_img_path_treatment(path_to_images = path_to_images_dir, 
             treatment_of_interest = name,
             anno = anno_df) %>% 
  get_img_data() %>% 
  plot_img(treatment_of_interest = name)
```

I hack away 

```{r}
## new figure
name <- c("WYE-125132 (WYE-132)", "DMSO")
set.seed(312)

get_img_path_treatment(path_to_images = path_to_images_dir, 
             treatment_of_interest = name,
             anno = anno_df) %>% 
  get_img_data()  %>% 
  plot_img(treatment_of_interest = name)


## new figure
name <- c("Irinotecan", "DMSO")
set.seed(312)

get_img_path_treatment(path_to_images = path_to_images_dir, 
             treatment_of_interest = name,
             anno = anno_df) %>% 
  get_img_data()  %>% 
  plot_img(treatment_of_interest = name)
```
