---
title: "Cell Line Differences (Human)"
author: "Jan Sauer"
# date: "10.11.2018"
output: 
   BiocStyle::html_document:
    code_folding: hide
    df_print: paged
    toc_depth: 4
    # toc_float: true
    # fig_height: 6
    # fig_width: 9.5
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message=FALSE}
library(h5)
library(MASS)
library(dplyr)
library(ggplot2)
library(plotly)
library(Rtsne)
library(pheatmap)
library(data.table)
library(knitr)
library(RColorBrewer)
source("../r_functions/PublicationTheme_ggplot.R")
source("../r_functions/feature_class.R")
source("../r_functions/Color_Palette.R")
source("../r_functions/Replicates.R")

knitr::opts_chunk$set(cache = TRUE, warning = FALSE, 
                      message = FALSE, cache.lazy = FALSE)

htmldir = "/collab-ag-fischer/PROMISE/data-10x-4t-c-16z/htmldir"

species = "human"
feature_type = "organoids"

lines = sort(get_lines(species))
colorScale = setNames(
  object = get_color_palette(species, length(lines)), 
  nm = lines)
```

# Introduction
I reduce the features and select the most informative features using a PCA, i.e. I retain the features that contribute the most to the variance as measured by their relative contribution to each PC component.

# Line Differences of DMSO Organoids
## Load Data
```{r c1}
dmso_dir = sprintf("%s/DMSO/results", species)
f = h5file(file.path(dmso_dir, sprintf("ReducedFeatures_%s_DMSO.h5", species)), "r")
features = data.frame(t(f[sprintf("features_%s", feature_type)][]))
colnames(features) = f[sprintf("feature_names_%s", feature_type)][]
metadata = data.frame(f[sprintf("metadata_%s", feature_type)][], stringsAsFactors = FALSE)
colnames(metadata) = f[sprintf("metadata_names_%s", feature_type)][]
h5close(f)
```

## Organoid Features
The features are sampled to ensure a balance between classes.
```{r c2}
sample_size = min(table(metadata$Line))
features_organoids = features %>% 
  group_by("Line" = metadata$Line) %>%
  sample_n(sample_size) %>% as.data.frame
```

### LDA
I want to test several feature set sizes to compare them visually
```{r c3}
expl_var_plots = list()
lda_proj_plots = list()
num_features_list = c(5, 10, 25, 50, 100, 200)
num_features_list = num_features_list[num_features_list < ncol(features)]
num_features_list = c(num_features_list, ncol(features))

for(num_features in num_features_list) {
  model = lda(
    x=features_organoids %>% select(-Line) %>%
      select(seq_len(num_features)), 
    grouping=features_organoids$Line)
  plda = predict(object = model, newdata = features_organoids %>% select(-Line) %>%
      select(seq_len(num_features)))
  transformed = as.data.frame(plda$x)
  transformed$Line = features_organoids$Line
  
  # Accuracy
  acc = mean(as.character(plda$class) == features_organoids$Line)
  print(sprintf("Accuracy for %s features = %s", num_features, acc))
  
  # Explained Variance
  expl_var_plots[[length(expl_var_plots) + 1]] = ggplot_df = data.frame(
    "Component" = paste0("LD", seq_along(model$svd)), 
    "Value" = model$svd**2 / sum(model$svd**2), 
    "NumFeatures" = num_features)
  # ggplot(data = ggplot_df) + 
  #   geom_col(aes(x = Component, y = Value)) + 
  #   xlab("") + ylab("") + ggtitle("Proportion of Variance") + 
  #   theme_Publication() + scale_fill_Publication() + 
  #   theme(legend.position = "none")
  
  # 2D LDA Projection
  transformed$NumFeatures = num_features
  lda_proj_plots[[length(lda_proj_plots) + 1]] = transformed[, c("LD1", "LD2", "Line", "NumFeatures")]
  # ggplot(data = transformed) +
  #   geom_point(aes(x = LD1, y = LD2, color = Line)) +
  #   theme_Publication() + scale_fill_Publication() +
  #   theme(legend.position = "right", legend.direction = "vertical",
  #         legend.key.size = unit(0.5, "cm"))
}

lda_proj_plots = do.call(rbind, lda_proj_plots)
expl_var_plots = do.call(rbind, expl_var_plots)

ggplot(data = lda_proj_plots) + 
  geom_point(aes(x = LD1, y = LD2, color = Line)) +
  facet_wrap(facets =  ~ NumFeatures) + 
  theme_Publication() + scale_color_manual(values = colorScale) +
  theme(legend.position = "right", legend.direction = "vertical",
        legend.key.size = unit(0.5, "cm"))
```



## Well Features

A better separation becomes visible if features are averaged across wells

```{r c4}
well_id = paste0(metadata$Plate, "_", metadata$Well)
features_wells = aggregate(x = features, by = list(well_id), FUN = median)
rownames(features_wells) = features_wells$Group.1
features_wells$Line = substr(features_wells$Group.1, 1, 7)
features_wells$Group.1 = NULL

metadata_wells = aggregate(x = metadata, by = list(well_id), FUN = first)
rownames(metadata_wells) = metadata_wells$Group.1
metadata_wells$Group.1 = NULL
```

### LDA
```{r c5}
expl_var_plots = list()
lda_proj_plots = list()
contribs = list()
num_features_list = c(5, 10, 25, 50, 100, 200)
num_features_list = num_features_list[num_features_list < ncol(features)]
num_features_list = c(num_features_list, ncol(features))

for(num_features in num_features_list) {
  model = lda(
    x=features_wells %>% select(-Line) %>%
      select(seq_len(num_features)), 
    grouping=features_wells$Line)
  plda = predict(object = model, newdata = features_wells %>% select(-Line) %>%
      select(seq_len(num_features)))
  transformed = as.data.frame(plda$x)
  transformed$Line = features_wells$Line
  
  # Accuracy
  acc = mean(as.character(plda$class) == features_wells$Line)
  print(sprintf("Accuracy for %s features = %s", num_features, acc))
  
  # Explained Variance
  expl_var_plots[[length(expl_var_plots) + 1]] = data.frame(
    "Component" = paste0("LD", seq_along(model$svd)), 
    "Value" = model$svd**2 / sum(model$svd**2), 
    "NumFeatures" = num_features)
  # ggplot(data = ggplot_df) + 
  #   geom_col(aes(x = Component, y = Value)) + 
  #   xlab("") + ylab("") + ggtitle("Proportion of Variance") + 
  #   theme_Publication() + scale_fill_Publication() + 
  #   theme(legend.position = "none")
  
  # 2D LDA Projection
  transformed$NumFeatures = num_features
  lda_proj_plots[[length(lda_proj_plots) + 1]] = transformed[
    , c("LD1", "LD2", "LD3", "Line", "NumFeatures")]
  # ggplot(data = transformed) +
  #   geom_point(aes(x = LD1, y = LD2, color = Line)) +
  #   theme_Publication() + scale_fill_Publication() +
  #   theme(legend.position = "right", legend.direction = "vertical",
  #         legend.key.size = unit(0.5, "cm"))
  
  # Feature contributions to LD1 and LD2
  contrib = list()
  for(ldc in colnames(model$scaling)) {
    contrib[[ldc]] = model$scaling[,ldc]
    contrib[[ldc]] = contrib[[ldc]][order(abs(contrib[[ldc]]), decreasing = TRUE)]
    contrib[[ldc]] = data.frame(
      "var" = names(contrib[[ldc]]), 
      "value" = contrib[[ldc]], 
      "LD" = ldc,
      "NumFeatures" = num_features)
  }
  contribs[[length(contribs) + 1]] = do.call(rbind, contrib)
}

lda_proj_plots = do.call(rbind, lda_proj_plots)
expl_var_plots = do.call(rbind, expl_var_plots)
contribs = do.call(rbind, contribs)

ggplot(data = lda_proj_plots) + 
  geom_point(aes(x = LD1, y = LD2, color = Line)) +
  facet_wrap(facets =  ~ NumFeatures) + 
  theme_Publication() + scale_color_manual(values = colorScale) +
  theme(legend.position = "right", legend.direction = "vertical",
        legend.key.size = unit(0.5, "cm"))

ggplot(data = lda_proj_plots) + 
  geom_point(aes(x = LD2, y = LD3, color = Line)) +
  facet_wrap(facets =  ~ NumFeatures) + 
  theme_Publication() + scale_color_manual(values = colorScale) +
  theme(legend.position = "right", legend.direction = "vertical",
        legend.key.size = unit(0.5, "cm"))
```

### PCA
```{r c6}
expl_var_plots = list()
pca_plots = list()
contribs = list()
num_features_list = c(5, 10, 25, 50, 100, 200)
num_features_list = num_features_list[num_features_list < ncol(features)]
num_features_list = c(num_features_list, ncol(features))

for(num_features in num_features_list) {
  model = prcomp(
    x = features_wells %>% select(-Line) %>%
      select(seq_len(num_features)),
    scale. = TRUE, center = TRUE)
  transformed = as.data.frame(model$x)
  transformed$Line = features_wells$Line
  
  # Explained Variance
  expl_var_plots[[length(expl_var_plots) + 1]] = data.frame(
    "Component" = paste0("PC", seq_along(model$sdev)), 
    "Value" = model$sdev / sum(model$sdev), 
    "NumFeatures" = num_features)
  
  # 2D LDA Projection
  transformed$NumFeatures = num_features
  pca_plots[[length(pca_plots) + 1]] = transformed[, c("PC1", "PC2", "Line", "NumFeatures")]
}

pca_plots = do.call(rbind, pca_plots)
expl_var_plots = do.call(rbind, expl_var_plots)

ggplot(data = pca_plots) + 
  geom_point(aes(x = PC1, y = PC2, color = Line)) +
  facet_wrap(facets =  ~ NumFeatures) + 
  theme_Publication() + scale_color_manual(values = colorScale) +
  theme(legend.position = "right", legend.direction = "vertical",
        legend.key.size = unit(0.5, "cm"))
```

### t-SNE
```{r c7}
tsne = Rtsne(features_wells %>% select(-Line), perplexity = 50)
ggplot_df = data.frame(
  "Line" = features_wells$Line,
  "X1" = tsne$Y[,1],
  "X2" = tsne$Y[,2])
ggplot(data = ggplot_df) + geom_point(aes(x = X1, y = X2, color = Line)) +
  theme_Publication() + xlab("X") + ylab("Y") +
  theme(legend.position = "right", legend.direction = "vertical",
        legend.key.size = unit(0.5, "cm")) + 
  scale_color_manual(values = colorScale)
```

## Line Clusters

```{r, eval = FALSE}
library(tidyverse)
#I load the prepared well feature data directly and skip most parts
features_wells <- readRDS("~/promise/local_data/image_analysis/WellFeatures_human.rds") %>%
  rownames_to_column("Line") %>%
  mutate(Line = substr(Line, 1, 7))

metadata_wells <- readRDS("~/promise/local_data/image_analysis/WellMetadata_human.rds")
```


I perform a clustering on the lines
```{r c8}

# model = lda(
#   x=features_wells %>% select(-Line), 
#   grouping=features_wells$Line)
# plda = predict(
#   object = model, 
#   newdata = features_wells %>% select(-Line))
# transformed = as.data.frame(plda$x)
# 
# # Explained Variance
# expl_var_plots = data.frame(
#   "Component" = paste0("LD", seq_along(model$svd)), 
#   "Value" = model$svd**2 / sum(model$svd**2),
#   "NumFeatures" = num_features)

model = prcomp(
  x = features_wells %>% select(-Line),
  scale. = TRUE, center = TRUE)
transformed = as.data.frame(model$x)
transformed$Line = features_wells$Line

features_lines = aggregate(
  x = transformed %>% select(-Line), 
  by = list(features_wells$Line), 
  FUN = median)
rownames(features_lines) = features_lines$Group.1
features_lines$Group.1 = NULL

rep_id = paste0(
  metadata_wells$Line, "_", 
  substr(metadata_wells$Plate, 12, 14), "_", 
  metadata_wells$Replicate)

features_replicates = aggregate(
  x = transformed %>% select(-Line),
  by = list("Replicate.ID" = rep_id),
  FUN = median)
rownames(features_replicates) = features_replicates$Replicate.ID
features_replicates$Replicate.ID = NULL

d = dist(features_lines[-7,])
hc = hclust(d, method = "ward.D2")

annotation = data.frame(
  "Line" = substr(rownames(features_replicates[-7,]), 1, 7) %>% unique(),
  row.names = substr(rownames(features_replicates[-7,]), 1, 7) %>% unique(),
  stringsAsFactors = FALSE)

pheatmap(
  mat = t(as.matrix(features_lines[-7, 1:9])), cluster_rows = FALSE, 
  cluster_cols = hc, cutree_cols = 6, annotation_col = annotation, border_color = NA)
```

```{r}
summary(model)$importance[2,c(1:9)] %>% as.data.frame() %>% rownames_to_column("PC") %>% gather(pc, proportion, -PC) %>% dplyr::select(-pc) %>% as.tibble() %>% ggplot(aes(PC, proportion)) + geom_point() + theme_classic()
```

Now I wonder what is the optimal number of clusters on this limited dataset

```{r}
library(fpc)
library(gclus)


set.seed(13)
df <- t(as.matrix(features_lines[,20:ncol(features_lines)])) %>% as.data.frame()

pamk.best <- pamk(df)
cat("number of clusters estimated by optimum average silhouette width:", pamk.best$nc, "\n")
## number of clusters estimated by optimum average silhouette width: 3
plot(pam(df, pamk.best$nc))
```


```{r}
library(cluster)
k <- clusGap(df, FUN = kmeans,  nstart = 2, K.max = 15)
k %>% plot()
```

```{r}
pheatmap(
  mat =scale(df), cluster_rows = FALSE, 
  cluster_cols = hc, cutree_cols = 6, annotation_col = annotation, border_color = NA)
```


### Quality Control
I repeat the clustering but for individual replicates of lines to test the screen quality.

```{r c8b}
rep_id = paste0(
  metadata_wells$Line, "_", 
  substr(metadata_wells$Plate, 12, 14), "_", 
  metadata_wells$Replicate)

annotation = data.frame(
  "Line" = substr(rownames(features_replicates), 1, 7),
  row.names = rownames(features_replicates, 1, 7),
  stringsAsFactors = FALSE)

anno_colorScale = list(
  "Line" = setNames(
    object = colorScale, 
    nm = sort(unique(annotation$Line))))

d = dist(features_replicates)
hc = hclust(d, method = "ward.D2")

pheatmap(
  mat = t(as.matrix(features_replicates[, 1:5])), 
  cluster_rows = FALSE, cluster_cols = hc, annotation_col = annotation, 
  #annotation_colors = anno_colorScale, 
  show_colnames = FALSE, 
  border_color = NA)

ph = pheatmap(
  mat = d, cluster_rows = hc, cluster_cols = hc, 
  annotation_col = annotation, # annotation_row = annotation,
  annotation_colors = anno_colorScale, 
  annotation_names_col = FALSE, annotation_names_row = FALSE, 
  border_color = NA)
```

### Example Wells

```{r c9}
set.seed(100)
num_imgs = 3
metadata_wells = aggregate(x = metadata, by = list(well_id), FUN = first)
random_wells = metadata_wells %>% dplyr::group_by(Line) %>% sample_n(num_imgs)
all_imgs = list()
captions = c()
for(line in sort(unique(random_wells$Line))) {
  subdat = random_wells[random_wells$Line == line, ]
  imgs = list()
  plates = c()
  for(ii in seq_len(num_imgs)) {
    plate = subdat[ii, "Plate"]
    plates = c(plates, as.character(plate))
    row = substr(subdat[ii, "Well"], 1, 1)
    col = substr(subdat[ii, "Well"], 2, 3)
    img_url = file.path(htmldir, plate, sprintf("%s_%s_%s_1.jpeg", plate, row, col))
    imgs[[ii]] = EBImage::readImage(img_url)
  }
  all_imgs[[length(all_imgs)+1]] = EBImage::abind(imgs, along = 1)
  captions = c(captions, paste0(plates, collapse = ", "))
}
```

```{r c10, fig.cap=captions[1], fig.width=10}
EBImage::display(all_imgs[[1]])
```

```{r c11, fig.cap=captions[2], fig.width=10}
EBImage::display(all_imgs[[2]])
```

```{r c12, fig.cap=captions[3], fig.width=10}
EBImage::display(all_imgs[[3]])
```

```{r c13, fig.cap=captions[4], fig.width=10}
EBImage::display(all_imgs[[4]])
```

```{r c14, fig.cap=captions[5], fig.width=10}
EBImage::display(all_imgs[[5]])
```

```{r c15, fig.cap=captions[6], fig.width=10}
EBImage::display(all_imgs[[6]])
```

```{r c16, fig.cap=captions[7], fig.width=10}
EBImage::display(all_imgs[[7]])
```

```{r c17, fig.cap=captions[8], fig.width=10}
EBImage::display(all_imgs[[8]])
```

```{r c18, fig.cap=captions[9], fig.width=10}
EBImage::display(all_imgs[[9]])
```

```{r c19, fig.cap=captions[10], fig.width=10}
EBImage::display(all_imgs[[10]])
```

```{r c20, fig.cap=captions[11], fig.width=10}
EBImage::display(all_imgs[[11]])
```

```{r c21, fig.cap=captions[12], fig.width=10}
EBImage::display(all_imgs[[12]])
```

```{r c22, fig.cap=captions[13], fig.width=10}
EBImage::display(all_imgs[[13]])
```

```{r c23, fig.cap=captions[14], fig.width=10}
EBImage::display(all_imgs[[14]])
```

```{r c24, fig.cap=captions[15], fig.width=10}
EBImage::display(all_imgs[[15]])
```

## Individual Features

The features are sorted by importance, i.e. how well they distinguish organoid lines. I look at the first few features to see how lines cluster along them.

### Organoid Features

```{r c25, fig.height=12}
ggplot_df = melt(features_wells %>% select(1:12, Line), id.vars = "Line")
ggplot(data = ggplot_df) +
  geom_histogram(mapping = aes(value, fill = Line), 
                 position = "identity", alpha = 0.5, bins = 50) + 
  facet_wrap(facets = ~variable, nrow = 4) + 
  theme_Publication() + scale_fill_manual(values = colorScale) + 
  labs(title = "Histogram of Organoid Area", x = "", y = "")
```

<!-- ### 2D Plots -->
<!-- I select pairs of features and show the resulting scatterplots -->
<!-- ```{r} -->
<!-- ggplot_df = features_wells %>% select(1:12) -->
<!-- ggplot_df = as.data.frame( -->
<!--   apply(ggplot_df, 2, function(x) (x - mean(x)) / sd(x))) -->
<!-- ggplot_df$Line = features_wells$Line -->

<!-- ggplot(data = ggplot_df) +  -->
<!--   geom_point(mapping = aes(x = x.0.s.area, y = x.c.h.f12.s4, color = Line)) +  -->
<!--   theme_Publication() + scale_color_manual(values = colorScale) -->
<!-- ``` -->

## Phenoprints

```{r c26, fig.height=8}
features_lines = aggregate(features, list(metadata$Line), median)
# Reshape into a ggplot-compatible format
features_lines = melt(features_lines, id.vars = "Group.1")
# Rename features into more readable names
rename_dict = c(
  "x.0.s.area" = "Area", "x.a.b.q05" = "Actin Intensity Median", 
  "x.b.b.q05" = "FITC Intensity Median", "x.c.b.q05" = "DAPI Intensity Median", 
  "x.b.h.asm.s2" = "FITC Haralick ASM", "x.b.b.q001" = "FITC Intensity 1-Percentile", 
  "x.Bac.b.mad" = "Actin*DAPI Intensity MAD", "x.a.h.f13.s4" = "Actin Haralick F13", 
  "x.Ba.h.sav.s4" = "Actin Haralick SAV", "x.c.h.f13.s4" = "DAPI Haralick F13")

features_lines = features_lines[features_lines$variable %in% names(rename_dict), ]
features_lines$variable = rename_dict[as.character(features_lines$variable)]
colnames(features_lines) = c("Line", "Feature", "Value")

ggplot(data = features_lines) + geom_col(aes(x = Feature, y = Value, fill = Feature)) + 
  facet_wrap(facets = ~Line) + theme_Publication() + xlab("") + ylab("") + 
  theme(legend.position = "right", legend.direction = "vertical", 
        axis.text.x = element_blank(), legend.key.size = unit(0.5, "cm")) + 
  scale_fill_manual(values = colorScale)
```

