---
title: "Human Organoid Unsupervised Exploration"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE)
```

# Setup

sudo apt-get install libpng16-16

```{r}
library(tidyverse)
library(here)
library(monocle3)
library(ggrastr)
library(cowplot)
library(princurve)
library(scico)
library(ggridges)

# I wish I could solve my path problems with the here() package, but experienced unreliable behaviour 
PATH = "/dkfz/groups/shared/OE0049/B110-Isilon2/promise/"

umap_tidy <- read_rds(paste0(PATH, "data/processed/PhenotypeSpectrum/umap_absolute_all_drugs_tidy.Rds"))
pca_tidy <- read_rds(paste0(PATH, "data/processed/PhenotypeSpectrum/pca_absolute_all_drugs_tidy.Rds"))
umap_sampled <- read_rds(paste0(PATH, "data/processed/PhenotypeSpectrum/umap_absolute_all_drugs_sampled.Rds"))
```

# Organoid size Distributions

I plot a size-distribution. 

```{r, eval = TRUE}
gg_size_dist <- umap_tidy %>% 
  ggplot(aes(size)) + 
  geom_histogram() + 
  theme_classic()

gg_size_dist_log <- gg_size_dist + 
  scale_x_log10() 
```

I add the eCDF,

```{r}
df <- umap_sampled

gg_ecdf <- ggplot(df %>% filter(drug == "DMSO")) +
  stat_ecdf(aes(x = size, group = line), 
              geom = "step", size = 1.5) +
  #scale_color_manual(values = c("#00AFBB", "#E7B800"))+
  labs(y = "f(size)",
       x = "organoid size [pixels]") + 
  theme_cowplot()

gg_ecdf
```

For more details about distributions, please refer to *reports/Phenotypespectrum/'xyz'_dist.pdf*.


```{r}
line_param <- umap_sampled %>% 
  nest(-line, -replicate) %>% 
  mutate(fit = map(data, ~ fitdistrplus::fitdist(.x$size, "lnorm")),
         param = map(fit, ~ .x$estimate %>% broom::tidy()))

df <- line_param %>% unnest(param) %>% 
  filter(names == "meanlog") %>% 
  group_by(line) %>% 
  mutate(mean_meanlog = mean(x)) %>% 
  arrange(mean_meanlog) %>% 
  mutate(line = factor(line, levels = .$line %>% unique()))

df %>% ggplot(aes(x, line)) + 
  geom_point() + 
  theme_cowplot()

organoid_size_factor <- df$line %>% levels()
```

```{r}
organoid_size_factor_09 <- umap_sampled %>% group_by(line) %>% 
  summarise(x = quantile(size_log, 0.9)) %>% 
  #summarise(x = mean(size_log)) %>% 
  arrange(x) %>% .$line

umap_sampled %>% filter(drug == "DMSO") %>% 
  mutate(line = factor(line, levels = organoid_size_factor_09)) %>% 
  ggplot() +
  geom_density_ridges_gradient(aes(y = line, x = size_log, fill = stat(x)), scale = 1) +
  #geom_density(aes(x = size_log, group = replicate, color = morphological_class)) + 
  #facet_wrap(~ line) + 
  scale_fill_viridis_c() +
  labs(caption = "DMSO treated organoids",
       x = "ln(size)") + 
  theme(legend.position = "bottom") +
  theme_cowplot() + 
  ggsave(paste0(PATH, "reports/figures/gg_size_dist_morph_ridge.pdf"), 
         width = 4, height = 4)
```

```{r}
ggplot(umap_sampled %>% filter(drug == "DMSO") %>% 
         filter(line %in% c("D021T01", "D055T01"))) +
  geom_density(aes(x = size_log, color = line, group = replicate)) + 
  #facet_wrap(~ line, ncol = 1) + 
  theme_cowplot() + ggsave(paste0(PATH, "reports/figures/gg_size_dist_example.pdf"), width = 3, height = 6)
```


# Paclitaxel Case Study


```{r}
drug_size <- umap_tidy %>% filter(drug == "DMSO" | drug == "Paclitaxel") %>% 
  mutate(concentration = ifelse(drug == "DMSO", 0, concentration)) %>%
  #filter(morphological_class == "disorganized") %>% 
  #filter(morphological_class != "other") %>% 
  mutate(concentration = factor(concentration, levels = c("0", "0.0016", "0.008", "0.04", "0.2", "1.0")))

ggdrug_size <- ggplot(drug_size) +
  geom_density(aes(x = log(size), group = concentration, color = concentration), size = 1.5) + 
  scico::scale_color_scico_d() +
  theme_cowplot() + 
  #scale_x_continuous(limits = c(0, 15000)) +
  theme(legend.position = "bottom") + 
  labs(color = "Paclitaxel Concentration Factor",
       title = "Organoid size distribution",
       x = "ln(size)")

ggdrug_size 
```

```{r}
drug_count <- drug_size %>%
  dplyr::count(concentration, line, replicate, well)

ggdrug_count <- ggplot(drug_count) +
  geom_density(aes(x = log(n), group = concentration, color = concentration), size = 1.5) + 
  scico::scale_color_scico_d() +
  theme_cowplot() + 
  theme(legend.position = "bottom") + 
  labs(color = "Paclitaxel Concentration Factor",
       title = "Organoid count distribution",
       x = "ln(n)")

ggdrug_count + ggsave(paste0(PATH, "reports/figures/gg_drug_dist_n_log.pdf"))
```


```{r}
set.seed(234)
loi = c("D022T01", "D055T01")


df <- umap_tidy  %>%
  filter(drug == "DMSO" | drug == "Paclitaxel") %>% 
  mutate(concentration = ifelse(drug == "DMSO", 0, concentration)) %>% 
  filter(line == loi)

gg_drug <- umap_sampled %>%
  dplyr::select(-line, -concentration) %>%
  ggplot(aes(v1, v2)) + 
  geom_point_rast(alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point_rast(data = df  %>%
    group_by(concentration) %>% 
    sample_n(1000, replace = TRUE),
  aes(color = concentration),alpha = 1, size = 1.5, shape=16) + 
  #facet_wrap( ~ concentration, ncol = 1) + 
  #scale_color_brewer(type = "seq", palette = "YlOrRd") + 
  #geom_density2d(color = "black") + 
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2",
       caption = paste0(paste(loi, collapse=" "), ", Paclitaxel"),
       color = "Concentration Factor") + 
  scico::scale_color_scico_d() + 
  facet_wrap(~ line, ncol = 2) +
  theme(legend.position = "bottom") +
  theme_cowplot(font_size = 8) + 
  theme(legend.position = "bottom")
 
gg_drug
```



```{r}
loi <- c("D022T01", "D055T01")

drug_size_param <- drug_size %>% 
  nest(-concentration, -line) %>% 
  mutate(fit = map(data, ~ fitdistrplus::fitdist(.x$size, "lnorm")),
         param = map(fit, ~ .x$estimate %>% broom::tidy()))

df <- drug_size_param %>% unnest(param) %>% 
  filter(names == "meanlog") %>%
  mutate(concentration = factor(concentration, levels = c("0", "0.0016", "0.008", "0.04", "0.2", "1.0"))) 

gg_size_drug <- df %>%
  #mutate(concentration = as.numeric(as.character(concentration))) %>%
  ggplot(aes(concentration, x)) + 
    geom_point(color = "grey") + 
  geom_line(aes(group = line), color = "grey") +
  geom_point(data = df %>% filter(line %in% loi), color = "black") +
  geom_line(aes(group = line), data = df %>% filter(line %in% loi), color = "black") +
  labs(y = 'size ln(mean)' ,
       caption = paste0(paste(loi, collapse=" "), ", Paclitaxel")) +
  #scale_x_log10()
  theme_cowplot()

gg_size_drug
```




I use an alternative, manual, plotting method. In order to plot the object data, I export a subset of organoids. I plot the cells by size.

```{r}
set.seed(234)

df <- umap_tidy  %>%
  mutate(cystic = if_else(line == "D013T01" &  well == "D24" & plate == "D013T01P001L02", TRUE, FALSE)) %>%
  mutate(compact = if_else(line == "D055T01" &  well == "D24" & plate == "D055T01P007L02", TRUE, FALSE))

df %>% 
  sample_frac(0.01) %>%
  ggplot(aes(v1, v2, color = size_log)) + 
  #scale_color_brewer(type = "qual", palette = 2) +
  geom_point_rast(alpha = 0.1, size = 0.35) + 
  geom_point_rast(color = "red", alpha = 1, size = 0.5, data = df %>% filter(cystic == TRUE)) +
  geom_point_rast(color = "blue", alpha = 1, size = 0.5, data = df %>% filter(compact == TRUE)) + 
  scale_color_viridis_c() +
  theme_cowplot() + 
  coord_fixed()
```

I wonder what the small, predominantly cystic cluster, is consisting off. In order to rule out potential segmentation artefacts, I take a closer look at wells, for which we have generated segmentation masks beforehand. 

```{r}
set.seed(234)

df <- umap_tidy  %>%
  mutate(cystic = if_else(line == "D013T01" &  well == "B24" & plate == "D013T01P001L02", TRUE, FALSE))

df %>% 
  sample_frac(0.01) %>%
  ggplot(aes(v1, v2, color = size_log)) + 
  #scale_color_brewer(type = "qual", palette = 2) +
  geom_point_rast(alpha = 0.1, size = 0.35) + 
  geom_point_rast(color = "red", alpha = 1, size = 0.5, data = df %>% filter(cystic == TRUE)) +
  scale_color_viridis_c() +
  theme_cowplot() + 
  coord_fixed() + 
  ggtitle("organoids with one well of cystic line marked red")
```


```{r}
umap_tidy %>%
  dplyr::filter(line == "D013T01" &  well == "B24" & plate == "D013T01P001L02") %>% 
  mutate(original_x = as.numeric(original_x),
         original_y = as.numeric(original_y)) %>%
  #filter(v1 < -7 & v2 > 0) %>%
  mutate(field = factor(field, levels = c(1, 2, 4, 3))) %>%
  ggplot(aes(original_x, -original_y, size = size_log, color = v1)) + 
  geom_point() + 
  facet_wrap(~field, ncol = 2) + 
  coord_fixed() +
  viridis::scale_color_viridis() +
  theme_cowplot()
```


I also create a supplement comparing DMSO with other treatments

```{r}
set.seed(123)

df <- umap_sampled

gg_size_supp <- df %>%
  mutate(drug = if_else(drug == "DMSO", "DMSO", "other")) %>%
  ggplot(aes(v1, v2, color = size_log)) + 
  geom_point_rast(alpha = 0.5, size = 0.35) + 
  scale_color_viridis_c() +
  theme_cowplot() +
  labs(x = "UMAP 1",
       y = "UMAP 2") + 
  theme(legend.position = "bottom") +
  facet_wrap(~ drug)

gg_size_supp + ggsave(paste0(PATH, "reports/figures/gg_size_all.pdf"))
```

# Organoid line differences

I create a single plot showing the two extreme organoid lines and their distribution within the embedding. 

```{r}
set.seed(123)

loi <- c("D055T01", "D007T01",  "D027T01", "D018T01") #c("D055T01", "D007T01",  "D021T01", "D019T01", "D027T01")
#loi <- umap_sampled$line %>% unique()

df <- umap_sampled %>%
  filter(drug == "DMSO")

gg_line <- df %>% dplyr::select(-line) %>%
  ggplot(aes(v1, v2)) + 
  geom_point_rast(alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point_rast(data = umap_tidy %>%
                    filter(drug == "DMSO") %>% 
                   # filter(line %in% c("D021T01")) %>%
    filter(line %in% loi) %>% 
    mutate(line = factor(line, levels = loi)) %>% 
      sample_frac(0.1),
    #mutate(line = factor(line, levels = c("D021T01"))),
  aes(color = line),alpha = .4, size = 0.35, shape=16) + 
  facet_wrap( ~ line, ncol =2) +
  scale_color_brewer(type = "qual", palette = "Set2") +
  #scale_color_manual(values = c(c("#D80D12", "#461C01", "#9a4c91", "#70BE6F", "#24345E"))) + 	
  #geom_density2d(color = "black") + 
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme_cowplot(font_size = 8) + 
  theme(legend.position = "nothing")

gg_line + ggsave(paste0(PATH, "reports/figures/gg_size_all.pdf"), width = 4, height = 4)
```

_E_13_contrastProjections.h5

# Concentration dependent phenotype transitions

## Illustration

```{r}
set.seed(123)

drug_order <- umap_tidy %>%
    #filter(drug == "DMSO" | grepl(drug, pattern = "Bortezomib")) %>%
  filter(grepl(drug, pattern = "Bortezomib")) %>%
    filter(line == "D046T01") %>% .$drug %>% unique()
drug_order <- drug_order[rev(c(2,1,4,3,5))]

gg_bortezomib <- umap_sampled %>%
  dplyr::select(-line, -concentration) %>%
  ggplot(aes(v1, v2)) + 
  geom_point_rast(alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point_rast(data = umap_tidy %>%
    #filter(drug == "DMSO" | grepl(drug, pattern = "Bortezomib")) %>%
      filter(grepl(drug, pattern = "Bortezomib")) %>%
  filter(line %in% c("D055T01", "D054T01", "D046T01")) %>%
    mutate(concentration = if_else(concentration == "nan", "0", concentration)) %>%
    mutate(drug = factor(drug, levels =drug_order)) %>%
    
    group_by(drug) %>% 
    sample_n(5000, replace = TRUE),
  aes(color = drug),alpha = 1, size = 1.5, shape=16, color = "black") + 
  facet_wrap( ~ concentration, ncol = 1) + 
  #scale_color_brewer(type = "seq", palette = "YlOrRd") + 
  #geom_density2d(color = "black") + 
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme_cowplot(font_size = 8) + 
  theme(legend.position = "bottom")
  
gg_bortezomib

gg_bortezomib + ggsave(paste0(PATH, "reports/figures/gg_bortezomib.pdf"), width = 2, height = 8)
```




## Aggregated Pseudotimes

Now I am building a function to create these tidy path objects.

```{r}
create_princurve_trace <- function(df){
  df_tmp_init <- df %>% 
  group_by(concentration) %>% 
  summarize(v1 = mean(v1),
            v2 = mean(v2)) %>% 
  ungroup()
  
  fit <- principal_curve(x = df %>% dplyr::select(v1, v2) %>% as.matrix(),
                       start = df_tmp_init %>% dplyr::select(v1, v2) %>% as.matrix(),
                       approx_points = FALSE, 
                       trace = TRUE, 
                       plot_iterations = FALSE, 
                       maxit = 30,
                       stretch =2)
  
  plot(fit)
  points(df_tmp_init%>% dplyr::select(v1, v2) %>% as.matrix())

  result <- list()
  
  result$fit <- fit$s[fit$ord,] %>% as_tibble()
  result$center <- df_tmp_init
  result$input <- df
  
  return(result)
}
```

```{r}
set.seed(123)

drug_order <- umap_tidy %>%
    filter(drug == "DMSO" | grepl(drug, pattern = "Trametinib")| grepl(drug, pattern = "Bortezomib")) %>%
  filter(!grepl(drug, pattern = "Dabrafenib")) %>% 
  filter(!grepl(drug, pattern = "GSK1120212")) %>% .$drug %>% unique() %>% sort()

center_df <- umap_tidy %>%
    filter(drug %in% drug_order) %>%
  #filter(line %in% c("D055T01", "D054T01", "D046T01")) %>%
  dplyr::select(v1, v2, drug, line) %>% 
  group_by(drug, line) %>% 
  summarize(v1 = mean(v1),
            v2 = mean(v2)) %>% 
  separate(drug, c("drug", "concentration"), sep = "__") %>% 
  mutate(concentration = as.numeric(concentration),
         drug_line = paste0(drug, "_", line))

gg_trambort <- center_df %>% 
  ggplot(aes(v1, v2)) + 
  geom_point_rast(data = umap_tidy %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point(aes(color = drug, group = drug_line),alpha = 1, size = 1.5, shape=16) + 
  geom_smooth(aes(group = drug, color = drug), se = FALSE) + 
  #geom_path(aes(group = drug, color = drug), arrow = arrow(angle =15, ends = "last", type = "closed", length = unit(0.15, "inches"))) + 
  #geom_path(aes(group = drug_line, color = drug)) + 
  scale_color_manual(values = c("#707070", "#000000", "#4285F4")) +
  
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme_cowplot()

gg_trambort
```

I am creating a new version of the plot above

```{r}
drug_order <- c("Trametinib")

center_df <- umap_tidy %>%
    filter(drug %in% drug_order) %>%
  filter(line %in% c("D055T01", "D054T01", "D046T01")) %>% #, "D027T01", "D019T01"
  mutate(drug = factor(drug, levels =drug_order)) %>% 
  arrange((drug)) %>% 
  group_by(drug, line, concentration) %>%
  sample_n(200, replace = TRUE) %>% ungroup() # adjust based on pan-line vs. 

plot_df <- center_df %>%
  nest(-line, -drug) %>%
  mutate(plot = purrr::map(data, ~ create_princurve_trace(.x)))


gg_tram_traj <- 

  ggplot() + 
  geom_point_rast(aes(v1, v2), data = umap_tidy %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point(data = plot_df %>% unnest(data),
             aes(color = line, v1, v2), alpha = 0.5, size = 0.35) +
  geom_path(data = plot_df %>% mutate(plot_trace = purrr::map(plot, ~ .x$fit %>% .[, ])) %>% unnest(plot_trace), 
            aes(v1, v2, group = paste0(drug, line), color = line), size = 1.5, 
            arrow = arrow(angle = 10, ends = "last", type = "closed", length = unit(0.15, "inches")))+ 
  scale_color_brewer(type = "qual", palette = "Set2") +
  theme_cowplot() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
  theme(legend.position = "bottom") + 
  facet_wrap(~ drug)

gg_tram_traj + ggsave(paste0(PATH, "reports/figures/gg_tram_traj.pdf"), width = 4, height = 4)
```

I am repeating the same figure for D027T01

```{r}
set.seed(123)

drug_order <- c("Trametinib")

center_df <- umap_tidy %>%
    filter(drug %in% drug_order) %>%
  filter(line %in% c("D027T01")) %>% 
  mutate(drug = factor(drug, levels =drug_order)) %>% 
  arrange((drug)) %>% 
  group_by(drug, line, concentration) %>%
  sample_n(400, replace = TRUE) %>% ungroup() # adjust based on pan-line vs. 

plot_df <- center_df %>%
  nest(-line, -drug) %>%
  mutate(plot = purrr::map(data, ~ create_princurve_trace(.x)))


gg_tram_traj <- 

  ggplot() + 
  geom_point_rast(aes(v1, v2), data = umap_tidy %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point(data = plot_df %>% unnest(data),
             aes(color = line, v1, v2), alpha = 0.5, size = 0.35) +
  geom_path(data = plot_df %>% mutate(plot_trace = purrr::map(plot, ~ .x$fit %>% .[1:1500,])) %>% unnest(plot_trace), 
            aes(v1, v2, group = paste0(drug, line), color = line), size = 1.5, 
            arrow = arrow(angle = 10, ends = "last", type = "closed", length = unit(0.15, "inches")))+ 
  scale_color_brewer(type = "qual", palette = "Set2") +
  theme_cowplot() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
  theme(legend.position = "bottom") + 
  facet_wrap(~ drug)

gg_tram_traj + ggsave(paste0(PATH, "reports/figures/gg_tram_traj_D27_D19.pdf"), width = 4, height = 4)
```


```{r}
drug_order <- c("WYE-132", "DMSO")

center_df <- umap_tidy %>%
    filter(drug %in% drug_order) %>%
  filter(line %in% c("D030T01", "D027T01", "D013T01")) %>%
  group_by(drug, line, concentration) %>%
  sample_n(700, replace = TRUE) %>% ungroup()

gg_mtor_all <- umap_tidy %>%
    
  filter(drug == "DMSO" | grepl(drug, pattern = "WYE-132")) %>%
  filter(line %in% c("D030T01", "D027T01", "D013T01")) %>%
  
  group_by(drug) %>% 
  sample_n(700) %>%
  ggplot(aes(v1, v2)) + 
  
  geom_point_rast(data = umap_sampled %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
 
  geom_point_rast(aes(color = drug),alpha = 1, size = 1, shape=16, color = "black") + 
  facet_wrap(~ drug) + 
 
  theme_cowplot() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+

  theme(legend.position = "nothing")

gg_mtor_all + ggsave(paste0(PATH, "reports/figures/gg_mtor.pdf"), width = 6, height = 3)
```

```{r}
set.seed(123)
drug_order <- c("CHIR-98014", "DMSO")

center_df <- umap_tidy %>%
    filter(drug %in% drug_order) %>%
  #filter(line %in% c("D030T01", "D027T01", "D013T01")) %>%
  group_by(drug, line, concentration) %>%
  sample_n(500, replace = TRUE) %>% ungroup()

gg_mtor_all <- umap_tidy %>%
    
  filter(drug == "DMSO" | grepl(drug, pattern = drug_order[1])) %>%
  #filter(line %in% c("D030T01", "D027T01", "D013T01")) %>%
  
  group_by(drug) %>% 
  sample_n(500) %>%
  ggplot(aes(v1, v2)) + 
  
  geom_point_rast(data = umap_sampled %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
 
  geom_point_rast(aes(color = drug),alpha = 1, size = 1, shape=16, color = "black") + 
  facet_wrap(~ drug) + 
 
  theme_cowplot() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+

  theme(legend.position = "nothing")

gg_mtor_all + ggsave(paste0(PATH, "reports/figures/gg_gsk.pdf"), width = 6, height = 3)
```


```{r, eval = FALSE}
set.seed(123)

drug_order <- umap_tidy %>%
  dplyr::select(drug) %>%
    filter(grepl(drug, pattern = "__")) %>% .$drug %>% unique() %>% sort()

curve_df <- umap_tidy %>%
    filter(drug %in% drug_order) %>%
  mutate(drug = factor(drug, levels =drug_order)) %>% 
  arrange((drug)) %>% 
  group_by(drug, line) %>%
  sample_n(1000, replace = TRUE) %>% # adjust based on pan-line vs. line level 
  ungroup() %>% 
  # running the princcurve algorithm
  separate(drug, c("compound", "concentration"), sep = "__", remove = FALSE) %>%
  nest(-compound, -line) %>%
  #head(10) %>% 
  mutate(plot = purrr::map(data, ~ create_princurve_trace(.x)))

curve_df %>% saveRDS(here::here("vignettes/07_organoid_unsupervised_exploration/curve_df.Rds"))
```

## Individual Pseudotimes

```{r, eval = FALSE}
set.seed(123)

drug_order <- umap_tidy %>%
  dplyr::select(drug) %>%
    filter(grepl(drug, pattern = "__")) %>% .$drug %>% unique() %>% sort()

curve_df <- umap_tidy %>%
    filter(drug %in% drug_order) %>%
  mutate(drug = factor(drug, levels =drug_order)) %>% 
  arrange((drug)) %>% 
  group_by(drug, line) %>%
  sample_n(1000, replace = TRUE) %>% # adjust based on pan-line vs. line level 
  ungroup() %>% 
  # running the princcurve algorithm
  separate(drug, c("compound", "concentration"), sep = "__", remove = FALSE) %>%
  nest(-compound, -line) %>%
  #head(10) %>% 
  mutate(plot = purrr::map(data, ~ create_princurve_trace(.x)))

curve_df %>% saveRDS(here::here("vignettes/07_organoid_unsupervised_exploration/curve_df.Rds"))
```


```{r, eval = FALSE}
gg_curve_df <- curve_df %>% 
  mutate(compound_line = paste0(compound, "__", line)) %>%
  mutate(plot_trace = purrr::map(plot, ~ .x$fit %>% .[, ])) %>% 
  unnest(plot_trace) 

gg_curve_df %>%
  ggplot(aes(v1, v2)) + 
  geom_point(data = curve_df %>% unnest(data),
             aes(color = compound), alpha = 0.05, size = 0.35) +
  
  geom_path(aes(group = compound_line, color = compound_line), size = 1.5, arrow = arrow(angle = 10, ends = "last", type = "closed", length = unit(0.15, "inches"))) + 
  
  facet_grid(~ compound) +
   
  theme_cowplot() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
  theme(legend.position = "bottom")
```

# Plot Export

```{r}
gg_size_dist + ggsave(paste0(PATH, "reports/figures/gg_size_dist.pdf"))
gg_size_dist_log +  ggsave(paste0(PATH, "reports/figures/gg_size_dist_log.pdf"))
ggdrug_size + ggsave(paste0(PATH, "reports/figures/gg_drug_dist_size_log.pdf"))
gg_ecdf + ggsave(paste0(PATH, "reports/figures/gg_size_dist_ecdf.pdf"))

gg_drug + ggsave(paste0(PATH, "reports/figures/gg_drug.pdf"), 
         width = 8, height = 4)
gg_size_drug + ggsave(paste0(PATH, "reports/figures/gg_trametinib_size_dose.pdf"), width = 3.65, height = 3.65)
```


# Supplement

Here I collect pieces of code that did not make it into the final analysis but can be run in theory. In order to access these peaces of code, you have to open the *.RMD* file. 

```{r}
knitr::knit_exit()
```

## Mixture modeling of Organoid size distribution


I estimate a log-gaussian mixture model of organoid size.

```{r, eval = TRUE}
library(mixtools)

df_mix <- normalmixEM(umap_tidy$size_log, k = 2)

df_fus <- df %>% dplyr::select(size_log) %>% mutate(posterior = "joined") %>% 
  rbind(data.frame(size_log = df_mix$x,
                   posterior = df_mix$posterior %>% as.data.frame() %>% round() %>% .$comp.1))

ggplot(df) + 
  geom_density(aes(x = size_log, y=..count..), alpha = 0.2) + 
  theme_cowplot()

ggplot(df_fus) + 
  geom_density(aes(x = size_log, y=..count.., fill = posterior), alpha = 0.2) + 
  theme_cowplot()
```

```{r, eval = TRUE}
df_mix
```



I scale this approach to all other organoid lines. 

```{r}
set.seed(0123349)

umap_mix <- umap_tidy %>% filter(drug == "DMSO")  %>% 
  nest(-line, -replicate) %>%
  mutate(mix_fit = map(data, ~ normalmixEM(.x$size_log, k = 2, maxit = 5000)),
         prob = map(mix_fit, ~ .x$posterior %>% as.data.frame() %>% .$comp.1)) 

umap_mix = umap_mix %>% unnest(prob, data) 
```

```{r}
set.seed(0.1)

df = umap_mix %>% mutate(posterior = round(prob)) %>% mutate(posterior = factor(posterior))

df %>% 
  sample_frac(0.01) %>%
  ggplot(aes(prob_dead, fill = posterior)) + 
  geom_histogram() + 
  scale_fill_brewer(type= "qual") + 
  scale_y_log10() + 
  theme_cowplot()

ggplot(df) + 
  geom_density(aes(x = size_log, fill = posterior, group = replicate), alpha = 0.5) + 
  facet_wrap(~ line) + 
  labs(caption = "Plotted are Replicates, DMSO treated organoids",
       x = "ln(size)") + 
  theme(legend.position = "bottom") +
  theme_cowplot()


ggplot(df) + 
  geom_density(aes(x = size_log, y=..count.., fill = posterior), alpha = 0.2) + 
  theme_cowplot() + 
  facet_wrap(~ line)
```

## Count distribution of Paclitaxel treated organoids 



```{r}
drug_count_param <- drug_count %>% 
  nest(-concentration, -line) %>% 
  mutate(fit = map(data, ~ fitdistrplus::fitdist(.x$n, "lnorm")))
        # param = map(fit, ~ .x$estimate %>% broom::tidy()))

drug_count_param %>% unnest(param) %>% 
  filter(names == "meanlog") %>%
  ggplot(aes(concentration, exp(x))) + 
    geom_point() + 
  theme_cowplot()
```


```{r}
drug_count_param %>% unnest(param) %>% 
  filter(names == "meanlog") %>% 
  dplyr::select(concentration, line, count = x) %>% 
  left_join(tram_size_param %>% unnest(param) %>% 
  filter(names == "meanlog") %>% dplyr::select(concentration, line, size = x)) %>% 
  
  ggplot(aes(count, size, color = concentration)) + 
  geom_point() + 
 # geom_path(aes(group = line)) +
  scico::scale_color_scico_d() + 
  theme_cowplot()


```




## Main Panel

```{r}
plot_grid(gg_size, 
          plot_grid(gg_line,
                    gg_bortezomib,
                    ncol = 2),
          labels = "AUTO",
          ncol = 2) + 
  gggsave(here::here("results/figures/panel_6.pdf"))
```

## Supplemental Panel

```{r}
plot_grid(gg_harmony_effect, 
          gg_size_supp,
          gg_size_dist
          )
```

## Running harmony on batches

We abandoned this approach as we did not screen organoid lines across batches. Reducing batch differences automatically led to a reduction of line specific effects. 


First, I illustrate batch effects and how they are reduced

```{r, eval = FALSE}
set.seed(123)

pca_metadata_frac <- pca_metadata %>% sample_frac(0.001)

pca_harmony <- pca_metadata_frac %>% dplyr::select(contains("V"))
metadata_harmony <- pca_metadata_frac %>% dplyr::select(screen_ID, Line)

my_harmony_embeddings <- HarmonyMatrix(
  pca_harmony, metadata_harmony, c("screen_ID"),
  do_pca = FALSE,
  verbose = TRUE, 
  return_object = TRUE
)

umap_pre <- uwot::umap(my_harmony_embeddings$Z_orig %>% t() %>% as_tibble(), verbose = TRUE)
umap_post <- uwot::umap(my_harmony_embeddings$Z_corr %>% t() %>% as_tibble(), verbose = TRUE)

harmony_effect <- umap_pre %>% as_tibble() %>% mutate(status = "pre") %>% cbind(id = 1:nrow(.)) %>% cbind(metadata_harmony)%>%
  rbind(.,umap_post %>% as_tibble() %>% mutate(status = "post")%>% cbind(id = 1:nrow(.)) %>% cbind(metadata_harmony)) 

harmony_effect %>% saveRDS(here::here("vignettes/07_organoid_unsupervised_exploration/harmony/harmony_effect.Rds"))
```

Here we can see the effect of Harmony on batch discordances.

```{r, eval = TRUE}
harmony_effect <- readRDS(here::here("vignettes/07_organoid_unsupervised_exploration/harmony/harmony_effect.Rds"))

gg_harmony_effect <- harmony_effect %>% 
  mutate(status = factor(status, levels = c("pre", "post"))) %>%
  ggplot(aes(V1, V2, color = screen_ID)) + 
  geom_point_rast(alpha = 0.5, size = 0.35) + 
  theme_cowplot() + 
  scale_color_brewer(type = "qual", palette = 2) +
  facet_wrap( ~ status) + 
  #ggsave(here::here("vignettes/07_organoid_unsupervised_exploration/pre_post_harmony.png")) + 
  NULL
```

Still, line-specific differences are conserved in this method.

```{r}
harmony_effect %>% 
 filter(status == "post") %>%
  ggplot(aes(V1, V2, color = screen_ID)) + 
  geom_point_rast(alpha = 0.5, size = 0.35) + 
  theme_cowplot() + 
  facet_wrap( ~ Line) + 
  scale_color_brewer(type = "qual") +
  #ggsave(here::here("vignettes/07_organoid_unsupervised_exploration/post_harmony_line.png")) + 
  NULL
```


## Running Harmony on replicate and on replicate+line level

```{r, eval = FALSE}
set.seed(123)

pca_harmony <- pca_metadata %>% dplyr::select(contains("V"))
metadata_harmony <- pca_metadata %>% dplyr::select(screen_ID, Line)

harmony_id <- HarmonyMatrix(
  pca_harmony, metadata_harmony, c("screen_ID"),
  do_pca = FALSE,
  verbose = TRUE, 
  return_object = TRUE
)

harmony_id$Z_corr %>% saveRDS(here::here("vignettes/07_organoid_unsupervised_exploration/harmony/harmony_pca_id.Rds"))
```

I load the Harmony results.

```{r, eval = FALSE}
harmony_id_corr <- readRDS(here::here("vignettes/07_organoid_unsupervised_exploration/harmony/harmony_pca_id.Rds"))
```



## Running harmony on both line and batch 

```{r, eval = FALSE}
harmony_id_line <- HarmonyMatrix(
  pca_harmony, metadata_harmony, c("screen_ID", "Line"),
  do_pca = FALSE,
  verbose = TRUE, 
  return_object = TRUE
)


harmony_id_line %>% saveRDS(here::here("vignettes/07_organoid_unsupervised_exploration/harmony/harmony_pca_id_line.Rds"))

```

## Plotting gigantic overview figure


```{r, eval = FALSE}
gg_line_overview <- plot_cells(ods, color_cells_by="morphological_class",
           alpha = 0.05, label_cell_groups = FALSE,
           rasterize = TRUE) + 
  facet_wrap(~ Line) + 
  theme(legend.position = "nothing") + 
  scale_color_manual(values = rev(c("#A2549B", "#D80D12", "#6E1614","#69B563", "#A9CEE1","#1E3B87"))) 
```

I directly write this supplementary plot to disc. 

```{r, eval = FALSE}
gg_line_overview + 
  ggsave(here("results/figures/single_organoid/line.pdf"), height = 8, width = 8)
```

## Running princurve on principle component data 


 In the past, I estimated a princurve directly from the UMAP embedding. I am not sure this is a good idea. Rather, I am going to test running the princurve algorithm on a set of PCs. 
What is the %-Variance explained per PC?

```{r}
vars_pca <- apply(reducedDims(ods)$PCA, 2, var)
vars_pca_scaled = vars_pca/sum(vars_pca)

vars_pca_scaled %>% plot()
```

This is the PCA-based princurve function. 

```{r}
create_princurve_trace_pca <- function(df){
  df_tmp_init <- df %>% 
  group_by(drug) %>% 
  summarise_at(vars(contains("pc")), funs(mean)) %>%
  ungroup()
  
  fit <- principal_curve(x = df %>% dplyr::select(pc1:pc5) %>% as.matrix(),
                       start = df_tmp_init %>% dplyr::select(pc1:pc5) %>% as.matrix(),
                       approx_points = FALSE, 
                       trace = TRUE, 
                       plot_iterations = FALSE, 
                       maxit = 30,
                       stretch =2)
  
  plot(fit)
  points(df_tmp_init%>% dplyr::select(pc1:pc5) %>% as.matrix())

  result <- list()
  
  result$fit <- fit$s[fit$ord,] %>% as_tibble()
  result$center <- df_tmp_init
  result$input <- df
  
  return(result)
}
```


```{r}
set.seed(123)

drug_order <- pca_tidy %>%
  dplyr::select(drug) %>%
    filter(grepl(drug, pattern = "__")) %>% .$drug %>% unique() %>% sort()

curve_df <- pca_tidy %>%
    filter(drug %in% drug_order) %>%
  mutate(drug = factor(drug, levels =drug_order)) %>% 
  arrange((drug)) %>% 
  group_by(drug, line) %>%
  sample_n(1000, replace = TRUE) %>% # adjust based on pan-line vs. line level 
  ungroup() %>% 
  # running the princcurve algorithm
  separate(drug, c("compound", "concentration"), sep = "__", remove = FALSE) %>%
  nest(-compound, -line) %>%
  head(1) %>% 
  mutate(plot = purrr::map(data, ~ create_princurve_trace_pca(.x)))
```


```{r}
set.seed(123)

gg_curve_df <- curve_df %>% 
  mutate(compound_line = paste0(compound, "__", line)) %>%
  mutate(plot_trace = purrr::map(plot, ~ .x$fit %>% .[, ])) %>% 
  unnest(plot_trace) 

pca_subset = pca_tidy %>% sample_frac(.01)
sample_index = pca_subset$uuid


curve_index = FNN::get.knnx(data = pca_subset %>% dplyr::select(pc1:pc5) %>% as.matrix(), gg_curve_df %>% dplyr::select(pc1:pc5) %>% as.matrix(), k=1, algorithm=c("kd_tree"))

gg_curve_df = cbind(gg_curve_df, umap_tidy %>% filter(uuid %in% sample_index) %>% .[curve_index$nn.index,] %>% dplyr::select(v1, v2))

gg_curve_df %>%
  ggplot(aes(v1, v2)) + 
  geom_point(data = curve_df %>% unnest(data) %>% dplyr::select(compound, line, uuid) %>% left_join(umap_tidy %>% dplyr::select(uuid, v1, v2)),
             aes(color = compound), alpha = 0.05, size = 0.35) +
  
  #geom_path(aes(group = compound_line, color = compound_line), size = 1.5, arrow = arrow(angle = 10, ends = "last", type = "closed", length = unit(0.15, "inches"))) + 
  geom_point(aes(group = compound_line, color = compound_line), size = 1.5) +
  #facet_grid(~ compound) +
  facet_wrap(~ compound_line) +
  theme_cowplot() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
  theme(legend.position = "bottom")
```


