---
title: "Human Organoid Unsupervised Exploration"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE)
```

Loading packages

```{r}
library(tidyverse)
library(here)
library(ggrastr)
library(cowplot)
library(princurve)
library(scico)
library(ggridges)

# I wish I could solve my path problems with the here() package, but experienced unreliable behaviour 
PATH = "/dkfz/groups/shared/OE0049/B110-Isilon2/promise/"

umap_tidy <- read_rds(paste0(PATH, "data/processed/PhenotypeSpectrum/umap_absolute_all_drugs_tidy.Rds"))
#pca_tidy <- read_rds(paste0(PATH, "data/processed/PhenotypeSpectrum/pca_absolute_all_drugs_tidy.Rds"))
umap_sampled <- read_rds(paste0(PATH, "data/processed/PhenotypeSpectrum/umap_absolute_all_drugs_sampled.Rds"))
```

# Organoid Size Distributions

I plot a size-distribution. 

```{r, eval = TRUE}
gg_size_dist <- umap_tidy %>% 
  ggplot(aes(size)) + 
  geom_histogram() + 
  theme_cowplot()

gg_size_dist_log <- gg_size_dist + 
  scale_x_log10() 
```

I add the eCDF.

```{r}
df <- umap_sampled

gg_ecdf <- ggplot(df %>% filter(drug == "DMSO")) +
  stat_ecdf(aes(x = size, group = line), 
              geom = "step", size = 1) +
  #scale_color_manual(values = c("#00AFBB", "#E7B800"))+
  labs(y = "f(size)",
       x = "organoid size [pixels]") + 
  theme_cowplot()

gg_ecdf
```

For more details about distributions, please refer to *reports/Phenotypespectrum/'xyz'_dist.pdf*.

```{r}
line_param <- umap_sampled %>% 
  nest(-line, -replicate) %>% 
  mutate(fit = map(data, ~ fitdistrplus::fitdist(.x$size, "lnorm")),
         param = map(fit, ~ .x$estimate %>% broom::tidy()))

df <- line_param %>% unnest(param) %>% 
  filter(names == "meanlog") %>% 
  group_by(line) %>% 
  mutate(mean_meanlog = mean(x)) %>% 
  arrange(mean_meanlog) %>% 
  mutate(line = factor(line, levels = .$line %>% unique()))

organoid_size_factor <- df$line %>% levels()

df <- df %>% 
    dplyr::select(line, replicate, x) %>%
    pivot_wider(names_from = replicate, 
                values_from = x) 

r_size = df %>% ungroup() %>% dplyr::select(-line) %>% as.matrix %>% cor() %>% min()

gg_size_replicate <- df %>% 
  ggplot(aes(`1`, `2`)) +
  geom_smooth(method = "lm", se = FALSE, color = "grey") +
  geom_point() + 
  theme_cowplot() + 
  labs(x = "Replicate 1, mu [ln size]",
       y = "Replicate 2, mu [ln size]",
       caption = paste0("2 replicates, r= ", round(r_size, 2))) + 
  coord_fixed(ratio = 1) 
  #geom_abline(slope = 1, color = "grey")
  

gg_size_replicate
```

```{r}
organoid_size_factor_09 <- umap_sampled %>% group_by(line) %>% 
  summarise(x = quantile(size_log, 0.9)) %>% 
  #summarise(x = mean(size_log)) %>% 
  arrange(x) %>% .$line

gg_size_dist_morph_ridge <- umap_sampled %>% filter(drug == "DMSO") %>% 
  mutate(line = factor(line, levels = organoid_size_factor_09)) %>% 
  ggplot() +
  geom_density_ridges_gradient(aes(y = line, x = size_log, fill = stat(x)), scale = 1) +
  #geom_density(aes(x = size_log, group = replicate, color = morphological_class)) + 
  #facet_wrap(~ line) + 
  scale_fill_viridis_c() +
  labs(caption = "DMSO treated organoids",
       x = "ln(size)",
       fill = "size") + 
  theme(legend.position = "bottom") +
  theme_cowplot() 

gg_size_dist_morph_ridge
```

```{r}
umap_size <- function(umap, main){
  umap %>%
  #filter(Size < 1000) %>%
  ggplot(aes(v1, v2, color = size_log)) + 
  geom_point_rast(alpha = 0.5, size = 0.35) + 
  scale_color_viridis_c() +
  theme_cowplot() +
  labs(x = "UMAP 1",
       y = "UMAP 2",
       title = main,
       color = "ln(size)") + 
  theme(legend.position = "bottom")
}

p2 <- umap_size(umap_sampled, "Raw Data, All data")
p2 +  ggsave(paste0(PATH, "notebooks/PhenotypeSpectrum/gg_size.pdf"))
```



# Organoid Size during Drug Treatment


```{r}
drug_size <- umap_tidy %>% filter(drug == "DMSO" | drug == "Paclitaxel") %>% 
  mutate(concentration = ifelse(drug == "DMSO", 0, concentration)) %>%
  #filter(morphological_class == "disorganized") %>% 
  #filter(morphological_class != "other") %>% 
  mutate(concentration = factor(concentration, levels = c("0", "0.0016", "0.008", "0.04", "0.2", "1.0")))

ggdrug_size <- ggplot(drug_size) +
  geom_density(aes(x = log(size), group = concentration, color = concentration), size = 1.5) + 
  scico::scale_color_scico_d() +
  theme_cowplot() + 
  #scale_x_continuous(limits = c(0, 15000)) +
  theme(legend.position = "bottom") + 
  labs(color = "Paclitaxel Concentration Factor",
       title = "Organoid size distribution",
       x = "ln(size)")

ggdrug_size 
```

```{r}
drug_count <- drug_size %>%
  dplyr::count(concentration, line, replicate, well)

ggdrug_count <- ggplot(drug_count) +
  geom_density(aes(x = log(n), group = concentration, color = concentration), size = 1.5) + 
  scico::scale_color_scico_d() +
  theme_cowplot() + 
  theme(legend.position = "bottom") + 
  labs(color = "Paclitaxel Concentration Factor",
       title = "Organoid count distribution",
       x = "ln(n)")

ggdrug_count
```


```{r}
set.seed(234)
loi = c("D022T01", "D055T01")


df <- umap_tidy  %>%
  filter(drug == "DMSO" | drug == "Paclitaxel") %>% 
  mutate(concentration = ifelse(drug == "DMSO", 0, concentration)) %>% 
  filter(line == loi)

gg_drug <- umap_sampled %>%
  dplyr::select(-line, -concentration) %>%
  ggplot(aes(v1, v2)) + 
  geom_point_rast(alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point_rast(data = df  %>%
    group_by(concentration) %>% 
    sample_n(1000, replace = TRUE),
  aes(color = concentration),alpha = 1, size = 1.5, shape=16) + 
  #facet_wrap( ~ concentration, ncol = 1) + 
  #scale_color_brewer(type = "seq", palette = "YlOrRd") + 
  #geom_density2d(color = "black") + 
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2",
       caption = paste0(paste(loi, collapse=" "), ", Paclitaxel"),
       color = "Concentration Factor") + 
  scico::scale_color_scico_d() + 
  facet_wrap(~ line, ncol = 2) +
  theme(legend.position = "bottom") +
  #theme_cowplot(font_size = 8) + 
  theme(legend.position = "bottom")
 
gg_drug
```



```{r}
loi <- c("D022T01", "D055T01")

drug_size_param <- drug_size %>% 
  nest(-concentration, -line) %>% 
  mutate(fit = map(data, ~ fitdistrplus::fitdist(.x$size, "lnorm")),
         param = map(fit, ~ .x$estimate %>% broom::tidy()))

df <- drug_size_param %>% unnest(param) %>% 
  filter(names == "meanlog") %>%
  mutate(concentration = factor(concentration, levels = c("0", "0.0016", "0.008", "0.04", "0.2", "1.0"))) 

gg_size_drug <- df %>%
  filter(line %in% loi) %>%
  #mutate(concentration = as.numeric(as.character(concentration))) %>%
  ggplot(aes(concentration, x)) + 
    geom_point(color = "grey") + 
  geom_line(data = df %>% dplyr::rename(line_h = line) , aes(group = line_h), color = "grey") +
  geom_point(data = df %>% dplyr::rename(line_h = line), color = "grey") +
  geom_point(color = "black") +
  geom_line(aes(group = line), color = "black") +
  labs(y = 'mu ln(size)' ,
       x = "Concentration Factor",
       caption = paste0(paste(loi, collapse=" "), ", Paclitaxel")) +
  facet_wrap(~ line)+
  theme_cowplot()

gg_size_drug
```

# Organoid Heterogeneity

I plot 2 organoid lines treated with DMSO control

```{r}
set.seed(234)

df <- umap_tidy  %>%
  mutate(cystic = if_else(line == "D013T01" &  well == "D24" & plate == "D013T01P001L02", TRUE, FALSE)) %>%
  mutate(compact = if_else(line == "D055T01" &  well == "D24" & plate == "D055T01P007L02", TRUE, FALSE))

gg_cys_comp <- df %>% 
  sample_frac(0.01) %>%
  ggplot(aes(v1, v2, color = size_log)) + 
  #scale_color_brewer(type = "qual", palette = 2) +
  geom_point_rast(alpha = 0.1, size = 0.35) + 
  geom_point_rast(color = "#F4B400", alpha = 1, size = 0.5, data = df %>% filter(cystic == TRUE)) +
  geom_point_rast(color = "#DB4437", alpha = 1, size = 0.5, data = df %>% filter(compact == TRUE)) + 
  scale_color_viridis_c() +
  labs(color = "size",
       caption = "yellow: D013T01, red: D055T01",
       x = "UMAP 1",
       y = "UMAP 2") +
  theme_cowplot() + 
  coord_fixed()

gg_cys_comp
```

In general, DMSO treated organoid lines cover the same latent space than drug treated organoids. This is likely influenced by the large number of untreated organoids in the dataset.

```{r}
set.seed(123)

df <- umap_sampled

gg_size_supp <- df %>%
  mutate(drug = if_else(drug == "DMSO", "DMSO", "other")) %>%
  ggplot(aes(v1, v2, color = size_log)) + 
  geom_point_rast(alpha = 0.5, size = 0.35) + 
  scale_color_viridis_c() +
  theme_cowplot() +
  labs(x = "UMAP 1",
       y = "UMAP 2") + 
  theme(legend.position = "bottom") +
  facet_wrap(~ drug)

gg_size_supp + ggsave(paste0(PATH, "reports/figures/gg_size_all.pdf"))
```

# Organoid line differences

I create a single plot showing the two extreme organoid lines and their distribution within the embedding. 

```{r}
set.seed(123)

loi <- c("D055T01", "D007T01",  "D027T01", "D018T01") #c("D055T01", "D007T01",  "D021T01", "D019T01", "D027T01")
#loi <- umap_sampled$line %>% unique()

df <- umap_sampled %>%
  filter(drug == "DMSO")

gg_line <- df %>% dplyr::select(-line) %>%
  ggplot(aes(v1, v2)) + 
  geom_point_rast(alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point_rast(data = umap_tidy %>%
                    filter(drug == "DMSO") %>% 
                   # filter(line %in% c("D021T01")) %>%
    filter(line %in% loi) %>% 
    mutate(line = factor(line, levels = loi)) %>% 
      sample_frac(0.1),
    #mutate(line = factor(line, levels = c("D021T01"))),
  aes(color = line),alpha = .4, size = 0.35, shape=16) + 
  facet_wrap( ~ line, ncol =2) +
  scale_color_brewer(type = "qual", palette = "Set2") +
  #scale_color_manual(values = c(c("#D80D12", "#461C01", "#9a4c91", "#70BE6F", "#24345E"))) + 	
  #geom_density2d(color = "black") + 
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme_cowplot(font_size = 8) + 
  theme(legend.position = "nothing")

gg_line + ggsave(paste0(PATH, "reports/figures/gg_size_all.pdf"), width = 4, height = 4)
```
I am focusing on cystic vs solid organoid lines

```{r}
#UMAP Cystic (Lines 18, 13, 27, 30) vs. Solid (others) treated with DMSO, for Figure 1 / matching expression analysis done for cystic vs. rest

set.seed(123)

loi <- c("D018T01", "D013T01",  "D027T01", "D030T01") #c("D055T01", "D007T01",  "D021T01", "D019T01", "D027T01")
#loi <- umap_sampled$line %>% unique()

df <- umap_tidy %>%
  filter(drug == "DMSO") %>% 
  mutate(morphology = case_when(line %in% loi ~ "cystic",
                                 TRUE ~ "other"))

gg_cystic <- umap_sampled %>% 
  ggplot(aes(v1, v2)) + 
  geom_point_rast(alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point_rast(data = df %>%
    filter(morphology == "cystic") %>% 
      sample_frac(0.05),
  aes(color = morphology),alpha = .4, size = 0.35, shape=16) + 
 # facet_wrap( ~ line, ncol =2) +
  scale_color_brewer(type = "qual", palette = "Set2") +
  #scale_color_manual(values = c(c("#D80D12", "#461C01", "#9a4c91", "#70BE6F", "#24345E"))) + 	
  #geom_density2d(color = "black") + 
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme_cowplot(font_size = 8) + 
  theme(legend.position = "nothing")

gg_cystic
```


# Concentration dependent phenotype transitions

## Aggregate Transisition

```{r}
set.seed(123)

drug_order <- umap_tidy %>%
    #filter(drug == "DMSO" | grepl(drug, pattern = "Bortezomib")) %>%
  filter(grepl(drug, pattern = "Bortezomib")) %>%
    filter(line == "D046T01") %>% .$drug %>% unique()
drug_order <- drug_order[rev(c(2,1,4,3,5))]

gg_bortezomib <- umap_sampled %>%
  dplyr::select(-line, -concentration) %>%
  ggplot(aes(v1, v2)) + 
  geom_point_rast(alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point_rast(data = umap_tidy %>%
    #filter(drug == "DMSO" | grepl(drug, pattern = "Bortezomib")) %>%
      filter(grepl(drug, pattern = "Bortezomib")) %>%
  filter(line %in% c("D055T01", "D054T01", "D046T01")) %>%
    mutate(concentration = if_else(concentration == "nan", "0", concentration)) %>%
    mutate(drug = factor(drug, levels =drug_order)) %>%
    
    group_by(drug) %>% 
    sample_n(5000, replace = TRUE),
  aes(color = drug),alpha = 1, size = 1.5, shape=16, color = "black") + 
  facet_wrap( ~ concentration, ncol = 1) + 
  #scale_color_brewer(type = "seq", palette = "YlOrRd") + 
  #geom_density2d(color = "black") + 
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
       #caption = "control treated organoids") + 
  theme(legend.position = "bottom") +
  theme_cowplot(font_size = 8) + 
  theme(legend.position = "bottom")+ 
  coord_fixed()
  
gg_bortezomib
```



```{r}
drug_order <- c("WYE-132", "DMSO")

center_df <- umap_tidy %>%
    filter(drug %in% drug_order) %>%
  filter(line %in% c("D030T01", "D027T01", "D013T01")) %>%
  group_by(drug, line, concentration) %>%
  sample_n(700, replace = TRUE) %>% ungroup()

gg_mtor_all <- umap_tidy %>%
    
  filter(drug == "DMSO" | grepl(drug, pattern = "WYE-132")) %>%
  filter(line %in% c("D030T01", "D027T01", "D013T01")) %>%
  
  group_by(drug) %>% 
  sample_n(700) %>%
  ggplot(aes(v1, v2)) + 
  
  geom_point_rast(data = umap_sampled %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
 
  geom_point_rast(aes(color = drug),alpha = 1, size = 1, shape=16, color = "black") + 
  facet_wrap(~ drug) + 
 
  theme_cowplot() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+

  theme(legend.position = "nothing")

gg_mtor_all
```

```{r}
set.seed(123)
drug_order <- c("CHIR-98014", "DMSO")

center_df <- umap_tidy %>%
    filter(drug %in% drug_order) %>%
  #filter(line %in% c("D030T01", "D027T01", "D013T01")) %>%
  group_by(drug, line, concentration) %>%
  sample_n(500, replace = TRUE) %>% ungroup()

gg_gsk_all <- umap_tidy %>%
    
  filter(drug == "DMSO" | grepl(drug, pattern = drug_order[1])) %>%
  #filter(line %in% c("D030T01", "D027T01", "D013T01")) %>%
  
  group_by(drug) %>% 
  sample_n(500) %>%
  ggplot(aes(v1, v2)) + 
  
  geom_point_rast(data = umap_sampled %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
 
  geom_point_rast(aes(color = drug),alpha = 1, size = 1, shape=16, color = "black") + 
  facet_wrap(~ drug) + 
 
  theme_cowplot() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+

  theme(legend.position = "nothing")


```

```{r}
set.seed(123)
drug_order <- c("Ulixertinib (BVD-523, VRT752271)", 
                "TAK-733",
                "Trametinib (GSK1120212)",
                "Cobimetinib (GDC-0973, RG7420)",
                "MEK162 (ARRY-162, ARRY-438162)",
                "PD0325901",
                "Refametinib (RDEA119, Bay 86-9766)",
                "AZD8330",
                "Pimasertib (AS-703026)",
                "Selumetinib (AZD6244)",
                "PD318088",
                "VX-11e",
                "DMSO")

center_df <- umap_tidy %>% filter(concentration == "nan") %>%
    filter(drug %in% drug_order) %>%
  group_by(drug, line, concentration) %>%
  sample_n(500, replace = TRUE) %>% 
  summarise(v1 = mean(v1),
            v2 = mean(v2)) %>%
  ungroup()

center_df %>%
  ggplot(aes(v1, v2)) + 
  geom_point_rast(data = umap_sampled %>% dplyr::select(-drug, -line), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point_rast(aes(color = drug),alpha = 1, size = 1, shape=16) + 
  facet_wrap(~ line) + 
  theme_cowplot() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
  theme(legend.position = "nothing")
```


```{r}
create_plot_df_bulk <- function(loi, drug_order, n_sample, v1_cut){
  set.seed(123)
  # I am pulling in all drug observations and set concentration to 1
  drug_df <- umap_tidy %>%
    filter(drug %in% drug_order) %>%
    filter(line %in% loi) %>%
  mutate(concentration = if_else(concentration == "nan", 1, concentration))
  # I identify DMSO treated observations and set a concentration to 0
  dmso_df <- umap_tidy %>%
    filter(drug == "DMSO") %>%
    filter(line %in% loi) %>%
    mutate(concentration = 0) 
  # I reassign every DMSO concentration to a drug of interest
  assigned = rep(drug_order, times = nrow(dmso_df)/length(drug_order))
  # I join back both tables having created artificial drug treatments with concetration 0 from DMSO data
  center_df <- dmso_df %>% 
    head(length(assigned)) %>%
    cbind(assigned) %>% 
    mutate(drug = assigned) %>% 
    dplyr::select(-assigned) %>% 
    rbind(drug_df) %>%
    group_by(drug, line, concentration) %>%
    filter(v1 <= v1_cut) %>%
    sample_n(n_sample, replace = TRUE) %>% ungroup() # adjust based on pan-line

  plot_df <- center_df %>%
  nest(-line, -drug) %>%
  mutate(plot = purrr::map(data, ~ create_princurve_trace(.x)))

  return(plot_df)
}

draw_trajectory_bulk <- function(df){
  gg <- ggplot() + 
  geom_point_rast(aes(v1, v2), data = umap_tidy %>% sample_frac(0.01) %>% dplyr::select(-line, -drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point(data = df %>% unnest(data),
             aes(color = drug, v1, v2), alpha = 0.5, size = 0.35) +
  geom_path(data = df %>% unnest(plot_trace), 
            aes(v1, v2, group = paste0(drug, line), color = drug), size = 1.5, 
            arrow = arrow(angle = 10, ends = "last", type = "closed", length = unit(0.15, "inches")))+ 
  scale_color_brewer(type = "qual", palette = "Set2") +
  theme_cowplot() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
  theme(legend.position = "bottom") + 
  facet_wrap(~ line)
  
  return(gg)
}
```


```{r}
plot_df_raw <- create_plot_df_bulk(drug_order = c("Ulixertinib (BVD-523, VRT752271)", 
                                                  "TAK-733",
                                                  # "Trametinib (GSK1120212)",
                                                  # "Cobimetinib (GDC-0973, RG7420)",
                                                  # "MEK162 (ARRY-162, ARRY-438162)",
                                                  # "PD0325901",
                                                  # "Refametinib (RDEA119, Bay 86-9766)",
                                                  # "AZD8330",
                                                  # "Pimasertib (AS-703026)",
                                                  # "Selumetinib (AZD6244)",
                                                  # "PD318088",
                                                  # "VX-11e",
                                                  "Gefitinib (ZD1839)",
                                                  "Afatinib (BIBW2992)_uncertain",
                                                  "Erlotinib",
                                                  "Afatinib"),
                          loi = c("D018T01", "D027T01", "D046T01", "D027T01", "D019T01"), #18, 27, 46 (current figure) or lines 27, 19
                          n_sample = 400,
                          v1_cut = 5)

plot_df <- plot_df_raw %>% 
  mutate(plot_trace = purrr::map(plot, ~ .x$fit)) 


draw_trajectory_bulk(plot_df)

```


## MEK inhibitor Pseudotimes

Now I am building a function to create these tidy path objects.

```{r}
create_princurve_trace <- function(df){
  df_tmp_init <- df %>% 
  group_by(concentration) %>% 
  summarize(v1 = mean(v1),
            v2 = mean(v2)) %>% 
  ungroup()
  
  fit <- principal_curve(x = df %>% dplyr::select(v1, v2) %>% as.matrix(),
                       start = df_tmp_init %>% dplyr::select(v1, v2) %>% as.matrix(),
                       approx_points = FALSE, 
                       trace = TRUE, 
                       plot_iterations = FALSE, 
                       maxit = 30,
                       stretch =2)
  
  plot(fit)
  points(df_tmp_init%>% dplyr::select(v1, v2) %>% as.matrix())

  result <- list()
  
  result$fit <- fit$s[fit$ord,] %>% as_tibble()
  result$center <- df_tmp_init
  result$input <- df
  
  return(result)
}


create_plot_df <- function(loi, drug_order, n_sample){
  set.seed(123)
  
  center_df <- umap_tidy %>%
    filter(drug %in% drug_order) %>%
  filter(line %in% loi) %>% #, "D027T01", "D019T01"
  mutate(drug = factor(drug, levels =drug_order)) %>% 
  arrange((drug)) %>% 
  group_by(drug, line, concentration) %>%
  sample_n(n_sample, replace = TRUE) %>% ungroup() # adjust based on pan-line vs. 

  plot_df <- center_df %>%
  nest(-line, -drug) %>%
  mutate(plot = purrr::map(data, ~ create_princurve_trace(.x)))

  return(plot_df)
}

draw_trajectory <- function(df){
  gg <- ggplot() + 
  geom_point_rast(aes(v1, v2), data = umap_tidy %>% sample_frac(0.01) %>% dplyr::select(-drug), alpha = 1, size = 0.35, color = "#f1f1f1") + 
  geom_point(data = df %>% unnest(data),
             aes(color = line, v1, v2), alpha = 0.5, size = 0.35) +
  geom_path(data = df %>% unnest(plot_trace), 
            aes(v1, v2, group = paste0(drug, line), color = line), size = 1.5, 
            arrow = arrow(angle = 10, ends = "last", type = "closed", length = unit(0.15, "inches")))+ 
  scale_color_brewer(type = "qual", palette = "Set2") +
  theme_cowplot() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
  theme(legend.position = "bottom") + 
  facet_wrap(~ drug)
  
  return(gg)
}
```

I am generating a pseudotime plot

```{r}
set.seed(1234)

plot_df_raw <- create_plot_df(drug_order = c("Trametinib"),
                          loi = c("D055T01", "D054T01", "D046T01"),
                          n_sample = 200)

plot_df <- plot_df_raw %>% 
  mutate(plot_trace = purrr::map(plot, ~ .x$fit)) 


gg_tram_traj <- draw_trajectory(plot_df)

gg_tram_traj 
```

I am repeating the same figure for D027T01

```{r}
set.seed(123)

plot_df_raw <- create_plot_df(drug_order = c("Trametinib"),
                          loi = c("D027T01"),
                          n_sample = 800) 

plot_df <- plot_df_raw %>% 
  mutate(plot_trace = case_when(line == "D027T01" ~ purrr::map(plot, ~ .x$fit %>% .[1:3000,]))) 

gg_tram_traj_sel <- draw_trajectory(plot_df)

gg_tram_traj_sel
```
UMAP Embedding with trajektories: Trametinib and Binimetinib for PDO lines 19 and 27
UMAP highlighting all MEK inhibitors used (from L02/L03 - without concentrations) in all PDO lines (or if plot is too busy only from lines 18, 27, 46 (current figure) or lines 27, 19 (used in new figure for trajektories)
(3. Maybe: EGFR-Phenotype as Comparison - same plots as above, but with Gefitinib, Afatinib, Erlotinib:
UMAP Embedding with trajektories: Afatinib and Gefitinib for PDO lines 19 and 27
UMAP highlighting all EGFR targeted therapies in L02/L03 in same lines as above)

```{r}
set.seed(13)

plot_df_raw <- create_plot_df(drug_order = c("Binimetinib"),
                          loi = c("D027T01", "D019T01"),
                          n_sample = 800) 

plot_df <- plot_df_raw %>% 
  mutate(plot_trace = case_when(line == "D027T01" ~ purrr::map(plot, ~ .x$fit %>% .[1:2000,]),
                                line == "D019T01" ~ purrr::map(plot, ~ .x$fit %>% .[1:3500,]))) 

gg_bini_traj_sel <- draw_trajectory(plot_df)

gg_bini_traj_sel
```


# Plot Export

```{r}
#gg_size_dist + ggsave(paste0(PATH, "reports/figures/gg_size_dist.pdf"))
#gg_size_dist_log +  ggsave(paste0(PATH, "reports/figures/gg_size_dist_log.pdf"))
#ggdrug_size + ggsave(paste0(PATH, "reports/figures/gg_drug_dist_size_log.pdf"))
#gg_ecdf + ggsave(paste0(PATH, "reports/figures/gg_size_dist_ecdf.pdf"))
#gg_size_dist_morph_ridge +  ggsave(paste0(PATH, "reports/figures/gg_size_dist_morph_ridge.pdf"), width = 4, height = 4)

# ggdrug_count + ggsave(paste0(PATH, "reports/figures/gg_drug_dist_n_log.pdf"))
# gg_drug + ggsave(paste0(PATH, "reports/figures/gg_drug.pdf"), width = 8, height = 4)
# gg_size_drug + ggsave(paste0(PATH, "reports/figures/gg_trametinib_size_dose.pdf"), width = 3.65, height = 3.65)

gg_bortezomib + ggsave(paste0(PATH, "reports/figures/gg_bortezomib.pdf"), width = 2, height = 8)

gg_mtor_all + ggsave(paste0(PATH, "reports/figures/gg_mtor.pdf"), width = 6, height = 3)
gg_mtor_all + ggsave(paste0(PATH, "reports/figures/gg_gsk.pdf"), width = 6, height = 3)

gg_tram_traj_sel  + ggsave(paste0(PATH, "reports/figures/gg_tram_traj_sel.pdf"), width = 4, height = 4)
gg_bini_traj_sel + ggsave(paste0(PATH, "reports/figures/gg_bini_traj_sel.pdf"), width = 4, height = 4)
gg_tram_traj + ggsave(paste0(PATH, "reports/figures/gg_tram_traj.pdf"), width = 4, height = 4)

gg_cystic + ggsave(paste0(PATH, "reports/figures/gg_cystic.pdf"), width = 4, height = 4)
```

```{r}
plot_grid(gg_size_dist_morph_ridge, gg_size_replicate,
          labels = c('A', 'B'), label_size = 12, ncol = 2) +
  ggsave(paste0(PATH, "reports/panel_size_dist.pdf"), width = 8, height = 4)

plot_grid(plot_grid(ggdrug_size, ggdrug_count, labels = c('A', 'B'), label_size = 12),
          gg_drug,
          gg_size_drug,
          labels = c('', 'C', 'D'), label_size = 12, ncol = 1) +
  ggsave(paste0(PATH, "reports/panel_size_drug.pdf"), width = 8, height = 12)

gg_mtor_all

gg_size_supp

```


# Supplement

Here I collect pieces of code that did not make it into the final analysis but can be run in theory. In order to access these peaces of code, you have to open the *.RMD* file. 

```{r}
knitr::knit_exit()
```





















## Mixture modeling of Organoid size distribution


I estimate a log-gaussian mixture model of organoid size.

```{r, eval = TRUE}
library(mixtools)

df_mix <- normalmixEM(umap_tidy$size_log, k = 2)

df_fus <- df %>% dplyr::select(size_log) %>% mutate(posterior = "joined") %>% 
  rbind(data.frame(size_log = df_mix$x,
                   posterior = df_mix$posterior %>% as.data.frame() %>% round() %>% .$comp.1))

ggplot(df) + 
  geom_density(aes(x = size_log, y=..count..), alpha = 0.2) + 
  theme_cowplot()

ggplot(df_fus) + 
  geom_density(aes(x = size_log, y=..count.., fill = posterior), alpha = 0.2) + 
  theme_cowplot()
```

```{r, eval = TRUE}
df_mix
```



I scale this approach to all other organoid lines. 

```{r}
set.seed(0123349)

umap_mix <- umap_tidy %>% filter(drug == "DMSO")  %>% 
  nest(-line, -replicate) %>%
  mutate(mix_fit = map(data, ~ normalmixEM(.x$size_log, k = 2, maxit = 5000)),
         prob = map(mix_fit, ~ .x$posterior %>% as.data.frame() %>% .$comp.1)) 

umap_mix = umap_mix %>% unnest(prob, data) 
```

```{r}
set.seed(0.1)

df = umap_mix %>% mutate(posterior = round(prob)) %>% mutate(posterior = factor(posterior))

df %>% 
  sample_frac(0.01) %>%
  ggplot(aes(prob_dead, fill = posterior)) + 
  geom_histogram() + 
  scale_fill_brewer(type= "qual") + 
  scale_y_log10() + 
  theme_cowplot()

ggplot(df) + 
  geom_density(aes(x = size_log, fill = posterior, group = replicate), alpha = 0.5) + 
  facet_wrap(~ line) + 
  labs(caption = "Plotted are Replicates, DMSO treated organoids",
       x = "ln(size)") + 
  theme(legend.position = "bottom") +
  theme_cowplot()


ggplot(df) + 
  geom_density(aes(x = size_log, y=..count.., fill = posterior), alpha = 0.2) + 
  theme_cowplot() + 
  facet_wrap(~ line)
```

## Count distribution of Paclitaxel treated organoids 



```{r}
drug_count_param <- drug_count %>% 
  nest(-concentration, -line) %>% 
  mutate(fit = map(data, ~ fitdistrplus::fitdist(.x$n, "lnorm")))
        # param = map(fit, ~ .x$estimate %>% broom::tidy()))

drug_count_param %>% unnest(param) %>% 
  filter(names == "meanlog") %>%
  ggplot(aes(concentration, exp(x))) + 
    geom_point() + 
  theme_cowplot()
```


```{r}
drug_count_param %>% unnest(param) %>% 
  filter(names == "meanlog") %>% 
  dplyr::select(concentration, line, count = x) %>% 
  left_join(tram_size_param %>% unnest(param) %>% 
  filter(names == "meanlog") %>% dplyr::select(concentration, line, size = x)) %>% 
  
  ggplot(aes(count, size, color = concentration)) + 
  geom_point() + 
 # geom_path(aes(group = line)) +
  scico::scale_color_scico_d() + 
  theme_cowplot()


```




## Main Panel

```{r}
plot_grid(gg_size, 
          plot_grid(gg_line,
                    gg_bortezomib,
                    ncol = 2),
          labels = "AUTO",
          ncol = 2) + 
  gggsave(here::here("results/figures/panel_6.pdf"))
```

## Supplemental Panel

```{r}
plot_grid(gg_harmony_effect, 
          gg_size_supp,
          gg_size_dist
          )
```

## Running harmony on batches

We abandoned this approach as we did not screen organoid lines across batches. Reducing batch differences automatically led to a reduction of line specific effects. 


First, I illustrate batch effects and how they are reduced

```{r, eval = FALSE}
set.seed(123)

pca_metadata_frac <- pca_metadata %>% sample_frac(0.001)

pca_harmony <- pca_metadata_frac %>% dplyr::select(contains("V"))
metadata_harmony <- pca_metadata_frac %>% dplyr::select(screen_ID, Line)

my_harmony_embeddings <- HarmonyMatrix(
  pca_harmony, metadata_harmony, c("screen_ID"),
  do_pca = FALSE,
  verbose = TRUE, 
  return_object = TRUE
)

umap_pre <- uwot::umap(my_harmony_embeddings$Z_orig %>% t() %>% as_tibble(), verbose = TRUE)
umap_post <- uwot::umap(my_harmony_embeddings$Z_corr %>% t() %>% as_tibble(), verbose = TRUE)

harmony_effect <- umap_pre %>% as_tibble() %>% mutate(status = "pre") %>% cbind(id = 1:nrow(.)) %>% cbind(metadata_harmony)%>%
  rbind(.,umap_post %>% as_tibble() %>% mutate(status = "post")%>% cbind(id = 1:nrow(.)) %>% cbind(metadata_harmony)) 

harmony_effect %>% saveRDS(here::here("vignettes/07_organoid_unsupervised_exploration/harmony/harmony_effect.Rds"))
```

Here we can see the effect of Harmony on batch discordances.

```{r, eval = TRUE}
harmony_effect <- readRDS(here::here("vignettes/07_organoid_unsupervised_exploration/harmony/harmony_effect.Rds"))

gg_harmony_effect <- harmony_effect %>% 
  mutate(status = factor(status, levels = c("pre", "post"))) %>%
  ggplot(aes(V1, V2, color = screen_ID)) + 
  geom_point_rast(alpha = 0.5, size = 0.35) + 
  theme_cowplot() + 
  scale_color_brewer(type = "qual", palette = 2) +
  facet_wrap( ~ status) + 
  #ggsave(here::here("vignettes/07_organoid_unsupervised_exploration/pre_post_harmony.png")) + 
  NULL
```

Still, line-specific differences are conserved in this method.

```{r}
harmony_effect %>% 
 filter(status == "post") %>%
  ggplot(aes(V1, V2, color = screen_ID)) + 
  geom_point_rast(alpha = 0.5, size = 0.35) + 
  theme_cowplot() + 
  facet_wrap( ~ Line) + 
  scale_color_brewer(type = "qual") +
  #ggsave(here::here("vignettes/07_organoid_unsupervised_exploration/post_harmony_line.png")) + 
  NULL
```


## Running Harmony on replicate and on replicate+line level

```{r, eval = FALSE}
set.seed(123)

pca_harmony <- pca_metadata %>% dplyr::select(contains("V"))
metadata_harmony <- pca_metadata %>% dplyr::select(screen_ID, Line)

harmony_id <- HarmonyMatrix(
  pca_harmony, metadata_harmony, c("screen_ID"),
  do_pca = FALSE,
  verbose = TRUE, 
  return_object = TRUE
)

harmony_id$Z_corr %>% saveRDS(here::here("vignettes/07_organoid_unsupervised_exploration/harmony/harmony_pca_id.Rds"))
```

I load the Harmony results.

```{r, eval = FALSE}
harmony_id_corr <- readRDS(here::here("vignettes/07_organoid_unsupervised_exploration/harmony/harmony_pca_id.Rds"))
```



## Running harmony on both line and batch 

```{r, eval = FALSE}
harmony_id_line <- HarmonyMatrix(
  pca_harmony, metadata_harmony, c("screen_ID", "Line"),
  do_pca = FALSE,
  verbose = TRUE, 
  return_object = TRUE
)


harmony_id_line %>% saveRDS(here::here("vignettes/07_organoid_unsupervised_exploration/harmony/harmony_pca_id_line.Rds"))

```

## Individual Pseudotimes


```{r, eval = FALSE}
set.seed(123)

drug_order <- umap_tidy %>%
  dplyr::select(drug) %>%
    filter(grepl(drug, pattern = "__")) %>% .$drug %>% unique() %>% sort()

curve_df <- umap_tidy %>%
    filter(drug %in% drug_order) %>%
  mutate(drug = factor(drug, levels =drug_order)) %>% 
  arrange((drug)) %>% 
  group_by(drug, line) %>%
  sample_n(1000, replace = TRUE) %>% # adjust based on pan-line vs. line level 
  ungroup() %>% 
  # running the princcurve algorithm
  separate(drug, c("compound", "concentration"), sep = "__", remove = FALSE) %>%
  nest(-compound, -line) %>%
  #head(10) %>% 
  mutate(plot = purrr::map(data, ~ create_princurve_trace(.x)))

curve_df %>% saveRDS(here::here("vignettes/07_organoid_unsupervised_exploration/curve_df.Rds"))
```


```{r, eval = FALSE}
set.seed(123)

drug_order <- umap_tidy %>%
  dplyr::select(drug) %>%
    filter(grepl(drug, pattern = "__")) %>% .$drug %>% unique() %>% sort()

curve_df <- umap_tidy %>%
    filter(drug %in% drug_order) %>%
  mutate(drug = factor(drug, levels =drug_order)) %>% 
  arrange((drug)) %>% 
  group_by(drug, line) %>%
  sample_n(1000, replace = TRUE) %>% # adjust based on pan-line vs. line level 
  ungroup() %>% 
  # running the princcurve algorithm
  separate(drug, c("compound", "concentration"), sep = "__", remove = FALSE) %>%
  nest(-compound, -line) %>%
  #head(10) %>% 
  mutate(plot = purrr::map(data, ~ create_princurve_trace(.x)))

curve_df %>% saveRDS(here::here("vignettes/07_organoid_unsupervised_exploration/curve_df.Rds"))
```


```{r, eval = FALSE}
gg_curve_df <- curve_df %>% 
  mutate(compound_line = paste0(compound, "__", line)) %>%
  mutate(plot_trace = purrr::map(plot, ~ .x$fit %>% .[, ])) %>% 
  unnest(plot_trace) 

gg_curve_df %>%
  ggplot(aes(v1, v2)) + 
  geom_point(data = curve_df %>% unnest(data),
             aes(color = compound), alpha = 0.05, size = 0.35) +
  
  geom_path(aes(group = compound_line, color = compound_line), size = 1.5, arrow = arrow(angle = 10, ends = "last", type = "closed", length = unit(0.15, "inches"))) + 
  
  facet_grid(~ compound) +
   
  theme_cowplot() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
  theme(legend.position = "bottom")
```

## Plotting gigantic overview figure


```{r, eval = FALSE}
gg_line_overview <- plot_cells(ods, color_cells_by="morphological_class",
           alpha = 0.05, label_cell_groups = FALSE,
           rasterize = TRUE) + 
  facet_wrap(~ Line) + 
  theme(legend.position = "nothing") + 
  scale_color_manual(values = rev(c("#A2549B", "#D80D12", "#6E1614","#69B563", "#A9CEE1","#1E3B87"))) 
```

I directly write this supplementary plot to disc. 

```{r, eval = FALSE}
gg_line_overview + 
  ggsave(here("results/figures/single_organoid/line.pdf"), height = 8, width = 8)
```

## Running princurve on principle component data 


 In the past, I estimated a princurve directly from the UMAP embedding. I am not sure this is a good idea. Rather, I am going to test running the princurve algorithm on a set of PCs. 
What is the %-Variance explained per PC?

```{r}
vars_pca <- apply(reducedDims(ods)$PCA, 2, var)
vars_pca_scaled = vars_pca/sum(vars_pca)

vars_pca_scaled %>% plot()
```

This is the PCA-based princurve function. 

```{r}
create_princurve_trace_pca <- function(df){
  df_tmp_init <- df %>% 
  group_by(drug) %>% 
  summarise_at(vars(contains("pc")), funs(mean)) %>%
  ungroup()
  
  fit <- principal_curve(x = df %>% dplyr::select(pc1:pc5) %>% as.matrix(),
                       start = df_tmp_init %>% dplyr::select(pc1:pc5) %>% as.matrix(),
                       approx_points = FALSE, 
                       trace = TRUE, 
                       plot_iterations = FALSE, 
                       maxit = 30,
                       stretch =2)
  
  plot(fit)
  points(df_tmp_init%>% dplyr::select(pc1:pc5) %>% as.matrix())

  result <- list()
  
  result$fit <- fit$s[fit$ord,] %>% as_tibble()
  result$center <- df_tmp_init
  result$input <- df
  
  return(result)
}
```


```{r}
set.seed(123)

drug_order <- pca_tidy %>%
  dplyr::select(drug) %>%
    filter(grepl(drug, pattern = "__")) %>% .$drug %>% unique() %>% sort()

curve_df <- pca_tidy %>%
    filter(drug %in% drug_order) %>%
  mutate(drug = factor(drug, levels =drug_order)) %>% 
  arrange((drug)) %>% 
  group_by(drug, line) %>%
  sample_n(1000, replace = TRUE) %>% # adjust based on pan-line vs. line level 
  ungroup() %>% 
  # running the princcurve algorithm
  separate(drug, c("compound", "concentration"), sep = "__", remove = FALSE) %>%
  nest(-compound, -line) %>%
  head(1) %>% 
  mutate(plot = purrr::map(data, ~ create_princurve_trace_pca(.x)))
```


```{r}
set.seed(123)

gg_curve_df <- curve_df %>% 
  mutate(compound_line = paste0(compound, "__", line)) %>%
  mutate(plot_trace = purrr::map(plot, ~ .x$fit %>% .[, ])) %>% 
  unnest(plot_trace) 

pca_subset = pca_tidy %>% sample_frac(.01)
sample_index = pca_subset$uuid


curve_index = FNN::get.knnx(data = pca_subset %>% dplyr::select(pc1:pc5) %>% as.matrix(), gg_curve_df %>% dplyr::select(pc1:pc5) %>% as.matrix(), k=1, algorithm=c("kd_tree"))

gg_curve_df = cbind(gg_curve_df, umap_tidy %>% filter(uuid %in% sample_index) %>% .[curve_index$nn.index,] %>% dplyr::select(v1, v2))

gg_curve_df %>%
  ggplot(aes(v1, v2)) + 
  geom_point(data = curve_df %>% unnest(data) %>% dplyr::select(compound, line, uuid) %>% left_join(umap_tidy %>% dplyr::select(uuid, v1, v2)),
             aes(color = compound), alpha = 0.05, size = 0.35) +
  
  #geom_path(aes(group = compound_line, color = compound_line), size = 1.5, arrow = arrow(angle = 10, ends = "last", type = "closed", length = unit(0.15, "inches"))) + 
  geom_point(aes(group = compound_line, color = compound_line), size = 1.5) +
  #facet_grid(~ compound) +
  facet_wrap(~ compound_line) +
  theme_cowplot() +
  labs(x = "UMAP 1",
       y = "UMAP 2")+
  theme(legend.position = "bottom")
```


