---
title: "incell6000_setup_final"
author: "Niklas R"
date: "22 2 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Check profiles of imaged plates
```{r}
library(readr)
library(StatMatch)
library(tibble)
library(dplyr)
library(magrittr)
library(ggplot2)
library(lazyeval)
library(tidyr)

ID = "D0"

setwd("/Volumes/collab-bernd-fischer/PROMISE/incoming/PROMISE-10x-4t-c-16z/")
temp <- list.files(pattern = ID) %>%
  as_tibble() %>%
  mutate(time = file.info(.$value)$atime < Sys.time()-180,
         csv = paste0(.$value, "/", .$value, "_optimize_intensity.csv")) %>%
  filter(time == TRUE) 



plates.list <- lapply(temp$csv, read_csv)
temp.df <- do.call("rbind",plates.list)

plates <- temp.df %>% 
  mutate(plate = rep(temp$value, each = nrow(plates.list[[1]])),
         line = substr(plate, 22,28),
         library = substr(plate, 33,35),
         plate_number = substr(plate, 29,32),
         barcode = substr(plate, 22,35)) 


#rm(temp.df)
```

```{r}
plot.profile.max <- function(df){
  require(tidyverse)
  
  df %>% 
    as_tibble() %>%
    mutate(Identifier.z_index = as.factor(Identifier.z_index),
         EmissionFilter.name = as.factor(EmissionFilter.name)) %>%
  ggplot(., aes(x = Identifier.z_index, 
                y = MinMaxMean.max, 
                color = EmissionFilter.name)) +
      geom_boxplot() +
      scale_y_continuous(limits = c(0,65000)) + 
      geom_hline(yintercept =1000) + 
      ggtitle(df$plate)
    
  
}

```

```{r}
plot.profile.mean <- function(df, level_vector){
  require(tidyverse)
  
  df %>% 
    as_tibble() %>%
    mutate(Identifier.z_index = as.numeric(Identifier.z_index),
         EmissionFilter.name = factor(EmissionFilter.name, levels = c("dsRed", "FITC", "DAPI")),
         plate = factor(plate, levels = c(unique(plates$plate)))) %>%
  ggplot(aes(x = Identifier.z_index, 
                       y = MinMaxMean.mean, 
                       color = EmissionFilter.name)) +
      geom_point(alpha = 0) +
      geom_smooth(se = TRUE, size = 1, alpha = 0.5) + 
      scale_y_log10(limits = c(100,300), breaks = c(120,150,200,250)) + 
      geom_hline(yintercept =120) + 
      #geom_hline(yintercept =1000) + 
      ggtitle("z-level vs. channel intensity across plates") + 
      facet_grid(plate_number ~ line)
    
  
}
```

```{r}
library(platetools)
library(viridis)
plot.plate <- function(df, channel){
  require(platetools)
  temp <- df %>% 
    filter(EmissionFilter.name == channel) %>%
    group_by(Well.label, plate_number, line) %>%
    summarise(mean_score = mean(MinMaxMean.mean-100)) %>%
    mutate(Well.label.short = paste0(substr(Well.label, 1, 1), substr(Well.label, 5, 6)))
  
  set <- plate_map_grid(temp$mean_score, temp$Well.label.short, paste0(temp$line, temp$plate_number))
  
  raw_grid(data = set$values,
           well = set$well,
           plate_id = set$plate_label, 
           ncols = 3, 
           plate = 384) + 
    scale_fill_viridis()
 
  
  
}
```

```{r}
#temp <- temp[c(1:64512),]
plates %>%
  #filter(grepl("W01P006", plate)) %>%
  plot.profile.mean() #6 3
```

```{r}
plates %>%
  plot.plate(., "FITC") #
```








```{r}
plot.summary <- function(df){
  require(tidyverse)
  
  df %>% 
    as_tibble() %>%
    mutate(Identifier.z_index = as.numeric(Identifier.z_index),
         EmissionFilter.name = factor(EmissionFilter.name, levels = c("dsRed", "FITC", "DAPI")),
         line = factor(line, levels = c("W01", "K02", "A03", "B04"))) %>%
  ggplot(aes(x = Identifier.z_index, 
                       y = sd.mean, 
                       color = EmissionFilter.name)) +
      geom_point(alpha = 0) +
      geom_smooth(se = TRUE, size = 1, alpha = 0.5) + 
      #scale_y_log10(limits = c(100,300), breaks = waiver()) + 
      #geom_hline(yintercept =120) + 
      #geom_hline(yintercept =1000) + 
      ggtitle("z-level vs. channel standard deviation across plates") + 
      facet_grid(plate_number ~ line)
    
}
```


```{r}
plates %>%
  group_by(line, library, EmissionFilter.name, Identifier.z_index) %>%
  summarise(mean.mean = mean(MinMaxMean.mean),
            var.mean = var(MinMaxMean.mean),
            sd.mean = sd(MinMaxMean.mean),
            mean.max = mean(MinMaxMean.max),
            var.max = var(MinMaxMean.max),
            sd.max = sd(MinMaxMean.max)) %>%
  as.data.frame() -> test

test %>%
  plot.summary()
```


```{r}
library(tidyverse)


plates %>%
  group_by(plate) %>%
    do(mean.plot = plot.profile.mean) -> temp
 # plot.profile.mean() #6 3

temp$mean.plot


```
The plate profiles for B and A plates are optimal; for W and K, however, the histogram is shifted to the left -suggesting a lower offset for imaging would be more appropiate. 
This is ,however, contraintuitive to the qualitative information gained yesterday and today. All organoids that could not be captured on the max projection are in higher z-levels, which could not be reached with the available settings.

```{r}
plates %>%
  filter(grepl("W01", plate) & grepl("L07", plate) & Well.label == "C - 5") %>%
  plot.profile.mean
```

```{r}
plates %>%
  filter()
```

