---
title: "Clustering of Organoid Features"
author: "Jan Sauer"
date: "10.11.2017"
output: 
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r, message=FALSE, warning=FALSE, include=FALSE}
library(PROMISE)
library(ggplot2)
library(rhdf5)
library(Rtsne)
# library(sparcl)
# library(fpc)
library(cluster)
library(ape)
library(abind)
library(EBImage)

# Directories and files
config_dir = "/collab-ag-fischer/PROMISE/data-10x-4t-c-16z/configdir"
feature_dir = "/collab-ag-fischer/PROMISE/data-10x-4t-c-16z/features"
source(file.path(config_dir, "watchdogConfig.R"))

# Parameters
feature_type = "organoids"
```

# Clustering of Organoids in well D004T01P005L08_A_09
The well in question shows a clear split in organoid types.

```{r}
plate = "D004T01P005L08"
row = "A"
col = "09"
preview_fn = thumbnailFilename(
  filedir = file.path(htmldir, plate), platename = plate, 
  row = row, col = col, configdir = config_dir)
```

!["D004T01P005L08_A_09"](`r preview_fn`)

## Load and Preprocess data

I use a set of pseudo-labels to get a rough idea of how well the clustering methods work. In particular, I've annotated blurry objects ("BLU"), shrapnel ("SHR"), live organoids ("NEG"), and dead organoids ("POS").
```{r}
# features_fn = file.path(feature_dir, plate, "wells_normalized", sprintf("%s_%s_%s_features_normalized.h5", plate, row, col))
# features = h5read(features_fn, sprintf("features_%s", feature_type))
# colnames(features) = h5read(features_fn, sprintf("feature_names_%s", feature_type))
# object_type = h5read(features_fn, sprintf("object_type_%s", feature_type))
features_fn = file.path(feature_dir, plate, "wells_normalized", sprintf("%s_%s_%s_features_normalized.h5", plate, row, col))
features = h5read(features_fn, sprintf("features_%s", feature_type))
colnames(features) = h5read(features_fn, sprintf("feature_names_%s", feature_type))
object_type = h5read(features_fn, sprintf("object_type_%s", feature_type))

# This is only for numeric stability
features = round(features, 8)
```

Remove features with any NA's and with constant values
```{r}
# Remove NA features
na_feature_counts = colSums(!is.finite(features))
features = features[, na_feature_counts == 0]
```

Discretize the features by rounding to the nearest 0.5
```{r}
features = round(features*2) / 2
```

```{r}
# Remove constant features
feature_range = apply(features, 2, function(x) mad(x, na.rm = TRUE))
features = features[, feature_range > 0]
```

<!-- Correlating such a high-dimensional space can be difficult, so the features are discretized to remove some noise. To make features comparable, I discretize each feature individually: -->
<!-- * if x <= 1st Quartile then set x to -1 -->
<!-- * if 1st Quartile <= x <= 3rd Quartile then set x to 0 -->
<!-- * if x >= 3rd Quartile then set x to 1 -->
<!-- ```{r} -->
<!-- features = apply(features, 2, function(x) { -->
<!--   q1 = quantile(x, 0.25, na.rm = TRUE) -->
<!--   q3 = quantile(x, 0.75, na.rm = TRUE) -->
<!--   out = rep(0, length(x)) -->
<!--   out[x <= q1] = -1 -->
<!--   out[x >= q3] = 1 -->
<!--   return(out) -->
<!-- }) -->
<!-- ``` -->

<!-- ## PCA -->
<!-- Remove useless features/constant features and perform a PCA. A look at the cumulative explained variance of the PCA shows that there are no "distinct" components; nearly 150 features are required to explain 90% of the variance in the dataset. -->
<!-- ```{r} -->
<!-- # Remove highly correlated features (rho >= 0.99) -->
<!-- features_cor = cor(features, use = "pairwise.complete.obs") -->
<!-- feature_cor_vector = features_cor[upper.tri(features_cor, diag = FALSE)] -->
<!-- features_cor[upper.tri(features_cor, diag = TRUE)] = 0 -->
<!-- features_to_keep = names(which(!apply(features_cor,2,function(x) any(abs(x) > 0.99, na.rm = TRUE)))) -->
<!-- features = features[, features_to_keep] -->

<!-- # PCA -->
<!-- pca = prcomp(features, center = TRUE, scale. = TRUE) -->
<!-- features_pca = pca$x[, cumsum(pca$sdev) / sum(pca$sdev) <= 0.9] -->
<!-- pca_df = data.frame( -->
<!--   "Cumulative.Variance" = cumsum(pca$sdev) / sum(pca$sdev),  -->
<!--   "Principal.Component" = seq_along(pca$sdev)) -->
<!-- ggplot(data = pca_df) + geom_point(aes(x = Principal.Component, y = Cumulative.Variance)) +  -->
<!--   ggtitle("Cumulative Explained Variance of PCA Components") -->
<!-- ``` -->

## t-SNE
I look at the t-SNE for various perplexities. The coloring indicates the type of object; I differentiate between blurry objects, "organoids" with an area >= 1500 pixels, and "shrapnel" with an area of < 1500 pixels. The clustering shows a separation between blurry objects and organoids.
```{r}
perplexities = c(30)
tsne_data = lapply(perplexities, function(x) Rtsne(
  features, perplexity = x, initial_dims = 250, 
  theta = 0, pca_scale = TRUE, pca_center = TRUE))
names(tsne_data) = perplexities
for(x in perplexities) {
  tsne_df = data.frame(tsne_data[[as.character(x)]]$Y)
  tsne_df$Object.Type = object_type
  plot(
    ggplot(data = tsne_df) + geom_point(aes(x = X1, y = X2, col = Object.Type)) + 
      ggtitle(sprintf("T-SNE Plot with perplexity %s", x)))
}
```

All further analyses will be performed without the blurry organoids.
```{r}
features_focus = features[object_type != "BLURRY", ]
object_type_focus = object_type[object_type != "BLURRY"]
```

```{r}
perplexities = c(20)
tsne_data = lapply(perplexities, function(x) Rtsne(
  features_focus, perplexity = x, initial_dims = 250, 
  theta = 0, pca_scale = TRUE, pca_center = TRUE))
names(tsne_data) = perplexities
for(x in perplexities) {
  tsne_df = data.frame(tsne_data[[as.character(x)]]$Y)
  tsne_df$Object.Type = object_type_focus
  plot(
    ggplot(data = tsne_df) + geom_point(aes(x = X1, y = X2, col = Object.Type)) + 
      ggtitle(sprintf("T-SNE Plot with perplexity %s", x), subtitle = "Only focused objects"))
}
```

## Hierarchical Clustering
An unsupervised hierarchical clustering was performed. Coloring is as with the t-SNE. If I ignore the shrapnel and blurry objects then the organoids themselves seem to cluster into 8 categories.
```{r, width=12, height=12}
# The normalization doesn't change the clustering, only the distances, making the 
# dendrogram easier to read
# features_norm = apply(features_focus, 2, function(x) {
#   (x - mean(x)) / sd(x)})
d = dist(features_focus)
h = hclust(d)
colors = c("red", "blue", "green", "black")
plot.phylo(
  as.phylo(h), cex = 0.6, label.offset = 0.5, type = "phylogram", 
  show.tip.label = FALSE, edge.color = colors[as.factor(object_type_focus)], 
  tip.color = colors[as.factor(object_type_focus)])
legend(x="topleft",legend = levels(as.factor(tolower(object_type_focus))), fill = colors,xpd = T, cex = 1)
```

## K-Means Clustering 
I take a look at the gap statistic (Tibshirani et al, 2000; Grün et al, 2015) to determine the optimal number of clusters.
```{r}
pam1 = function(x, k) list(cluster = pam(x, k, cluster.only = TRUE))
gapstat = clusGap(x = features_focus, FUNcluster = pam1, K.max = 20, B = 50, verbose = FALSE)
plot(gapstat, main = "Gap Statistic for Full Feature Set")
```

As seen above, organoids separate very nicely from the shrapnel. I attempt to recreate this split using only those objects. While the optimal number of clusters is clearly k = 1, a significant local maximum can also be seen at k = 7
```{r}
gapstat = clusGap(
  x = features_focus[object_type_focus == "Organoid", ], FUNcluster = pam1, 
  K.max = 20, B = 50, verbose = FALSE)
plot(gapstat, main = "Gap Statistic for Organoid Features")
```

## Examples of clustered organoids
Next, I show some examples of the individual clusters to visually confirm the clustering. The clustering isn't ideal, as three of the clusters have only one entry.
```{r}
object_id = seq_len(nrow(features))
object_id_organoids = object_id[object_type == "Organoid"]

# Perform clustering
clusters = pam1(x = features[object_type == "Organoid", ], k = 7)$cluster
print(table(clusters))
```

```{r}
# Load and normalize projection
data_fn = projectionHdf5Filename(
  filedir = file.path(hdf5projection, plate), platename = plate, 
  row = row, col = col, configdir = config_dir)
proj = h5read(data_fn, "images")
for(ch_i in 1:3) {
  for(field_i in 1:4) {
    q = quantile(proj[,,ch_i,field_i], 0.999)
    proj[,,ch_i,field_i][proj[,,ch_i,field_i] > q] = q
  }
  proj[,,ch_i,] = proj[,,ch_i,] - min(proj[,,ch_i,])
  proj[,,ch_i,] = proj[,,ch_i,] / max(proj[,,ch_i,])
}

# Load unaltered features for coordinates
orig_features_fn = file.path(
  feature_dir, plate, "wells", 
  sprintf("%s_%s_%s_features.h5", plate, row, col))
features_original = h5read(orig_features_fn, sprintf("features_%s", feature_type))
colnames(features_original) = h5read(orig_features_fn, sprintf("feature_names_%s", feature_type))

# Extract sample organoids
num_samples = 5
valid_clusters = as.numeric(names(table(clusters)[table(clusters) >= num_samples]))
set.seed(20171110)
all_segments = list()
for(cl in valid_clusters) {
  cl_object_ids = sample(object_id_organoids[clusters == cl], num_samples, replace = FALSE)
  cl_coords = features_original[cl_object_ids, c("x.0.m.cx", "x.0.m.cy", "x.0.s.radius.max", "FIELD")]
  colnames(cl_coords) = c("x", "y", "radius", "field")
  segments = list()
  for(oid in seq_len(nrow(cl_coords))) {
    xx = round(cl_coords[oid, "x"])
    yy = round(cl_coords[oid, "y"])
    rad = ceiling(cl_coords[oid, "radius"]) + 10
    fld = cl_coords[oid, "field"]
    minx = max(xx - rad, 0)
    maxx = min(xx + rad, dim(proj)[1])
    miny = max(yy - rad, 0)
    maxy = min(yy + rad, dim(proj)[2])
    segments[[oid]] = proj[minx:maxx, miny:maxy, 1:3, fld]
  }
  all_segments[[as.character(cl)]] = segments
}

# Embed all segments into a single image
max_x = 0
max_y = 0
for(cl in all_segments) {
  for(seg in cl) {
    max_x = max(max_x, dim(seg)[1] + 5)
    max_y = max(max_y, dim(seg)[2] + 5)
  }
}

all_embed_segments = list()
ii = 1
for(cl in all_segments) {
  embed_segments = list()
  jj = 1
  for(seg in cl) {
    new_embed = array(0, dim = c(max_x, max_y, 3))
    ddim = round((dim(new_embed) - dim(seg))[1:2] / 2)
    new_embed[ddim[1]:(ddim[1] + dim(seg)[1] - 1), ddim[2]:(ddim[2] + dim(seg)[2] - 1), ] = seg
    new_embed[1:dim(new_embed)[1], 1, 1:3] = 1
    new_embed[1, 1:dim(new_embed)[2], 1:3] = 1
    embed_segments[[jj]] = new_embed
    jj = jj + 1
  }
  all_embed_segments[[ii]] = abind(embed_segments, along = 2)
  ii = ii + 1
}
all_embed_segments = Image(abind(all_embed_segments, along = 1), colormode = "color")
writeImage(all_embed_segments, "OrganoidClusters.png")
```

[Organoid Clusters. Each column shows 5 randomly chosen members of one of the clusters with more than one member](OrganoidClusters.png)